### 2015-07-03 V4046Sgr2013/ms_v4046sgr2013_chapter.tex
initial commit

\chapter{The Structure of the Evolved Circumbinary Disk around V4046 Sgr}{\label{c:v4046_structure}

\section*{abstract}
We present sensitive, sub-arcsecond resolution Submillimeter Array observations 
of the protoplanetary disk around the nearby, pre-main sequence spectroscopic 
binary V4046 Sgr.  We report for the first time a large inner hole ($r=29$\,AU) 
spatially resolved in the 1.3\,mm continuum emission and study the structure of 
this disk using radiative transfer calculations to model the spectral energy
distribution (SED), continuum visibilities, and spectral line emission of CO and
its main isotopologues.  Our modeling scheme demonstrates that the majority of 
the dust mass is distributed in a narrow ring (centered at 37\,AU with a FWHM of
16\,AU) that is $\sim$5$\times$ more compact than the gas disk.  This structure 
implies that the dust-to-gas mass ratio has a strong spatial variation, ranging 
from a value much larger than typical of the interstellar medium (ISM) at the 
ring to much smaller than that of the ISM at larger disk radii.  We suggest that
these basic structural features are potentially observational signatures of the 
accumulation of solids at a local gas pressure maximum.  These models also 
require a substantial population of $\sim$$\mu$m-sized grains inside the central
disk cavity.  We suggest that this structure is likely the result of dynamical 
interactions with a low-mass companion, although photoevaporation may also play 
a secondary role.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the past few years, high angular resolution millimeter/radio-wave
observations have facilitated rapid development in our understanding of some
fundamental aspects of protoplanetary disk evolution.  For example, there is
mounting evidence for substantial discrepancies between the spatial
distributions of mm-sized dust particles and molecular gas in protoplanetary
disks, possibly caused by a dramatic decrease in the dust-to-gas ratio at large 
radii \citep[e.g.,][]{panic09,andrews12}.  Coupling those results with new 
analyses that identify a systematic decrease in dust particle sizes in the 
outer disk \citep{guilloteau11,perez12}, there seems to be good progress toward 
the promise of empirical constraints on the key processes tied to the growth 
and migration of disk solids.  Meanwhile, there has been marked improvement in 
the characterization of another key factor in disk evolution: dynamical 
interactions with companions.  \citet{harris12} provided a sweeping 
quantitative confirmation of the seminal work by \citet{jensen96}, 
demonstrating that tidal interactions in multiple star systems are a prominent 
issue for disk survival, especially for separations in the $\sim$3-30\,AU 
range.  And \citet{andrews11} (among others) have speculated that the ring-like 
millimeter-wave dust continuum emission morphologies noted for the so-called 
``transition" disks are most likely created by an analogous dynamical 
interaction process, though in this case the perturbers may be embedded {\it 
planetary} companions.

For the time being, most of these data-driven studies of disk evolution are 
somewhat piecemeal in terms of both diagnostics and target samples, due to 
practical observational limitations.  While that should soon change 
dramatically with the completion of the Atacama Large Millimeter Array (ALMA)
facility, it is important to recognize that some individual targets will always
serve as particularly illuminating case studies for different aspects of disk
evolution.  Here, we focus on the disk around V4046 Sgr, a remarkable system
with observable characteristics that are likely simultaneously shaped by {\it 
all} of the fundamental processes involved in disk evolution.

V4046 Sgr is a double-lined spectroscopic binary with a 2.4-day orbital period 
\citep{byrne86,delareza86,quast00,stempels04}.  Radial velocity monitoring
indicates a nearly equal-mass pair of K-type stars on a close ($a \approx 
0.045$\,AU), circular ($e \le 0.001$) orbit \citep{stempels13}.
\citet{kastner11} suggested that GSC 07396$-$00759 could be an additional
distant ($\sim$2\farcm8) companion, which itself might be an unresolved close
binary \citep[see also][]{nataf10}.  Two independent, dynamical methods to
estimate the V4046 Sgr stellar masses are in agreement, with $M_{\ast,1} = 
0.90\pm0.05$ and $M_{\ast,2} = 0.85\pm0.04$\,M$_{\odot}$ for the primary and
secondary, respectively \citep{rosenfeld12,stempels13}.  Coupling these
dynamical mass measurements to the inferred temperatures and luminosities,
pre-main sequence evolution models suggest that V4046 Sgr has an age of
$\sim$10-20\,Myr \citep{rodriguez10,donati11,rosenfeld12}.  Those ages are 
consistent with the conjecture of \citet{torres06,torres08} that V4046 Sgr is a 
member of the $\beta$ Pic moving group, with a kinematic parallax distance of 
only 73\,pc from the Sun.  Evidence for a disk around the V4046 Sgr binary first
came from emission line accretion signatures \citep[H$\alpha$ equivalent widths 
of $\sim$30-120\AA;][]{merrill50,henize76,hbc88} and an infrared continuum 
excess in the {\it IRAS} bands \citep[e.g.,][]{johnson86,weintraub90,weaver92}. 
Strong (sub)millimeter emission was detected by \citet{jensen96}, suggesting a 
relatively large disk mass.  \citet{jensen97} modeled the infrared SED and 
concluded that it was consistent with an extended circumbinary disk truncated at
an inner edge radius of $\sim$0.2\,AU, as would be expected from dynamical 
interactions with the central binary \citep[e.g.,][]{artymowicz94}.  Later, 
\citet{kastner08} discovered a substantial reservoir of molecular gas in the 
disk orbiting V4046 Sgr, which was subsequently imaged and found to span 
$\sim$800\,AU in diameter 
\citep[roughly 10\arcsec\ on the sky;][]{rodriguez10,oberg11,rosenfeld12}.  
Fitting an elliptical Gaussian to their 230\,GHz continuum visibilities, 
\citet{rodriguez10} suggested that the mm-wave dust emission 
is concentrated inside a $\sim$40\,AU radius.

Taken together, these properties make the V4046 Sgr system a significant 
benchmark for studies of protoplanetary disk evolution for three
key reasons.  First, the advanced age of the disk suggests that evolutionary
effects have had more time to progress, and therefore should exhibit more
obvious observational signatures than for a typical $\sim$1\,Myr-old disk.
Evolutionary mechanisms like particle growth and migration
\citep[e.g.,][]{birnstiel12a}, giant planet formation
\citep[e.g.,][]{pollack96,hubickyj05}, and gas dissipation via photoevaporative
winds \citep[e.g.,][]{clarke01,alexander09} should all be relevant in shaping
disk properties at the V4046 Sgr age.  For reference, such disks are rare: V4046
Sgr harbors the {\it only} gas-rich disk in the $\beta$ Pic moving group, and 
is one of a handful of gas-rich disks known to be associated with
T Tauri stars in local young stellar groups \citep[the others orbit TW Hya, MP 
Mus, and T Cha;][]{kastner97,kastner10,sacco13}.
Second, the proximity of 
V4046 Sgr is a substantial practical advantage in measuring key evolutionary 
diagnostics: the same observations would be $4\times$ more sensitive and probe 
2$\times$ smaller spatial scales for V4046 Sgr compared to disks around 
younger T Tauri stars that are associated with the nearest star-forming 
clouds (at distances $\sim 140$\,pc).  And third, from the perspective of a 
stellar host, V4046 Sgr introduces some interesting environmental issues that 
could influence the evolution of its disk.  Aside from dynamical clearing due to
its stellar multiplicity, the binary at the center of the V4046 Sgr disk makes 
for an unusual dichotomy.  The combined mass of the two central stars means that
dynamical timescales in the disk are relatively short.  However, the 
irradiation environment is not much different than for a single star, so 
timescales tied to thermal, energetic, or chemical processes are comparatively 
unaffected.  In essence, the evolutionary behavior of the V4046 Sgr disk can be 
described as a hybrid of a typical Herbig Ae disk and T Tauri disk, depending on
the relevant timescale that dominates a given evolution mechanism.

In this article, we present new, sensitive, high angular resolution observations
of the 1.3\,mm continuum and $^{12}$CO/$^{13}$CO/C$^{18}$O $J$=2$-$1 line 
emission from the V4046 Sgr circumbinary disk.  Using these data and a suite of 
radiative transfer tools, we aim to construct a preliminary, representative 
model of the disk structure in an effort to help characterize the observational 
signatures of different disk evolution mechanisms.  In the following sections, 
we describe our observations with the Submillimeter Array (SMA) and the relevant
data calibration procedures (\S 2), present some basic observational results 
(\S 3), develop models for the disk structure (\S 4), and comment on their 
implications for our understanding of disk evolution (\S 5).  A summary is 
provided in \S 6.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Observations and Data Reduction} \label{sec:observations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We observed V4046 Sgr with the Submillimeter Array \citep[SMA;][]{ho04} on
Mauna Kea, Hawaii on four occasions in 2009 and 2011.  These observations and
their calibration were already described by \citet{rodriguez10} and
\citet{rosenfeld12}, but a brief summary of the key points is provided here for
completeness.  In these observations, the individual 6-m array elements were
arranged in each of the four available SMA configurations, spanning baseline 
lengths from 8 to 509\,m.  The dual-sideband receiver backends and SMA 
correlator were configured with a local oscillator (LO) frequency of 
225.360\,GHz (1.33\,mm) and a $\sim$2\,GHz-wide intermediate frequency (IF) 
band $\pm$4-6\,GHz from the LO: in 2011, a second IF band was included 
$\pm$6-8\,GHz from the LO.  Each sideband/IF band combination was composed of 
24 spectral chunks of 104\,MHz width (although typically only the central 
82\,MHz are used).  In the first IF band, three chunks (in each sideband) were 
split into 512 spectral channels, to sample the CO isotopologue emission at 
200\,kHz ($\sim$0.25\,km s$^{-1}$) resolution.  All other chunks were coarsely 
split into 32 channels (3.25\,MHz each) to observe the continuum.  Observations 
of V4046 Sgr were interleaved with visits to J1924-292 ($\sim$15\degr\ away) on 
a 5-15\,minute cycle, as well as J1733-130 (22\degr\ away) on a 
$\sim$45\,minute cycle.  Additional observations of 3C 454.3 and available 
planets/satellites (Neptune, Ceres, Callisto) were conducted for calibration 
purposes.  A summary of relevant observational parameters is provided in Table 
\ref{tab:obs}.

The visibility data in each IF band and from each SMA observation were
calibrated independently using the {\tt MIR} package, as described by
\citet{rosenfeld12}.  After confirming their consistency over all IF/sideband
combinations and on overlapping baseline lengths, the continuum data were
spectrally averaged.  Spectral visibilities that cover each CO isotopologue
transition were continuum-subtracted and combined.  The overall data quality is
exceptional for the SMA and the low declination of V4046 Sgr, due primarily to 
the excellent observing conditions (precipitable water vapor levels were only 
$\sim$1\,mm throughout all of the observations).  Synthesis images were made 
using the {\tt CLEAN} deconvolution algorithm in the {\tt MIRIAD} software 
package for each emission tracer.  For the 1.3\,mm continuum, we emphasized the 
emission on smaller spatial scales with a Briggs robust (-1) weighting scheme, resulting in a map with a
$0\farcs74\times0\farcs38$ synthesized beam (at P.A. = 14\degr) and an RMS
noise level of 1.0\,mJy beam$^{-1}$.  Channel maps of the $^{12}$CO line 
emission were generated with natural weighting to produce a 
$1\farcs1\times0\farcs9$ beam with an RMS noise level of 40\,mJy beam$^{-1}$ in
25 binned 0.4\,km s$^{-1}$ velocity channels, centered on the systemic LSR
velocity, $+$2.87\,km s$^{-1}$ \citep{rodriguez10}.  Analogous channel maps of 
the CO isotopologue lines were made by employing a 0\farcs5 Gaussian taper, 
producing a slightly larger synthesized beam ($1\farcs5\times1\farcs2$) and a 
similar RMS noise level ($\sim$35\,mJy beam$^{-1}$).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results} \label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A summarized representation of the SMA observations of V4046 Sgr is provided in 
Figure \ref{fig:summary}.  The synthesized map of 1.3\,mm dust continuum 
emission is shown in Figure \ref{fig:summary}(a), with contours starting at 
5\,mJy beam$^{-1}$ (5\,$\sigma$) and increasing at 10\,mJy beam$^{-1}$ 
(10\,$\sigma$) intervals.  The integrated flux density recovered in this map is 
$283\pm28\,$mJy (dominated by a 10\% systematic calibration uncertainty), 
consistent with a reasonable extrapolation from single-dish submillimeter 
photometry measurements \citep{jensen96}.  We find the mm-wave emission is 
concentrated in a bright, narrow ring centered at the mean stellar position 
\citep[$\alpha = 18^{\rm h}16^{\rm m}10\fs49$, $-32\degr47\arcmin34\farcs50$, 
J2000;][]{zacharias10}, with a double-peaked morphology due to limb brightening 
at the projected ring ansae \citep[see][]{andrews11}.  The apparent rotation 
between the semi-major axis of the disk and the line joining the two ansae 
is mostly an artifact of the $uv$ sampling and is reproduced in our models (see 
\S \ref{sec:model3}).  These emission peaks are separated by 0\farcs75 
($\sim$55\,AU) and have peak intensities of $50\pm5$\,mJy beam$^{-1}$ (dominated
by a 10\%\ calibration uncertainty; S/N $\approx$ 50).  The western peak appears
slightly brighter ($\sim$5\,mJy beam$^{-1}$) than its eastern counterpart, but a
proper evaluation of the significance of this discrepancy requires a more 
detailed analysis (see \S 5).  The emission ring itself is at best only 
marginally resolved, implying a width smaller than the 0\farcs4 ($\sim$30\,AU) 
minor axis of the synthesized beam.  Figure \ref{fig:summary}(b) shows the 
azimuthally-averaged profile of the real and imaginary continuum visibilities as
a function of deprojected baseline length, constructed assuming the disk viewing
geometry derived by \citet{rosenfeld12} ($i=33\fdg5$, PA$=76$\degr).  The 
visibility amplitudes exhibit the distinctive oscillation pattern that is 
characteristic of an emission ring, with nulls at $\sim$150 and 
350\,k$\lambda$.  The zero-spacing amplitude is estimated to be $\sim$315\,mJy, 
suggesting that roughly 10\%\ of the total flux density from the disk was 
filtered out of the synthesized map in Figure \ref{fig:summary}(a): presumably 
that emission is distributed on larger scales with low surface brightness.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{V4046Sgr2013/f01.eps} \caption{An observational summary of the V4046
Sgr disk.  (a) An image of the 1.3\,mm dust continuum emission, with contours
drawn at 10\,mJy beam$^{-1}$ ($10$\,$\sigma$) intervals, starting at 5\,mJy
beam$^{-1}$.  The synthesized beam is shown in the lower left corner.  (b) The
azimuthally averaged real ({\it black}) and imaginary ({\it gray}) components
of the 1.3\,mm continuum visibilities as a function of deprojected baseline
length.  (c) The broadband SED, with a composite photosphere model for the two
stars marked as a thin gray curve \citep[see][]{rosenfeld12}.  The thick gray
curve shows the {\it Spitzer} IRS spectrum.  (d) The velocity-integrated
intensities (0$^{\rm th}$ moment; contours) overlaid on the intensity-weighted
velocities (1$^{\rm st}$ moment; colors) for the $^{12}$CO $J$=2$-$1 line
emission.  Intensity contours start at 0.16\,Jy beam$^{-1}$ km s$^{-1}$ and
increase in 0.24\,Jy beam$^{-1}$ km s$^{-1}$ steps, and the color scale spans
an LSR velocity width of $\pm$3\,km s$^{-1}$ from the systemic value.  (e) The
same as (d), but for the $^{13}$CO $J$=2$-$1 line emission.  Intensity contours
are drawn at 0.15\,Jy beam$^{-1}$ km s$^{-1}$ intervals, starting at 0.10\,Jy
beam$^{-1}$ km s$^{-1}$.  (f) The integrated line profiles of the $^{12}$CO
({\it black}), $^{13}$CO ({\it red}), and C$^{18}$O $J=2-1$ ({\it blue}; not
detected) emission inside square regions 12\arcsec\ on a side.
\label{fig:summary}}
\end{figure}

These SMA continuum measurements demonstrate that the V4046 Sgr binary joins 
the growing ranks of stars that host massive protoplanetary ``transition" disks 
\citep[e.g.,][]{andrews11}, with a central cavity of radius $\sim$29\,AU (see 
\S \ref{sec:model3}) that is substantially depleted of mm-sized dust particles. 
This classification is commensurate with the SED for V4046 Sgr, shown in Figure 
\ref{fig:summary}(c) and constructed from photometry in the literature 
\citep{hutchinson90,weaver92,jensen96,jensen97,rodriguez10,oberg11}, the 2MASS 
\citep{skrutskie06} and {\it IRAS} \citep{beichmann88} point source catalogs, 
and an archival {\it Spitzer} IRS spectrum.  The SED features the standard 
signature of a transition disk, with a distinctive ``dip" in the continuum near 
10\,$\mu$m suggesting that small, warm dust is preferentially depleted (although
{\it not} absent; see \S\ref{sec:model1}) near the central binary.  An accurate 
determination of the size of this putative cavity from the SED alone is not 
trivial \citep[e.g.,][]{calvet02}, although the very weak excess in the 
near-infrared makes it clear that the dust optical depths must be significantly 
diminished within at least a few AU of the central stars.  Note that the 
observed dip in the continuum covers a much wider wavelength range than was 
inferred by \citet{jensen97}.  Simulations of interactions between a stellar 
binary and its circumbinary disk suggest that the disk material should be 
inwardly truncated at a radius $\sim$2-5$\times$ the binary separation 
\citep[e.g.,][]{artymowicz94}: for V4046 Sgr, the disk truncation should occur 
at $\sim$0.2\,AU, much smaller than the continuum data suggest.  Therefore, the 
observed dust cavity is {\it not} related to interactions with the central 
binary.

The gas phase of the V4046 Sgr disk traced by CO isotopologue emission lines is
represented in Figures \ref{fig:summary}(d)-(f).  The $^{12}$CO and $^{13}$CO 
$J$=2$-$1 emission are displayed as moment maps in Figures \ref{fig:summary}(d) 
and (e).  The velocity-integrated intensities (0${\rm th}$ moment) are shown as 
contours at 3\,$\sigma$ intervals, starting at 2\,$\sigma$ (RMS noise levels 
are 0.08\,Jy beam$^{-1}$ km s$^{-1}$ for $^{12}$CO, 0.05\,Jy beam$^{-1}$ km 
s$^{-1}$ for $^{13}$CO), overlaid on the intensity-weighted velocities (1$^{\rm 
st}$ moment), with colors marking the velocity shift relative to line center.  
These maps exhibit the standard pattern of rotation, more clearly manifested in 
the individual channel maps shown together in Figure \ref{fig:chmaps}.  Line 
emission is firmly detected ($>$3\,$\sigma$, or 0.12\,Jy beam$^{-1}$ in each 
0.4\,km s$^{-1}$ channel) out to $\pm4.4$ or $\pm4.0$\,km s$^{-1}$ from the 
line center for $^{12}$CO and $^{13}$CO, respectively.  Given the V4046 Sgr 
stellar mass and disk inclination angle \citep[$i = 33\fdg5$, PA$ = 76$\degr 
with an ambiguity in the absolute orientation;][]{rosenfeld12}, those maximal 
projected velocities correspond to disk radii of $\sim$25-30\,AU, which is 
similar to the size of the dust cavity.  Therefore, these data do not have 
sufficient sensitivity in the line wings to rule whether or not there is 
CO inside the dust cavity.  The integrated line intensities derived from the 
0$^{\rm th}$ moment maps are $34.5\pm3.5$\,Jy km s$^{-1}$ for $^{12}$CO and 
$9.4\pm0.9$\,Jy km s$^{-1}$ for $^{13}$CO.  The peak intensities in the channel 
maps are $0.95\pm0.10$\,Jy beam$^{-1}$ ($22\pm 2$\,K; peak S/N $\sim$20) and 
$0.47\pm0.08$\,Jy beam$^{-1}$ ($6.5\pm0.8$\,K; peak S/N $\sim$11), respectively.
The C$^{18}$O emission is faint, and at best only marginally detected.  We 
estimate an integrated intensity of $\sim$0.6\,Jy km s$^{-1}$ from a 
0$^{\rm th}$ moment map, but suggest that this number be treated with caution: 
there is no firm detection in individual channel maps.  For reference, spatially
integrated spectra for the CO isotopologue lines are shown together in Figure 
1(f).

\begin{figure}[t!]
\epsscale{0.95}
\plottwo{V4046Sgr2013/f02.eps}{V4046Sgr2013/f03.eps}
\caption{Naturally weighted channel maps of the $^{12}$CO $J=2-1$ 
({\it left}) and $^{13}$CO $J=2-1$ ({\it right}) emission towards the V4046 
Sgr disk.  Channels are 0.4\,km s$^{-1}$ wide with the synthesized beam 
marked in the bottom left corner.  Contour levels are drawn at intervals of 
0.12 Jy\,beam$^{-1}$ ($3\,\sigma$).
\label{fig:chmaps}}
\end{figure}

One of the notable features of Figure 1 is the disparity in the apparent 
spatial extents of the CO line and dust continuum emission.  The optically thick
$^{12}$CO line emission spans a diameter of 10\arcsec\ on the sky, corresponding
to a projected disk radius of $\sim$365\,AU.  Comparing Figures 1(a) and (d), we
find the dust ring extent is roughly 5$\times$ smaller.  The fainter, optically 
thinner $^{13}$CO emission still subtends $\sim$8\arcsec\ on the sky, 
$\sim$4$\times$ larger than the 1.3\,mm continuum.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Modeling Analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Having highlighted the basic observational characteristics of the gas and dust
tracers in the V4046 Sgr circumbinary disk, we now move to a more quantitative
effort to explore what the data can tell us about the disk structure.  We
concentrate our analysis on extracting a model for the radial distributions of
gas and dust based on the observational results described in \S 3.  To do that,
we first describe a flexible modeling framework for constructing disk 
structures given an arbitrary surface density profile and dust-to-gas ratio (\S
4.1), generalized from a similar analysis of the TW Hya disk 
\citep{andrews12}.  Using this formalism, we then consider in turn the SED 
and the resolved 1.3\,mm continuum observations and present model disk 
structures that optimally account for these data (\S 4.2.1 and 4.2.2) under the 
assumption of a constant dust-to-gas ratio.  After weighing the successes and 
failures of these two models, we relax the latter assumption and construct a 
model (\S 4.2.3) that is more consistent with all of the data: the broadband 
SED, the resolved dust continuum emission, and the resolved line emission from 
two CO isotopologues.  The ultimate success of this latter model provides some
insights on the shortcomings of our typical assumptions when analyzing disk
structures, which are discussed further in \S 5.

\subsection{Physical and Practical Overview}

Here we describe a generic prescription for constructing a parametric model of 
the two-dimensional density and temperature structure of both dust and gas in a 
protoplanetary disk, which builds on the previous work by 
\citet{andrews09,andrews10,andrews11,andrews12}.  We assume an axisymmetric,
two-dimensional model in a cylindrical reference frame with coordinates ($r$,
$z$).

Since dust grains dominate the opacity in a disk, we construct a model starting
with the distribution of its constituent solids.  In a general case with
different dust populations, the density distribution is simply defined as the
discrete sum $\rho_{\rm dust}(r,z) = \sum_j \rho_j(r,z)$, where
\begin{equation}
\rho_j(r,z) = \frac{\Sigma_j(r)}{\sqrt{2\pi} H_j(r)} \exp \left[-\frac{1}{2} \left(\frac{z}{H_j(r)}\right)^2\right]. 
\end{equation}
In Eq.~(1), $\Sigma_j(r)$ is the surface density profile and $H_j(r)$ is the
scale height profile for each dust population (labeled by index $j$).  To keep
the problem tractable, we consider only two dust populations in this study: a 
``midplane" population that dominates the disk mass and is composed of larger
grains that are vertically settled, and a less-abundant ``atmosphere" population
of smaller grains that is distributed to larger heights from the midplane.  We 
define parametric scale height profiles 
\begin{eqnarray}
H_{\rm atm}(r) &=& H_0 \,\, (r/r_0)^{\psi} \\
H_{\rm mid}(r) &=& \chi \,\, H_{\rm atm}(r),
\end{eqnarray}
where $H_{0}$ is the scale height at $r = r_0$, $\psi$ is the flaring index,
and $\chi$ is a scaling factor (in the range 0 to 1) that mimics dust 
settling: we assign a fixed $\chi = 1/2$ for simplicity.  The absorption and 
scattering opacities for each dust population were computed with a Mie code, 
assuming segregated spherical particles with the \citet{pollack94} mineralogical
composition and optical constants and a size distribution $n(a) \propto 
a^{-3.5}$ between 5\,nm and a maximum size (here $a$ is the grain radius).  In
this specific case, we fix $a_{\rm max}({\rm atm}) = 10$\,$\mu$m and
$a_{\rm max}({\rm mid}) = 1$\,cm, which reproduces the shape of the 
(sub)mm-wavelength SED well (see models described in \S\ref{sec:model1}, 
\ref{sec:model2}, and \ref{sec:model3}).

For any functional form for the surface densities, $\Sigma_j(r)$, the
two-dimensional distribution of dust densities (and optical depths, accounting
for the grain properties) can be specified with the above prescription.  Given
$\rho_{\rm dust}(r,z)$, we calculate the corresponding temperature structure,
$T_{\rm dust}(r,z)$, using the Monte Carlo radiative transfer code {\tt
RADMC-3D}\footnote{\url{http://www.ita.uni-heidelberg.de/$\sim$dullemond/software/radmc-3d/}}, assuming an incident radiation field from the central binary.
For the latter, we use a composite spectrum of two \citet{lejeune97} model
photospheres with \{$T_{\rm eff,1} = 4350$\,K, $L_{\ast,1} = 
0.35$\,L$_{\odot}$, $M_{\ast,1} = 0.90$\,M$_{\odot}$\} and \{$T_{\rm eff,2} = 
4060$\,K, $L_{\ast,2} = 0.25$\,L$_{\odot}$, $M_{\ast,2} = 0.85$\,M$_{\odot}$\} 
\citep[for details, see][]{rosenfeld12}.

To construct a corresponding gas disk structure, we define a vertically 
integrated dust-to-gas mass ratio $\zeta$ that may vary spatially, such that 
$\Sigma_{\rm gas}(r) = \Sigma_{\rm dust}(r)/\zeta(r)$, where $\Sigma_{\rm 
dust}(r) = \sum_j \Sigma_j(r)$.  The two-dimensional density structure of the 
gas, $\rho_{\rm gas}(r, z)$, is determined by numerically integrating the 
differential equation that describes vertical hydrostatic equilibrium for a 
given gas temperature distribution, $T_{\rm gas}(r, z)$ \citep[see][their 
Eq.~5]{andrews12}.  At the midplane ($z = 0$), the gas and dust are assumed to 
be thermally coupled: $T_{\rm gas}(r, z = 0) = T_{m} = T_{\rm dust}(r, z = 
0)$.  At larger vertical heights, the gas temperatures are permitted to deviate 
from the dust temperatures.  To facilitate those departures, we adopt the 
parameterization introduced by \citet{dartois03},
\begin{equation}
 T_{\rm gas}(r, z) = \left\{
\begin{array}{ll}
T_a + \left(T_m - T_a\right) \left[\cos{\frac{\pi z}{2 z_q}}\right]^{2\delta} & \mbox{if $z < z_q$} \\
T_a & \mbox{if $z \ge z_q$} 
\end{array},
\right. 
\end{equation}
where $T_a = T_{a,0} (r/r_0)^{-q}$ is a parametric temperature profile for the
gas in the disk atmosphere, which is explicitly defined here as heights larger
than a fiducial value, $z_q$.  We fix $z_q = 2 H_{\rm gas} = 2 c_s/\Omega$,
where $H_{\rm gas}$ is the hydrostatic gas scale height evaluated at the
midplane temperature ($T_m$), defined as the ratio of the sound speed ($c_s$)
to the Keplerian angular velocity ($\Omega$).  The index parameter $\delta$
determines the shape of the vertical temperature gradient of the gas, and is 
fixed here to $\delta = 2$, following \citet{dartois03}.  Finally, to insure a 
physically realistic model, we impose the additional criterion that the gas 
cannot be colder than the dust: if $T_{\rm gas}(r, z)$ calculated using Eq.~(4) 
is less than $T_{\rm dust}(r, z)$ computed from the radiative transfer 
simulation, we set $T_{\rm gas} = T_{\rm dust}$.  In practice, this formulation
provides a straightforward means of permitting extra gas heating relatively 
near the star, while (if desired) maintaining thermal coupling between the gas
and dust at all heights at large disk radii ($r \gtrsim 100$\,AU; see \S 4.2).
For the kinematic structure of the disk, we assume the gas is in Keplerian
rotation around the central stars, 
${v_{\theta}}^2=GM_\ast/r$, 
and that the emission line profile widths are determined from the quadrature sum
of a thermal and turbulent broadening term, with a fixed and constant value for 
the latter of $\xi = 0.1$\,km s$^{-1}$.

Having defined the gas disk structure, we need to assign an abundance
distribution for the tracer molecule, CO.  Following \citet{aikawa99}, we
assume that 80\%\ of the gas is composed of H$_2$ and then assume a {\it
constant} fractional abundance $f_{\rm co} = n(^{12}$CO)/$n({\rm H}_2)$ such
that
\begin{equation}
n(^{12}{\rm CO}) = 0.8 f_{\rm co} \frac{\rho_{\rm gas}(r, z)}{\mu m_{\rm H}}
\end{equation}
in the region of the disk we term the ``abundant layer".  In Eq.~(5), 
$\mu$ is the mean molecular weight of the gas ($\mu = 2.37$) and $m_{\rm H}$ is 
the mass of a hydrogen atom.  The abundant layer is defined based on the 
treatment of \citet{qi08,qi11}, which assumes CO is present in the gas phase if:
(1) the local gas temperatures are high enough that the CO is not frozen onto 
the dust grains, and (2) the vertically-integrated column density is higher than
the penetration depth of photodissociating radiation.  At each radius,
Eq.~(5) is used to compute the CO densities if
\begin{eqnarray}
T_{\rm frz} &\le& T_{\rm gas}(r, z) \\
\sigma_s &\le& f_{\rm H} \int_z^{\infty} n_{\rm gas}(r, z^{\prime}) \,\, dz^{\prime} 
\end{eqnarray}
are valid: here, $T_{\rm frz}$ is the CO freezeout temperature, $\sigma_s$ is
the photodissociation column, and $f_{\rm H}$ is the fraction of H nuclei in 
the gas.  Guided by \citet{aikawa99} and \citet{qi11}, we assume $f_{\rm H} = 
0.706$ and fix $T_{\rm frz} = 19$\,K and $\sigma_s = 
5\times10^{20}$\,cm$^{-2}$.  At locations where Eqs.~(6) and (7) are not met, 
we sharply (and arbitrarily) reduce the $^{12}$CO abundance by scaling 
$f_{\rm co}$ down by a factor of $10^8$.  The CO isotopologue abundances are 
assumed to follow $^{12}$CO with the interstellar medium ratios inferred by 
\citet{wilson99}: $n(^{12}$CO)/$n(^{13}$CO$) = 69$ and 
$n(^{12}$CO)/$n($C$^{18}$O$) = 557$.

The formalism described above and encapsulated in Eqs.~(1)-(7) fully defines the
dust and gas structure of a model disk.  In practice, the dust structure 
depends on some description of the surface density profiles for each dust 
species, $\Sigma_j(r)$, and their corresponding vertical height profiles, 
$H_j(r)$: here, the latter is fully specified with two parameters, \{$H_0$, 
$\psi$\}.  Likewise, the gas structure is determined by a dust-to-gas mass ratio
profile $\zeta(r)$, and three additional free parameters: \{$T_{a,0}$, $q$\} to 
define the atmosphere temperature profile and \{$f_{\rm co}$\} to assign the CO 
molecular abundance.  For a given model, we generate synthetic data products to 
compare with the observations and evaluate our parameter choices and the 
underlying model characterization.  For the dust, we use the ray-tracing 
capability of {\tt RADMC-3D} to produce a synthetic SED and set of 1.3\,mm
continuum visibilities sampled at the same spatial frequencies as observed by
the SMA.  For the gas, we use the molecular excitation and line radiation 
transfer code {\tt LIME} \citep{brinch10} to calculate the non-local 
thermodynamic equilibrium level populations based on information from the LAMDA
database \citep{schoier05} and then calculate synthetic spectral visibilities
for each of the CO isotopologues in the velocity channels and spatial 
frequencies sampled by the SMA.  In all models, we assume the disk inclination
($i = 33\fdg5$), major axis position angle (PA = 76\degr), total stellar mass
($M_{\ast} = 1.75$\,M$_{\odot}$), and disk center position derived by
\citet{rosenfeld12}, as well as the distance ($d = 73$\,pc) inferred by
\citet{torres06,torres08}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Model Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The modeling framework we have constructed is relatively complex, with some
fairly severe parameter degeneracies \citep[for a more detailed discussion,
see][]{andrews09,andrews11,andrews12,qi11}.  Moreover, the computational
effort required for the radiative transfer and non-LTE excitation calculations
remains costly, severely limiting our capability to use either stochastic or
deterministic optimization algorithms to robustly probe the parameter-space.
In practice, we build structure models manually, using small explorations of
each parameter to reproduce the key observational features of interest.  This
means that the models presented below are not quantitatively optimized, nor
unique.  However, they are still of great interest since they serve as
qualitative illustrations of key features in the V4046 Sgr disk structure that
are not otherwise easily accessible.  In the following sections (\S 4.2.1 and
4.2.2), we aim to provide a pedagogical guide to the logic behind the more
complex disk structure we ultimately will advocate for the V4046 Sgr disk (\S
4.2.3).

\subsubsection{Model to Reproduce the SED}\label{sec:model1}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{V4046Sgr2013/f04.eps}
\caption{Results for the modeling effort focused on the SED (\S 4.2.1).  (a)
Surface density profiles for the gas ({\it dashed}), the large ``midplane"
({\it green}) and small ``atmosphere" ({\it blue}) grain populations.  (b) The
SED ({\it black points} and {\it solid gray curve}) compared with the model
({\it red}).  The stellar photosphere model is shown as a dashed gray curve.
(c) The deprojected, azimuthally-averaged (real) 1.3\,mm continuum visibility
profile, with model overlaid (in {\it red}); gray points mark the residuals.
(d) The observed, simulated, and residual 1.3\,mm emission synthesized maps.
Contours are drawn at 5\,mJy beam$^{-1}$ (5\,$\sigma$) intervals.  Note the
smaller map size compared to Figure \ref{fig:summary}.  (e) A comparison of the
observed and simulated $^{12}$CO moment maps.  (f) The same as panel (e), but
for the $^{13}$CO line.  Contour levels and map sizes for panels (e) and (f)
are the same as in Figure \ref{fig:summary}.  \label{fig:modsum1}}
\end{figure}

As a starting point, we focus on building a structure model that is designed to
reproduce the broadband SED of V4046 Sgr, with some input from the CO
observations to guide our description of the gas disk structure.  Following the
modeling prescription for transition disks described by \citet{andrews11}, we
adopt a truncated version of the \citet{lynden-bell74} similarity solution for
a thin, Keplerian accretion disk with a time-independent, power-law viscosity
profile to describe the surface densities of the dust \citep[see
also][]{hartmann98},
\begin{equation}
 \Sigma_{\rm dust}(r) = \left\{
\begin{array}{ll}
\Sigma_{ss}(r) = \Sigma_c \left(\frac{r}{r_c}\right)^{-\gamma} \exp \left[-\left(\frac{r}{r_c}\right)^{2-\gamma}\right]  & \mbox{if $r \ge r_{\rm cav}$} \\
\Sigma_{\rm cav} & \mbox{if $r_{\rm in} \le r < r_{\rm cav}$} 
\end{array},
\right. 
\end{equation}
where $r_c$ is a characteristic radius, $\gamma$ is an index parameter, 
$\Sigma_c = e \cdot \Sigma_{\rm dust}(r_c)$, $r_{\rm cav}$ is a ``cavity" 
radius, $\Sigma_{\rm cav}$ is a constant, and $r_{\rm in}$ is the inner edge of
the model.  Both grain populations were forced to follow the behavior in 
Eq.~(8): outside the disk cavity, 90\%\ of the mass was apportioned to the
midplane grains ($\Sigma_{\rm mid} = 0.9\,\Sigma_{\rm dust}$, $\Sigma_{\rm atm} 
= 0.1\,\Sigma_{\rm dust}$, for $r \ge r_{\rm cav}$), but the tenuous material
inside the cavity was assumed to be all atmosphere grains ($\Sigma_{\rm mid} = 
0$, $\Sigma_{\rm atm} = \Sigma_{\rm cav}$, for $r < r_{\rm cav}$).  Since the
SED is composed of unresolved photometric measurements, it has no ability to
constrain the gradient $\gamma$.  Therefore, we fixed $\gamma = 1$, a typical
value for disks \citep{andrews09,andrews10}.  The inner radius of the model was
set to $r_{\rm in} = 0.2$\,AU, the outer edge of the zone expected to be
cleared by dynamical interactions with the central binary
\citep{artymowicz94}.  For the gas phase, we assumed a standard, spatially 
uniform dust-to-gas ratio, $\zeta(r) = \zeta = 0.01$ 
\citep{pollack94,dalessio01}.

After some iteration between dust and gas structures, we identified a model
that can reproduce well both the broadband SED and SMA observations of CO line
emission: the results are highlighted in Figure \ref{fig:modsum1}.  Using the 
morphology of the continuum ``dip" in the infrared SED, we inferred a cavity 
radius $r_{\rm cav} = 3$\,AU, roughly 15$\times$ larger than could be explained 
by tidal stripping from the V4046 Sgr binary.  As with most other transition 
disks, this cavity is not empty: the weak infrared excess (and weak silicate 
emission feature) noted in Fig.~\ref{fig:modsum1}b is accommodated with 
$\Sigma_{\rm cav} \approx 10^{-4}$\,g cm$^{-2}$.  In the mid-infrared, the 
excess spectrum was fit with a scale height of $H_0 = 0.4$\,AU at $r_0 = 
10$\,AU and a flaring index $\psi = 1.25$ \citep[although the latter is not 
well-determined in transition disks; see][]{andrews11}.  The total mass of the 
dust structure (the integral of Eq.~8 over the disk area) was $M_{\rm dust} 
\approx 9\times10^{-4}$\,M$_{\odot}$, determined from the luminosity of the 
millimeter-wave SED.  The characteristic size of the disk was estimated to be 
$r_c = 75$\,AU, based on the extent of the $^{12}$CO emission.  Those data were 
described well with a CO abundance $f_{\rm co} = 3\times10^{-6}$ and a steep 
atmospheric temperature profile with $T_0 = 200$\,K (at $r_0 = 10$\,AU) and 
$q = 0.8$.  In practice, this $T_a(r)$ profile means the gas and dust are 
thermally coupled for $r \gtrsim 150$\,AU, but there is some source of 
additional gas heating in the disk atmosphere at smaller radii.  One possible 
source for that heating is the strong X-ray emission from the central binary 
(see \S\ref{sec:discussion}).

Although by design this model provides a good match to the SED and CO emission,
it should be obvious from Figure \ref{fig:modsum1} that it is
irreconcilable with the resolved 1.3\,mm continuum data.  However, these 
failures are instructive in refining the model.  The residuals in Figures 
\ref{fig:modsum1}(c)-(d) point out two key issues.  First, the cavity size 
derived from the SED alone substantially under-estimates the size that can be 
{\it directly} measured from the resolved 1.3\,mm data.  In \S 3, we estimated 
a cavity radius of $\sim$29\,AU (and will quantify that further below), 
$\sim$10$\times$ larger than derived here.  This kind of discrepancy is not 
uncommon for transition disks \citep[e.g., DM Tau;][]{andrews11}: any inference 
of a size scale from unresolved observations is inherently uncertain and may 
be a signature of how the dynamics of grains depends upon their size (see 
\S\ref{sec:discussion}).  Second, 
the spatial extent of this model is much larger than the 1.3\,mm continuum 
distribution.  This latter point is manifested in the poor match to the 
visibility profile in Figure \ref{fig:modsum1}(c) on large spatial scales 
(short baselines), where the model is clearly more resolved than the data 
permit (the negative residuals outside the emission ring in Figure 
\ref{fig:modsum1}(d) represent the same aspect of this model failure).  Since 
$r_c$ was determined from the CO data, this discrepancy is not surprising: we 
already highlighted in \S 3 how the CO emission is much more extended than the 
dust emission (see Fig.~\ref{fig:summary}).

\subsubsection{Model to Reproduce the Resolved 1.3\,mm Continuum Emission}
\label{sec:model2}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{V4046Sgr2013/f05.eps}
\caption{Results for the modeling effort focused on the resolved 1.3\,mm 
dust continuum emission (first model described in \S 4.2.2).  See Figure 
\ref{fig:modsum1} for a description of the contents of individual panels.  
\label{fig:modsum2a}}
\end{figure}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{V4046Sgr2013/f06.eps}
\caption{Results for the modeling effort focused on the resolved 1.3\,mm 
dust continuum emission (second model described in \S 4.2.2).  See Figure 
\ref{fig:modsum1} for a description of the contents of individual panels.  
\label{fig:modsum2}}
\end{figure}

Informed by the failures of the previous model, we shift the focus here to
building a model that is more consistent with the resolved morphology of the 
1.3\,mm dust emission.  Two prescriptions for the surface density profile are 
presented, both of which can reproduce the size and shape of the mm-wave 
emission.  Considering the visibility profile (Fig.~\ref{fig:summary}a), the former 
is constrained by the short baselines 
($\mathcal{R}\lesssim150$\,k$\lambda$) and the latter is dictated by the 
position of the null and deep trough at higher spatial frequencies.  The results
shown in Figure \ref{fig:modsum2a} are for the same relative ratios of midplane 
and atmosphere grains and model parameterization defined in \S\ref{sec:model2}, 
but with the the cavity size, $r_{\rm cav}$, enlarged to 21\,AU (i.e., adjusted
without regard to the infrared SED).  In addition, 
this model has $\gamma=-1.5$, $r_c=45$\,AU, $\psi = 1.5$, 
$H_0 = 0.45$\,AU at $r_0 = 10$\,AU, and 0.0011\,$M_\odot$ of dust.  Note that a 
negative $\gamma$ implies a density profile that {\it rises} with 
radius out to $r_c$, but then is sharply truncated.  In essence, it naturally
produces a relatively narrow ``ring" of material \citep{isella09}, making it a
practical means of describing the resolved morphologies of transition disks 
\citep[e.g.,][]{isella10,isella12,andrews11b,brown12}.  We also 
consider an alternative parameterization of the surface density profile where 
the viscous disk similarity solution, $\Sigma_{\rm dust}(r) = \Sigma_{ss}(r)$ 
(see Eq.~8), is valid for all $r$. Figure \ref{fig:modsum2} summarizes this 
second model which features a large, negative $\gamma = -3.5$ with 
$r_c = 48$\,AU, $\psi = 1.5$, $H_0 = 0.4$\,AU at $r_0 = 10$\,AU and 
0.0010\,$M_\odot$ of dust.

Although both of these models reproduces well the morphology of the 1.3\,mm 
emission ring (Figs.~\ref{fig:modsum2a},\ref{fig:modsum2}c,d), they each 
fail to account for the observations of the V4046 Sgr disk in three similar, 
illuminating ways.  We also experimented with alternative prescriptions for the
surface densities (e.g., a power-law with an outer edge cut-off), and found 
qualitatively similar results.  First, they produce too
deep and wide of a ``dip" in the infrared SED in Figures 
\ref{fig:modsum2a},\ref{fig:modsum2}(b), because there is not enough dust 
interior to the density peak.  Note that the mid/far-infrared
excess of the models is a manifestation of this lack of inner disk material: in 
the absence of an inner disk, a larger surface area of more distant material in 
the model ring can be directly illuminated by the central stars, producing a 
``bump" in the continuum at $\sim$50\,$\mu$m.  Second, if we assume a standard, 
uniform dust-to-gas ratio as before ($\zeta = 0.01$), there is not enough gas 
outside the density peak in either of these models to account for the observed 
spatial extent of the CO emission 
(Figs.~\ref{fig:modsum2a},\ref{fig:modsum2}e,f).  For a constant $\zeta$, this 
size discrepancy is present regardless of the assumed CO abundance or gas 
temperature distribution, although the results in Figures \ref{fig:modsum2a} and
\ref{fig:modsum2} use the same $f_{\rm co}$ and $T_a(r)$ that were adopted in 
\S 4.2.1.  And third, the models do not quite generate enough 1.3\,mm dust 
emission on large angular scales (deprojected baselines $\le$50\,k$\lambda$; 
see Figs.~\ref{fig:modsum2a},\ref{fig:modsum2}c).  The origins of 
this last discrepancy are indeed subtle.  We are unable to account for all the 
emission through a simple scaling of the dust mass because the model ring is 
already nearly optically thick (this is exacerbated by our assumption of 
isotropic scattering off large, spherical grains).  Increasing the densities 
further effectively impedes the
radiation transfer to the midplane, making the dust cooler and actually {\it 
decreasing} the 1.3\,mm luminosity.  Instead, we interpret this emission 
mismatch as an intrinsic model deficiency: the addition of a low-density, 
spatially-extended dust component could reconcile the model visibilities with 
the data.  Note that this faint ($\sim$10-15\%\ of 
the total flux) emission ``halo" would be strongly spatially filtered in the 
synthesized images, as was found for the SMA observations in \S 3.

\subsubsection{A Hybrid Model to Reproduce All the Data}\label{sec:model3}

The three different models constructed above illustrate the key features of the 
V4046 Sgr disk structure.  To review those results, we have identified three 
basic structural elements that are required in any model prescription that aims 
to successfully reproduce the observations:
\begin{enumerate}
\item The vast majority of the large grain population must be strongly
concentrated in a {\it narrow} ring with a large central cavity to explain the
1.3\,mm continuum visibilities;
\item Inside that cavity, a reservoir of small dust particles (which produce
little mm-wave emission) at radii of a few AU are required to account for the
morphology of the infrared SED;
\item A large-scale, extended halo structure is necessary to explain the CO
line emission size and a faint, strongly spatially filtered 1.3\,mm continuum 
component, but there appears
to be substantially less mass in the large dust particles relative to the gas
outside the concentrated ring component.
\end{enumerate}
Here, our goal is to combine the successful aspects of the models described
above into a hybrid structure that incorporates these three elements and is
commensurate with all of the observations.

We define a surface density profile for the dust that is composed of three
individual components, each corresponding to one of the structural elements
enumerated above:
\begin{equation}
\Sigma_{\rm dust}(r) = \Sigma_{\rm ring}(r) + \Sigma_{\rm in}(r) + \Sigma_{\rm halo}(r).
\end{equation}
We elect to use simple Gaussian profiles for the distribution of material in the
ring and inner disk components,
\begin{equation}
\Sigma(r) = \frac{\Sigma_0}{\sqrt{2\pi}\sigma} \exp \left[-\frac{1}{2}\left(\frac{r-\mu}{\sigma}\right)^2\right],
\end{equation}
where we define parametric constants that correspond to peak radii \{$\mu_{\rm 
ring}$, $\mu_{\rm in}$\}, profile widths \{$\sigma_{\rm ring}$, $\sigma_{\rm 
in}$\}, and normalizations \{$\Sigma_{0, {\rm ring}}$, $\Sigma_{0, {\rm in}}$\}
for both structure components separately.  We assume that the ring component is
composed solely of the large ``midplane" grains, and that the inner component
is made up entirely of the small ``atmosphere" grains.  For the halo component,
we again assume a truncated similarity solution model (as in \S 4.2.1), but
eschew the sharp inner edge for one with a smoother, Gaussian taper:
\begin{equation}
\Sigma_{\rm halo}(r) = \Sigma_{ss}(r; \{\Sigma_c, r_c, \gamma\}) \, \exp \left[-\left(\frac{\mu_{\rm ring}}{r}\right)^2\right].
\end{equation}
The functional form of Eq.~(11) is arbitrary, and was selected entirely for
purposes of convenience in the radiative transfer modeling (tying the
exponential turnover scale to the ring component center makes a joint
exploration of the model space less cumbersome).  We fix $\gamma = 1$ as before 
and use a 6:1 mass ratio between the midplane and atmosphere grains.  The gas is
assumed to follow the small grains in both the halo and inner disk components
separately -- i.e. decoupled from the large grain population -- each with a 
ratio $\zeta$.  We use the halo gas profile as an upper limit on the surface 
density of midplane grains in the ring. For radii where the ratio between the 
two would exceed unity (e.g. in the space between the two Gaussian 
distributions), we reduce $\Sigma_{\rm ring}$ to equal the gas surface density 
from the halo component.  Either increasing the gas or reducing the dust density
has no significant effect upon our observables since the subtended region is 
very narrow and the densities are already low.

A model constructed with the recipe outlined above is superior to those 
presented in \S\ref{sec:model1} and \ref{sec:model2} in terms of reproducing all
the relevant observations of the V4046 Sgr transition disk, as
demonstrated in Figure \ref{fig:modsum3}.  In this case, we assumed that the 
same dust scale height distributions used in the previous sections were 
applicable to all of the structure elements.  For the ring component, we 
identify its center at $\mu_{\rm ring} = 37$\,AU and width $\sigma_{\rm ring} = 
7$\,AU (corresponding to a FWHM of 16\,AU, or 0\farcs2 projected on the sky; 
notably smaller than the SMA angular resolution), and allocate a large total 
dust mass of 0.004\,M$_{\odot}$.  The representative size of the 
central cavity is estimated to be $29$\,AU by subtracting one half of the FWHM 
from the ring radius.  In the inner disk, we find 
$\mu_{\rm in} = 4$\,AU, $\sigma_{\rm in} = 1$\,AU, and a total dust mass of
$2\times10^{-5}$\,M$_{\odot}$.  To account for the faint, extended 1.3\,mm 
emission component and the size of the CO emission, a halo component with a 
characteristic radius $r_c = 75$\,AU and total dust mass of 
$\sim$10$^{-4}$\,M$_{\odot}$ is sufficient.  To produce sufficient CO line 
intensities for the same gas temperature and molecular abundance structures
highlighted in \S 4.2.1, we set $\zeta = 0.0014$ in the outer disk, 
{\it relative to the halo component only}.  The weak C$^{18}$O emission of this 
model is consistent with the observations (i.e. a non-detection).  In the inner 
disk, we have kept 
$\zeta = 0.01$ fixed as before.  The total (gas+dust) mass in this model is 
$\sim$0.094\,M$_{\odot}$ and the disk-integrated dust-to-gas mass ratio is 
$\sim$0.047.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{V4046Sgr2013/f07.eps}
\caption{Results for the modeling effort that successfully reproduces all of 
the data (SED, 1.3\,mm continuum visibilities, and CO spectral images).  See 
Figure \ref{fig:modsum1} for descriptions of individual panels.  
\label{fig:modsum3}}
\end{figure}

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We have presented sensitive, high angular resolution SMA observations of the 
1.3\,mm dust continuum and spectral lines of three CO isotopes from the V4046 
Sgr circumbinary disk. We have modeled the dust and gas structures and 
identified three fundamental signatures of profound evolution in this disk.  
First, the majority of the large grains are distributed in a narrow ring 
centered creating a large central cavity ($r\sim29$\,AU).  Second, there exists 
a significant population of $\mu$m-sized grains that have not been cleared from 
the center of this cavity.  Third, the gas disk extends much farther than the 
compact dust ring, suggesting that this disk has a radial variation in the 
dust-to-gas mass ratio.

To explain the striking discrepancy between the size of the gas and dust 
distributions, we invoke a model with a strong radial variation in the relative 
mass ratio of the large, midplane grains and the gas.  The dust-to-gas mass
ratio in the model is large at the dust ring ($\zeta \sim 0.1$ -- $1$) and small
outside of it ($\zeta \sim 10^{-3}$).  Similar discrepancies between the gas and
dust distributions have been noted for other disks, such as those around IM Lup 
\citep{panic08}, TW Hya \citep{andrews12}, and LkCa 15 \citep{isella12}.
While our preferred model does feature an extended, but tenuous, dusty halo, 
models where the dust uniformly follows the gas disk either produce too much 
continuum or too little CO emission at large radii and clearly disagree with 
the observations (see \S\ref{sec:model1} and \S\ref{sec:model2}).  We caution 
that the model $^{12}$CO and $^{13}$CO line emission is optically thick and so 
the derived dust-to-gas mass ratios are upper limits.  Additionally, the dust 
mass in the model is sensitive to the dust opacities and so other values of the
dust-to-gas mass ratio and total disk mass are possible. The precise form and 
normalization of $\zeta(r)$ are model dependent, but the observations seem to 
suggest that it is a strongly decreasing function.

While the underlying physical cause of the discrepancy between the size of the 
gas and dust distributions is not clear, there is a 
natural explanation in the growth and inward radial drift of dust particles 
\citep{weidenschilling80,brauer08a,birnstiel10}.  Drift is a fundamental feature
in the migration of disk solids, generated because the radial pressure gradient
in the disk causes the gas to rotate at slightly sub-Keplerian velocities while 
the dust particles are on Keplerian orbits.  The resulting headwind creates a 
drag-force on the dust that relates to the size of the particle, preferentially
moving the large particles towards the central star \citep{weidenschilling77,
takeuchi02}.  As these particles drift inwards, their size distribution is 
continuously evolving as they encounter higher local densities that are 
conducive to growth \citep{birnstiel10}.  The collective effect is to shrink the
radial extent of the dust mass distribution, as the gas disk continues to
viscously spread to larger radii \citep{birnstiel12a}.  Since these processes 
operate on dynamical timescales, this phenomenon should be particularly 
pronounced for the V4046 Sgr disk due to the binarity (increased $M_*$) and 
advanced age of its stellar hosts.  Resolved observations taken across the 
(sub-)millimeter and radio bands may clarify whether the radial grain size 
distribution is consistent with this drift and particle growth scenario 
\citep{guilloteau11,perez12}.

In order to reproduce the observed morphology of the mm continuum emission, we 
required a model that has its large dust grain population distributed in a 
narrow ring.  This is suggestive of a physical mechanism acting to both
concentrate the particles radially and prevent them from moving further into 
the inner disk.  Particle growth alone is insufficient to produce mm-wave 
cavities \citep{birnstiel12b}, but this kind of ring-like dust concentration is 
a natural consequence of particle evolution models that also employ feedback 
from a large-scale maximum in the radial pressure gradient of the gas 
\citep{pinilla12a}.  The dynamics of the dust particles are strongly dictated by
the surrounding gas and while negative radial pressure gradients induce inward 
radial drift, a {\it positive} pressure gradient may stop the dust as it streams
towards the star.  This pressure maximum forms a ``pocket'' that collects dust 
and, if sufficiently broad and high-amplitude, can significantly alter the 
observed continuum emission \citep{pinilla12b}.

While many physical phenomena may generate these particle traps 
\citep[e.g.][]{barge95,klahr97,alexander07,johansen09}, we focus on an origin in
the density maximum at a gap edge produced by dynamical interactions between a 
low-mass companion embedded in the gas disk \citep{lin79,goldreich80,crida06}.  
Simulations by \citet{pinilla12a} produced a ring of large grains in a broad 
surface density bump exterior to the radius of such a gap.  Furthermore, their 
models of the mutual evolution of the gas and dust predicted a strong radial 
variation in the dust-to-gas mass ratio (see their Figure 6) with heavily 
depleted dust densities in the outer disk ($\zeta \sim 10^{-3}$ -- $10^{-4}$) 
and a large concentration within the trap ($\zeta \sim 1$).  Both the 
ring-shaped surface density profile and radially decreasing dust-to-gas ratio 
produced by their calculation were necessary features of the model we derived 
for the V4046 Sgr disk.

Recent models that account for the expected substructure in these ring-like 
pressure traps suggest that a modest variation in the gas density 
(peak-to-valley amplitude ratio of $\gtrsim$1.5) can produce strong azimuthal 
asymmetries in the dust distribution \citep{regaly12,birnstiel13}.  This 
phenomenon may be sufficient to qualitatively explain some observations of the 
``lopsided'' emission from some transition disks \citep[e.g.][]{brown09,
mayama12,casassus13,vandermarel13}.  As noted in \S\ref{sec:results}, the western peak of the 
V4046 Sgr dust ring appears slightly brighter than the eastern peak by 
$\sim 5$\,mJy\,beam$^{-1}$.  We find that this subtle asymmetry is apparent in 
the continuum visibilities as well, and therefore is not likely to be an 
artifact of the imaging/deconvolution process.  Figure \ref{fig:asym} shows a 
cut of the imaginary component of the complex visibilities binned along the 
$v$-axis in the Fourier plane (corresponding to the E-W direction).  For a 
symmetric morphology, the imaginary visibilities should be zero: however, we 
find a statistically significant deviation in a sinusoid pattern.  The 
asymmetry can be reproduced reasonably well with a faint point source 
($\sim$8\,mJy) located $\sim$0\farcs23 to the west of the disk center.  However,
verification and characterization of this low-level apparent asymmetry beyond 
such a simple model will require observations with higher resolution and 
sensitivity.

\begin{figure}[t!]
\epsscale{0.6}
\plotone{V4046Sgr2013/f08.eps}
\caption{ The complex visibilities of the 1.3\,mm emission binned along the 
$v$ axis ({\it black points}) with a model 8\,mJy point source ({\it red}) 
offset by $\Delta \alpha = -0\farcs23$ from the disk center ({\it red curve}).  
The inset shows the continuum image with 5\,$\sigma$ contour levels and the 
model point source ({\it red dot}).
\label{fig:asym}}
\end{figure}

The ring-shaped dust disk and tentative asymmetry seem to be well-matched to 
predictions for a large pressure trap generated by a companion interacting with 
the gas disk.  However, the central spectroscopic binary itself is too compact 
and circular to have dynamically truncated the disk at such a large radius: this
scenario requires (at least) a third body.  One constraint on the properties of 
a putative companion comes from the dynamical mass estimate by 
\citet{rosenfeld12}.  Since this method utilizes the Keplerian rotation pattern
of the disk, it is sensitive to the cumulative mass at the disk center: the 
spectroscopic binary plus any other companion(s).  If the central binary and 
disk are co-planar, \citet{rosenfeld12} showed that the disk-based dynamical 
mass estimate was consistent with that inferred from spectroscopic monitoring of
the stellar orbits \citep{stempels13}.  Therefore, any additional companion must
have a low mass: the 3\,$\sigma$ uncertainties on the dynamical mass estimate of
\citet{rosenfeld12} require that $M_{\rm{comp}} \le 0.3$\,M$_\odot$.  A second,
complementary constraint can be made from the contribution of any companion to 
the near-infrared spectrum.  The composite stellar photosphere model adopted 
here accounts for 99, 95, and 85\%\ of the observed emission in the $K$, 
$L^{\prime}$ and $M$ bands, respectively \citep{jensen97,hutchinson90,
skrutskie06}.  If we attribute the remaining emission to a co-eval low-mass 
companion, rather than the tenuous inner dust ring described in \S 4, pre-main 
sequence models \citep[e.g.][]{baraffe98} suggest that the spectrum requires
$M_{\rm comp} \le 0.2$\,M$_\odot$.

\begin{figure}[t!]
\epsscale{0.6}
\plotone{V4046Sgr2013/f09.eps}
\caption{Limits on a companion mass with regions ruled out by the dynamical 
mass estimate \citep[{\it orange dots};][]{rosenfeld12}, infrared luminosities 
\citep[{\it solid blue};][]{jensen97,hutchinson90,skrutskie06}, and infrared 
aperture-masking observations ({\it thick rose lines}; A.~L.~Kraus, private 
communication).  The lower mass limit consistent with the reported velocity 
shifts is also shown \citep[{\it gray line};][]{quast00,donati11,rodriguez10}. 
The position and 2\,$\sigma$ widths of the main and inner dust rings are marked 
by floating error bars.
\label{fig:companion}}
\end{figure}

A third constraint on a potential companion was suggested by \citet{donati11}, 
based on their claim of a shift in the systemic velocity of the spectroscopic 
binary.  They measured $V_{\rm sys}=-5.7\pm0.2$\,km s$^{-1}$ (in the 
heliocentric frame) at a mean epoch of 2009.7, which is significantly different
than the previous measurements of the binary \citep[$-6.94\pm0.01$\,km 
s$^{-1}$ at epoch 1985.5;][]{quast00} and the gas disk 
\citep[$-6.26\pm.05$\,km s$^{-1}$;][]{rodriguez10}.  Assuming the total mass 
determined by \citet{rosenfeld12} and co-planar orbits, we can derive a lower 
limit on the companion mass as a function of semi-major axis (or orbital period)
from these apparent velocity shifts: $M_{\rm comp} \gtrsim 0.05$\,M$_\odot 
(a/{\rm 1AU})^{1/2}$, where $a$ is the semi-major axis.  It should be 
noted that these apparent velocity shifts are likely consistent with one another
when considering the typical systematic uncertainties in the velocity
calibration between different instruments.  A final, more direct, constraint on
the companion mass can be made from infrared aperture-masking observations 
(A.~L.~Kraus; private communication). No companion is detected in those data, 
with a $K$-band contrast limit of 6-7 magnitudes outside of 0\farcs04 from the 
central binary.  Assuming a co-eval companion and the \citet{baraffe98} models,
these data constrain $M_{\rm comp} \le 0.07$\,M$_\odot$ for $a>2.9$\,AU.  
Figure \ref{fig:companion} summarizes these four constraints on $M_{\rm comp}$ 
and $a$.

Given the results in Figure \ref{fig:companion}, it is clear that the mass of 
any companion must be low, making a low-mass brown dwarf or massive giant planet
likely candidates.  The position of this hypothetical planet from the observed 
ring is highly uncertain and with no conclusive evidence for reduced densities 
in the cavity, our CO observations cannot help constrain $M_{\rm comp}$ or $a$. 
However, sensitive observations with high angular resolution may resolve 
structure in the inner gas disk \citep{casassus13} or signatures of clearing in 
the high-velocity wings \citep{dutrey08}.  An alternative way of
locating a companion is by looking at the small grains that filter past it 
and using this flow of small grains entrained in the gas to quantify the mass of
the companion \citep{rice06,zhu12}.  While the distribution of $\mu$m-sized dust
particles in our model are closer to the central binary than the inner working
angle of the SEEDS survey \citep[0\farcs1-0\farcs15]{tamura09},
the V4046 Sgr disk is a natural candidate to test this scenario with scattered 
light observations \citep[e.g.][]{dong12}.

A compelling potential alternative mechanism for shaping the distinctive 
structure of the V4046 Sgr disk is photoevaporation \citep{hollenbach94,
clarke01,alexander06a,alexander06b,owen12} which can also create large pressure 
traps \citep{alexander07}.  Strong photoionizing radiation from the central 
stars drive a wind off the disk surface, and consequently can open a gap in the 
disk on $\sim$AU scales.  V4046 Sgr has a large UV excess \citep{delareza86,
hutchinson90,huenemoerder07} and large X-ray luminosity 
\citep[$L_X \sim (7-10)\times 10^{29}$\,erg s$^{-1}$, based on archival XMM 
data; see also][]{gunther06,argiroffi12}.  According to recent models
\citep{gorti09a,gorti09b,owen10,owen11}, this high-energy emission should drive 
a photoevaporative wind.  Indeed, the presence of such a wind has been 
inferred via analysis of mid-infrared [Ne II] emission from V4046 Sgr 
\citep{sacco12}.  However, given the suite of models presented by 
\citet{owen11} and the present accretion rate of V4046 Sgr \cite[$\log\dot{M}
= -9.3\pm0.3$\,M$_\odot$ yr$^{-1}$;][]{donati11,curran11}, the X-ray luminosity 
is too low to explain the large cavity size inferred here.  However, X-ray 
photoevaporation (and perhaps even EUV/FUV photoevaporation) might find a 
natural role in explaining the depletion of material inside the innermost dust 
ring ($r \lesssim 3$\,AU inferred from the SED, see \S\ref{sec:model1}).

Perhaps the greatest mystery surrounding the V4046 Sgr disk is its existence: 
most theoretical and observational work suggests that disk dispersal processes 
should substantially diminish the observational signatures studied here well 
within the lifetime of this system and the few others like it \citep{clarke01,
alexander06b,takeuchi05,currie07,currie09}.  The modeling analysis 
conducted here unfortunately is unable to directly answer the question of why 
this gas-rich disk persists even at an age of 10-20\,Myr.  However, the inferred
disk structure might be providing some clues.  It might be possible that the 
cavity was formed early in the disk lifetime, stifling the viscous evolution 
process by severely limiting the gas accretion flow onto the central stars, and
thereby slowing the natural evolution of the disk material.  Alternatively, 
\citet{alexander12} suggested that the tidal torque induced by a close binary 
on its disk may inhibit accretion and increase the lifetime of the circumbinary 
disk compared to that of a disk around a single star. Such scenarios are 
speculative: but it is interesting to note that most (if not all) of the known 
long-lived, gas-rich disks have evidence for low-density cavities at their 
centers.

Overall, the signatures of evolution in the transition disk hosted by V4046 Sgr
are remarkable, suggesting that this target merits a focused effort moving 
forward with ALMA.  Its close proximity and large physical size enables high 
resolution and high sensitivity observations.  Furthermore, the stellar 
binarity aids the observations: it is too tight to affect the disk structure, 
but the high stellar mass means that the widths of gas disk emission lines is 
large and so spectral line observations can have good velocity resolution 
without sacrificing sensitivity.  Observations are further aided by the fact 
that it is an isolated system unaffected by cloud contamination.  With all of 
this in mind, V4046 Sgr promises to be an excellent laboratory for studying disk
evolution and planet formation: its old age means that evolutionary features are
well-advanced, and its rich gas disk provides an excellent target for both 
detection efforts and spatially resolved analysis.  Lastly, the X-ray luminosity
makes V4046 Sgr a natural target for studying photoevaporation and its effects 
on disk evolution.

\section{Summary}

We have studied the structure of the V4046 Sgr circumbinary disk using high 
resolution, high sensitivity mm-wave observations from the SMA.  By modeling the
SED, spatially resolved 1.3\,mm dust continuum, and line emission of three CO 
isotopes, we have identified this disk as an exemplar evolved protoplanetary 
disk.  The key conclusions of our analysis are:

\begin{enumerate}
\item The CO gas disk is significantly more extended than the compact 
distribution of large grains.  In order to reproduce this feature we decouple 
the dust and gas surface density profiles, reducing the dust-to-gas ratio in the
outskirts of our disk model by a factor of 10.  We interpret this phenomenon as 
the result of the simultaneous growth and inward radial drift of mm/cm-sized 
particles in a viscously evolving gas disk \citep{birnstiel12a}.

\item The V4046 Sgr disk features a large central cavity ($r=29$\,AU) depleted 
of mm/cm-sized dust grains.  These grains are concentrated in a narrow ring 
($\mu=37$\,AU, FWHM=$16$\,AU).  There is no conclusive evidence from the CO 
line emission whether there is an analogous depletion in the gas disk. This 
distinctive morphology is consistent with simulations that feature large 
pressure traps generated by the dynamical interaction of a massive companion 
with a gas disk and the evolution of solid material \citep{pinilla12a}.

\item The cavity contains a significant population of $\mu$m-sized grains.  
We infer that this distribution of small dust particles has a smaller central
clearing that may be consistent with X-ray photoevaporation \citep{owen11}, but 
is too large to be truncated by the central binary \citep{artymowicz94}.  This 
population might originate from small grains entrained in the gas flowing 
past the pressure barrier that traps the mm/cm-sized grains in the large ring
\citep{rice06}.

\end{enumerate}

\section*{Chapter Acknowledgments} We are grateful to Adam Kraus for sharing his $K$-band limit on
a companion, Andrea Isella for teaching K.R. about asymmetries, and Moritz 
G{\"u}nther for helping us with the X-ray luminosities.  We are also indebted to
Kees Dullemond, Guillermo Torres, and Matt Payne for insightful discussions.  
We acknowledge support from NASA Origins of Solar Systems grant No. NNX11AK63.  
J.K.'s research on V4046 Sgr and other nearby protoplanetary disks is supported 
in part by National Science Foundation grant AST-1108950 to RIT.  The 
Submillimeter Array is a joint project between the Smithsonian Astrophysical 
Observatory and the Academia Sinica Institute of Astronomy and Astrophysics and 
is funded by the Smithsonian Institution and the Academia Sinica. This work is 
based [in part] on archival data obtained with the Spitzer Space Telescope, 
which is operated by the Jet Propulsion Laboratory, California Institute of 
Technology under a contract with NASA. Support for this work was provided by 
NASA.

\clearpage

\begin{table}
\begin{center}
\centering
\caption{Summary of Observations\label{tab:obs}}\medskip
\centering
\begin{threeparttable}
 \renewcommand\TPTminimum{\linewidth} %%% we want full width
\centering
\begin{tabular}{lcccccc}
  \toprule
UT date & configuration & antennas & baselines & $t_{\rm on}$ & $\Delta \nu_{\rm cont}$ & $\tau_{\rm atm}$ \\
 & &  & [k$\lambda$] & [min] & [GHz] & (225\,GHz) \\
  \midrule
2009 Feb 23 & extended      & 8 & 17-132 & 110 & 3.4 & 0.05 \\
2009 Apr 25 & compact-N     & 6 & 6-84   & 220 & 3.4 & 0.06 \\
2011 Mar  8 & sub-compact   & 7 & 4-50   & 90  & 7.4 & 0.05 \\
2011 Sep  4 & very extended & 8 & 19-390 & 180 & 7.4 & 0.04 \\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

### 2015-07-03 hd163296/ms_hd163296_chapter.tex
initial commit

\chapter{A Spatially Resolved Vertical Temperature Gradient in the HD 163296 Disk}\label{c:hd163296}

\section*{abstract}
We analyze sensitive, sub-arcsecond resolution ALMA Science Verification 
observations of CO emission lines in the protoplanetary disk hosted by the 
young, isolated Ae star HD 163296.  The observed spatial morphology of the 
$^{12}$CO $J$=3$-$2 emission line is asymmetric across the major axis of the 
disk; the $^{12}$CO $J$=2$-$1 line features a much less pronounced, but similar,
asymmetry.  The $J$=2$-$1 emission from $^{12}$CO and its main isotopologues 
have no resolved spatial asymmetry.  We associate this behavior as the direct 
signature of a vertical temperature gradient and layered molecular structure in 
the disk.  This is demonstrated using both toy models and more sophisticated 
calculations assuming non-local thermodynamic equilibrium (non LTE) conditions. 
A model disk structure is developed to reproduce both the distinctive spatial 
morphology of the $^{12}$CO $J$=3$-$2 line as well as the $J$=2$-$1 emission 
from the CO isotopologues assuming relative abundances consistent with the 
interstellar medium.  This model disk structure has $\tau=1$ emitting surfaces 
for the $^{12}$CO emission lines that make an angle of $\sim 15^\circ$ with 
respect to the disk midplane.  Furthermore, we show that the spatial and 
spectral 
sensitivity of these data can distinguish between models that have 
sub-Keplerian gas velocities due to the vertical extent of the disk and its 
associated radial pressure gradient (a fractional difference in the bulk gas 
velocity field of $\gtrsim 5$\%).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Spectral line emission from molecules can be a powerful tool for studying the 
structure of protoplanetary disks.  Abundant molecules in disks, such as CO or 
CN, trace the spatial distribution of gas temperatures and the radial extent of 
the molecular hydrogen that composes most of the disk mass 
\citep{koerner93,guilloteau13}.  Analysis of multiple lines can determine the 
vertical temperature structure of the disk \citep{dartois03} or indicate 
freezeout of the gas phase molecules onto dust grains in the cold midplane 
\citep{qi11}.  Observations of various species are useful for studying the 
abundance distributions of molecules and processes like ionization 
\citep{oberg11}, grain surface reactions \citep{dutrey11}, and fractionation 
\citep{oberg12}.  Furthermore, these spectral lines probe the bulk motions, or 
kinematics, of the gas \citep{beckwith93}.  Observations from interferometers 
are particularly useful since the emission is resolved both spatially and 
spectrally and can be used to derive a dynamical mass of the young star 
\citep{guilloteau98,simon00}, probe the structure of the disk 
\citep{dutrey08,panic10}, or detect non-thermal line broadening from turbulence 
\citep{hughes11,guilloteau12}.

The Herbig Ae star HD 163296 (spectral type A1) hosts a protoplanetary disk with
bright continuum emission \citep{allen76,mannings97,natta04} and a rich
molecular spectrum \citep{qi01,thi01,thi04,qi13}.  It is an isolated system, not
known to be associated with any star forming region or young moving cluster 
\citep{finkenzeller84}.  Observations of CO lines show that the disk exhibits
Keplerian rotation \citep{mannings97,isella07,hughes08} and is remarkably large,
extending past a radius of $\sim 500$\,AU \citep{isella07}.  The disk is seen in
scattered light \citep{grady00} and is associated with an asymmetric Herbig-Haro
outflow \citep{devine00} and molecular disk wind \citep{klaassen13}.  The system
is young \citep[$\sim 5$\,Myr;][]{natta04} and its {\it Hipparcos} parallax 
indicates it is nearby \citep[$d= 122^{+17}_{-13}$\,pc;][]{vandenancker98}. 
\citet{qi11} developed a density and temperature structure for this disk that 
was consistent with Submillimeter Array (SMA) observations of both the dust 
continuum and multiple CO emission lines.  Including an abundance distribution 
that accounts for both freezeout and photodissocation, their analysis suggested 
that the disk has a midplane CO snow line at $r\sim155$\,AU.  An ancillary 
result of their model is that the disk is colder in the midplane than in the 
atmosphere where the higher $J$ emission lines are produced.  Evidence for a 
vertical temperature gradient in this disk was independently presented by 
\citet{akiyama11}.

We present an analysis of the spectral line emission from the main CO 
isotopologues observed by the Atacama Large Millimeter/sub-millimeter Array 
(ALMA) towards the HD 163296 disk in Bands 6 and 7.  We first use a toy model 
comprised of a rotating double cone to guide our intuition of this data. 
Leveraging the high spatial and spectral resolution of these observations, we 
then demonstrate that the vertical temperature gradient in the disk is directly 
resolved in the $^{12}$CO $J$=3$-$2 maps alone.  We develop a disk structure 
model that is roughly self-consistent with those data and the $J$=2$-$1 
transitions of $^{12}$CO, $^{13}$CO, and C$^{18}$O.  In addition, we explore the
velocity structure of the disk and show that the excellent spatial and spectral 
resolution of the sensitive Band 7 observations can distinguish between various 
classes of disk models which have bulk gas velocities that deviate at the 
$\gtrsim 5$\% level.  We describe the observations in \S\ref{sec:observations} 
and show the results in \S\ref{sec:results}.  Our analysis follows in 
\S\ref{sec:analysis}, where we present the vertical temperature gradient 
(\S\ref{sec:models}) and explore sub-Keplerian motions due to the vertical 
geometry of the disk and its radial pressure gradient (\S\ref{sec:velocity}).  
We conclude with a brief discussion in \S\ref{sec:discussion} and a summary in 
\S\ref{sec:summary}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Observations and Data Reduction}\label{sec:observations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

HD 163296 ($\alpha = 17^{\rm h} 56^{\rm m} 21\fs287$, 
$\delta = -21\degr57\arcmin22\farcs39$, J2000) was observed by ALMA in Band 6 
and Band 7 as part of its commissioning and science verification (SV) program. 
The raw and calibrated visibility data were publicly released through the 
science portal, and were accompanied by a set of detailed calibration and 
imaging scripts provided by the ALMA SV team\footnote{\url{https://almascience.nrao.edu/alma-data/science-verification/overview}}.  
We started with the calibrated measurement set and used the CASA software 
package \citep[v3.4;][]{mcmullin07} to produce self-calibrated and continuum 
subtracted spectral visibilities for the $^{12}$CO, $^{13}$CO, C$^{18}$O 
$J$=2$-$1 lines at 230.538, 220.399, 219.560\,GHz in Band 6 and $^{12}$CO 
$J$=3$-$2 at 345.796\,GHz in Band 7.

The Band 6 observations were taken using 24 of the ALMA 12\,m antennas on 2012 
June 9, June 23, and July 7 for a total on-source time of 84 minutes (including 
latency).  The baselines spanned a range of 20 to 400\,m.  The correlator was 
configured to simultaneously observe four spectral windows (SpW), with two in 
each sideband:  SpWID \#1 included the $^{13}$CO and C$^{18}$O $J$=2$-$1
transitions, while SpWID\#2 covered the  $^{12}$CO $J$=2$-$1 line.  SpWIDs \#0 
and \#3 observed line-free continuum \citep[but see][]{qi13} and were centered 
at 217 and 234\,GHz.  The flux calibrator for each of the nights was Juno, 
Neptune, and Mars, respectively, and the bandpass solution was generated from 
observations of the quasar J1924-292.  Integrations on the science target were 
interleaved with $\sim 2$ minute long observations of the nearby quasar 
J1733-130 for phase calibration.  For the July 7 observations, the flux density 
of J1733-130 was determined to be 1.55\,Jy, using a bootstrap from the flux 
calibrator, Mars.  Similarly, HD 163296 was observed in Band 7 on 2012 June 9, 
Jun 11, June 22, and July 6 with the same antenna configuration.  Juno (June 9 
only) and Neptune were observed as primary flux calibrators.  When Neptune was 
observed, the flux density of J1733-130 was bootstrapped for each individual 
SpW.  An additional bootstrap from these observations was used to determine the 
flux density of J1733-130 in each SpW observed on June 9th.  The total 
on-source time was 140 minutes.  These observations used the same bandpass and 
phase calibration strategy as the Band 6 observations.  The correlator was set 
up to observe four spectral windows, with two in each sideband: SpWIDs \#0 and 
\#3 were centered at 360 and 347\,GHz for continuum, HCO$^+$ $J$=4$-$3 at 
356.734\,GHz was detected in SpWID \#1, and the $^{12}$CO $J$=3$-$2 line was 
covered by SpWID \#2.\footnote{The transformation to the absolute sky frequency 
in the Band 7 measurement set provided by the SV team is incorrect.  The 
$^{12}$CO $J$=3$-$2 line suggests that the systemic velocity, $v_{\rm sys}$, of 
HD 163296 as measured in the local standard of rest frame (radio definition; 
LSRK) is $6.99$\,km s$^{-1}$.  This disagrees with other measurements 
\citep[$v_{\rm sys}=5.8$\,km s$^{-1}$;][]{mannings97,isella07,hughes08} as well 
as the ALMA Band 6 observations.  This velocity offset was confirmed by the ALMA
Helpdesk, but is only a minor inconvenience since the line morphology is, for 
this small offset, sensitive only to the velocity differences.}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f01.pdf}
\caption{The disk hosted by HD 163296 imaged in four CO lines. The 
disk orientation is indicated by the solid gray lines, the synthesized beam 
dimensions are shown in the lower left corner panels.  Each line has been 
regridded onto the same velocity resolution (written relative to the systemic 
velocity in km s$^{-1}$).  The emission is shown in both color and with 
$\sigma \times [3,6,10 + 5n]\,(n = 0,1,2,\ldots)$ contours (see Table 
\ref{tab:lines} for noise estimates).  The last panel for each line contains the
0th (contours) and 1st (color scale) integrated moment maps.
\label{fig:datachmaps}}
\end{figure}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f02.pdf}
\caption{The $^{12}$CO $J$=3$-$2 line imaged with a channel width of
$\Delta v = 0.11$\,km s$^{-1}$.  
\label{fig:CO32chmaps}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure \ref{fig:datachmaps} shows channel maps of the $^{12}$CO $J$=3$-$2 line 
along with the $J$=2$-$1 lines from $^{12}$CO, $^{13}$CO, and C$^{18}$O.  
Before imaging, the visibilities have been regridded onto the same 
0.34\,km s$^{-1}$ wide channels; only the central 15 channels are shown, where 
the emission pattern is clearly resolved for all lines.  The central 49 channels
of the $^{12}$CO $J$=3$-$2 line imaged at the native 
$\Delta v=0.11$\,km s$^{-1}$ channel spacing is also shown in Figure 
\ref{fig:CO32chmaps}.  We generated these synthesized images using the CASA 
implementation of the CLEAN algorithm \citep{hogbom74} with natural weighting.  
Table \ref{tab:lines} summarizes the noise characteristics, synthesized beams, 
native channel widths and integrated intensities for each line.  The reported 
line intensities agree with previous measurements by the SMA \citep{qi11} 
within the 10\% systematic uncertainty estimated for the absolute calibration of
the visibility amplitudes.

In order to produce spectral images with such high fidelity, we developed a 
CLEAN mask based on the emission pattern expected from a disk in Keplerian 
rotation \citep{beckwith93}, which requires some assumption about the stellar 
mass, disk inclination, disk size, phase offset, position angle, and systemic 
velocity (see \S\ref{sec:toy}).  Except for the stellar mass, all of these
parameters can be estimated from an initial imaging of the data with a square 
mask; the final imaging results are consistent for any reasonable choices of 
these parameters.  We identified the region of the rotating disk where the 
projected line of sight velocities corresponded to each individual spectral 
channel.  This region was then convolved with a fixed Gaussian kernel 
($\sigma = 0\farcs7$) to broaden the mask and ensure that all of the significant
emission was covered and that the CLEANed area was always much larger than the 
synthesized beam size.  This recipe produces a separate mask for each channel, 
within which the measurement set was CLEANed deeply.  Some additional CLEAN 
iterations were then applied with no mask.  This masking technique significantly
improves the quality of the images compared to the reference maps provided in 
the SV data package (the signal-to-noise ratio per channel improves by a factor 
of $\sim$2--3).

The $^{12}$CO $J$=3$-$2 emission observed toward the disk around HD 163296 is 
asymmetric in its spatial morphology across the major axis of the disk 
(PA$=312^\circ$).  For channels offset from the line center by velocities 
$\gtrsim 1$\,km s$^{-1}$, the resolved emission appears systematically closer 
to the southern semi-minor axis (see Figure \ref{fig:CO32chmaps} and the top set
of channel maps in Figure \ref{fig:datachmaps}).  Furthermore, 
channel-by-channel, the southern half of the disk is brighter than the northern.
The $^{12}$CO $J$=2$-$1 line has these same features (refer to Figure 
\ref{fig:datachmaps}), but the effect is much less obvious; the morphological 
asymmetry can only be clearly seen for a few channels, 
$|\Delta v| \approx 1.7$--2.0\,km s$^{-1}$.  The $^{13}$CO and C$^{18}$O 
spectral emission have no apparent spatial asymmetries (see Figure 
\ref{fig:datachmaps}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis}\label{sec:analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Before attempting to explain the detailed morphologies of these CO data, we 
provide a brief primer on molecular line emission from a rotating disk
\citep[e.g.][]{omodaka92,beckwith93}.  The observed line is centered around some
transition frequency, $f_0$, with a linewidth that is sensitive to the 
quadrature sum of both the thermal velocity and turbulent motions of the gas.  
If the emitting gas has some bulk velocity along the line of sight, 
$v_{\rm los}$, then the line center will be Doppler shifted away from the 
transition frequency, $f_0$, by an amount $\Delta f = -v_{\rm los}(f_0/c)$.  
Observations taken by (sub-)millimeter interferometers like ALMA have both 
spectral and spatial resolution, and produce an image of the source for a set of
frequency, or equivalently velocity, channels.  These spectral line observations
comprise a three dimensional data cube of position, position, and velocity.  For
an inclined disk in Keplerian rotation, the emission observed in a single 
channel ``highlights'' what part of the disk has that same projected velocity as
it is Doppler shifted into the channel center.

The morphology of the line emission in any given channel changes depending on 
both the physical conditions and bulk motions of the gas.  Modeling the line 
emission requires a calculation of the molecular excitation state of the gas and
an integration of the radiative transfer equation along each line of sight, $s$,
through the disk structure,
\begin{equation}
I_\nu = \int_0^\infty S_\nu(s) \exp\left[-\tau_\nu(s)\right] K_\nu(s) ds,
\label{eq:radiativetransfer}
\end{equation}
where $\tau_\nu(s) = \int_0^s K_\nu(s')ds'$ is the optical depth, $K_\nu$ is the
absorption coefficient, and $S_\nu$ is the source function.  All three of these 
terms depend upon the local temperature, $T_{\rm gas}$, and density, 
$\rho_{\rm gas}$, of the disk.  Determining the structure of a disk using 
optically thick CO lines is difficult, and so previous studies have leveraged 
observations of multiple lines to break the model degeneracies 
\citep{dartois03,pietu07}.

To parse what these data reveal about the disk hosted by HD 163296,
we will first present a toy model that qualitatively explains the observed 
morphological asymmetries noted in the $^{12}$CO transitions (\S\ref{sec:toy}).
Next, we describe a more complex modeling procedure for testing whether a given
disk structure produces the observed line intensity in addition to the 
morphological and brightness asymmetries (\S\ref{sec:models}).  We then present 
a simple model developed for pedaogogical purposes (\S\ref{sec:pedagogic}) 
followed by a more realistic hydrostatic disk model (\S\ref{sec:hydro}).  
Finally, we demonstrate that some care must be taken when defining the rotation 
pattern of the disk, in light of the disk structure we derive with its large 
radial extent and molecular layer extending relatively high above the midplane 
(\S\ref{sec:velocity}).

%**************************
\subsection{A Toy Model}\label{sec:toy}
%**************************

The asymmetric shape of the $^{12}$CO $J$=3$-$2 line emission cannot be 
explained by the emission pattern predicted for a vertically thin rotating disk.
The coordinates of an inclined thin disk ($x,y,z=0$), are related to the 
sky-plane relative to the observer, ($x',y'$), by
\begin{equation}
\left(\begin{array}{c} x \\ y \\ z \end{array}\right) = \left(\begin{array}{c} x' \\ y' / \cos i \\ 0 \end{array}\right)
\end{equation}
where $i$ is the disk inclination (0$^\circ$ is seen face-on).  For a disk 
rotating in circular, Keplerian motion, the projected line of sight velocity as a
function of position is
\begin{equation}
v_{\rm obs}(x',y') = \sqrt{\frac{GM_\ast}{r}} \sin i \cos \theta,
\end{equation}
with an argument $\theta = \arctan(y/x)$ and modulus $r=\sqrt{x^2+y^2}$.  This 
function is mirror symmetric across the $x'$-axis and so the shape of the 
emission observed at some velocity, $v_{\rm obs}$, should appear symmetric about
the $x'$-axis of the sky-plane.  Of course, this geometrical argument is grossly
simplified and considers neither the physical structure of the disk, nor the 
associated radiative transfer effects for spectral line emission (see 
\S\ref{sec:models}).  However, protoplanetary disks are thought to be 
geometrically thin \citep{pringle81}, with radial surface densities and 
temperatures that roughly follow power-laws \citep[i.e. slowly changing $\rho$ 
and $T$ in the outer disk;][]{weidenschilling77a,adams88}.  Under these 
conditions, this approximation works well and has been both a useful and 
effective guide in the study of line emission from protoplanetary disks.  
However, the observation analyzed here reveals an emission pattern that requires
a modified conceptual framework \citep[see also][]{pavlyuchenkov07,semenov08,
guilloteau12}.

The toy model that we propose is a differentially rotating double cone, oriented
along the $z$ axis with an opening angle $\psi \in (0,\pi/2)$ measured from the 
$xy$ plane (i.e., the disk midplane). In this case, rays originating from the 
the sky-plane will intersect the surface at
\begin{equation}
\left(\begin{array}{c} x \\ y \\ z \end{array} \right) = \left(\begin{array}{c} x' \\\ y' / \cos i + t \sin i \\ t\cos i\end{array}\right),
\end{equation}
where $t$ solves the quadratic equation\footnote{\url{http://www.geometrictools.com/Documentation/IntersectionLineCone.pdf}}
\begin{equation}
0 = t^2 \left[\cos(2i) + \cos(2\psi)\right] - 2 \sin^2(\psi)\left[{x'}^2 + {y'}^2 \sec^2(i) + 2 t y' \tan(i)\right].
\end{equation}
The positive and negative roots of this equation correspond to the near and 
far halves of the double cone respectively (where we define the near half as 
the one closer to the observer). Any given line of sight will intersect the near
cone at {\it larger} $y$ than for the flat disk, by an additional factor 
$t\sin i$ (assuming $i < 180^\circ$).  Therefore, if this near cone is in 
circular, Keplerian rotation, ${v_\theta}^2 =GM/\sqrt{x^2 + y^2}$ with no $v_z$ 
component, the observed isovelocity contours will be systematically shifted to 
{\it smaller} $y$.   The same argument follows for the far cone, except that 
isovelocity contours are systematically shifted to {\it larger} $y$.  If 
$i>180^\circ$ (which is the case for the HD 163296 disk; see 
\S\ref{sec:models}), then the effects are the same except the solutions switch 
to describe the opposite half of the double cone.  Larger values of $\psi$ will 
amplify the differences between these curves.  The emission morphology predicted
from this double cone structure clearly mimics the observed asymmetry in the 
$^{12}$CO $J$=3$-$2 line, as demonstrated in Figure \ref{fig:cone}.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f03.pdf}
\caption{The middle two panels separately show the observed velocities ({\it 
in color}) from the near (second panel) and far (third panel) halves of a double
cone structure in circular, Keplerian rotation with 
$\psi = 15^\circ$, $i=44^\circ$, 
$M_\ast = 2.3$\,M$_\odot$, and semi-major axis aligned with the $x'$-axis of the
observer (see text for parameter definitions).  Dashed contours ({\it in white})
show the $\pm$0.75, 1.5, and 2.25\,km s$^{-1}$ isovelocity contours.  The same 
quantities are shown for a flat disk ($\psi=0$; first panel).  The 
$\pm 1.5$\,km s$^{-1}$ contours for all three models are shown together in the 
fourth panel ({\it solid curve} is the flat disk, {\it dashed curve} is the near
cone, {\it dotted curve} is the far cone).
\label{fig:cone}}
\end{figure}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f04.pdf}
\caption{The solid curves show the isovelocity contours of the toy model,
corresponding to the channel velocities of the $^{12}$CO $J$=3$-$2 line 
({\it shown in gray}).  Both the near ({\it in black}) and far ({\it in red})
halves of the double cone structure are shown out to a radius of 600\,AU.  The 
model parameters are the same as in Figure \ref{fig:cone}.
\label{fig:cone_chmaps}}
\end{figure}

The brightness and morphological asymmetries of the $^{12}$CO $J$=3$-$2 and 
$^{12}$CO $J$=2$-$1 lines can be understood if the $\tau=1$ emitting surface is 
something like a double cone structure with $\psi \sim 15^\circ$.  The bright 
portion of the line corresponds to the near cone that is systematically shifted 
along the minor axis (to smaller $y$).  The dim ``shadows'', which are 
asymmetric in the opposite sense, represent the far cone.  Figure 
\ref{fig:cone_chmaps} shows the velocity pattern of this toy model overlaid on 
the regridded channel maps of the $^{12}$CO $J$=3$-$2 emission.   The key 
feature of the double-cone is that it mimics an emission structure that has 
little contribution from the disk midplane: the structure is in some 
observational sense ``hollow'' at its center.  Furthermore, the angle that the
cone makes with the disk midplane ($\psi \sim 15^\circ$) is fairly large 
\citep[but not abnormally so;][]{vanzadelhoff01} and suggests that the disk 
cannot be treated as geometrically thin.

Detailed radiative transfer and chemical models 
\citep[][]{aikawa99,aikawa06,semenov11,walsh12} predict a cone-like structure 
for molecules like CO.  In these calculations, the midplane temperatures are 
coupled to the cold dust, greatly reducing the CO gas-phase abundance via 
freezeout onto dust grains.  Higher in the disk, direct stellar radiation heats 
the material, liberating the CO from the grain surfaces and ensuring that 
CO is both abundant and warm.  Above this molecular layer, where gas column 
densities are low, the CO is depleted by photodissociation via energetic photons
from the star and background radiation field.  It is this first layer, the cold 
midplane, that naturally explains the ALMA observations.  In this region, the 
gas is too cold to densely populate the higher $J$ CO transitions, so the 
observed emission is low.  Additionally, the CO number densities should be 
heavily depleted by freezeout onto dust grains, further reducing the emission.  
Instead, most of the emission is from the warm molecular layer which, being high
above the midplane, should produce asymmetric emission with respect to the major
axis of the disk.  This scenario also explains why the less abundant $^{13}$CO 
and C$^{18}$O emission is not asymmetric: these lines originate from deeper 
layers in the disk, which correspond to lower $\psi$ and less severe asymmetries
that are not resolved by the observations.  Indeed, calculations by 
\citet{pavlyuchenkov07} and \citet{semenov08} of HCO$^+$ $J$=4$-$3 emission for 
axisymmetric, inclined disks ($i\sim60^\circ$) predicted this same morphology.

%**************************
\subsection{Physical Model Analysis}\label{sec:models}
%**************************

While the toy model of a differentially rotating double-cone provides a useful
qualitative explanation for the asymmetric morphology of the $^{12}$CO $J$=3$-$2
line emission, we now undertake a more physically motivated, quantitative 
analysis.  We build upon the disk structures developed by \citet{dartois03} and 
\citet{qi11} to reproduce the morphology of the $^{12}$CO $J$=3$-$2 line along 
with the line intensities of the $J$=2$-$1 emission from the $^{12}$CO, 
$^{13}$CO, and C$^{18}$O isotopes.  In this section, we introduce the modeling 
scheme that we will use to illustrate the sensitivity of the observations to the
temperature structure and velocity field of the disk.

Following the formalism explained by \citet{andrews12}, axisymmetric density 
and temperature structures are defined using a polar cylindrical 
coordinate system ($r,z$).  The gas temperature structure, $T_{\rm gas}(r,z)$,
is specified parametrically throughout this work.  The gas surface density 
profile is assumed to be the self-similar model of a thin, viscous accretion 
disk 
\citep{lynden-bell74,hartmann98},
\begin{equation}
\Sigma_{\rm gas}(r) = \Sigma_c \left(\frac{r}{r_c}\right)^{-\gamma} \exp \left[-\left(\frac{r}{r_c}\right)^{2-\gamma}\right],
\end{equation}
where $r_c$ sets the radial size of the gas disk, $\gamma$ is a power-law index,
and $\Sigma_c = M_{\rm gas} (2-\gamma)/(2\pi r_c^2) $ is the normalization, 
where $M_{\rm gas}$ is the total gas mass.  The density structure, $\rho(r,z)$, 
is either parametically defined (\S\ref{sec:pedagogic}) or calculated from the 
temperature structure using the equation of hydrostatic equilibrium 
(\S\ref{sec:hydro}).  To calculate the CO number density, we assume that 80\% of
the gas (by number; $\mu = 2.37$) is composed of molecular hydrogen and that the
CO:H$_2$ relative abundance is a constant, $f_{\rm co}$.  Molecular depletion 
due to freezeout is implemented by decreasing $f_{\rm co}$ by a factor of 
10$^{8}$ in the region of the disk where gas temperatures fall below a freezeout
temperature, $T_{\rm gas} < T_{\rm frz}$.  A photodissociation boundary is 
calculated by vertically integrating the H nuclei density 
\citep[$0.706n_{\rm gas}$, the same procedure as in \citet{qi11};][]{aikawa99} and thresholding for 
heights, $z_{\rm phot}$, where the column density is less than a constant value,
\begin{equation}
\sigma_s > 0.706 \int_{z_{\rm phot}}^\infty n_{\rm gas}(r,z') dz'.
\end{equation}
This surface density is equivalent to the unitless $\Sigma_{21}$ defined by
\citet{aikawa06} and used by \citet{qi11} where 
$\Sigma_{21} = \sigma_s / 1.59 \times 10^{21}$\,cm$^{-2}$.

We assume a constant turbulent velocity width, $\xi=0.01$\,km s$^{-1}$, and disk
inclination, $i=224^\circ$.  The bulk gas velocities are described in each 
section and we fix several important model parameters: the stellar mass 
\citep[$M_\ast = 2.3$\,M$_\odot$;][]{natta04}, source distance 
\citep[$d=122$\,pc;][]{vandenancker98}, and major axis position angle 
(measured East of North, PA = 312$^\circ$).  Note that the asymmetries in the 
$^{12}$CO $J$=3$-$2 line emission effectively resolve the absolute disk spin 
orientation \citep{pietu07}, so our PA and inclinations differ by 
$\sim 180^\circ$ from the models presented by \citet{isella07} and 
\citet{qi11}.  See \S\ref{sec:discussion} for a more detailed discussion.

For any given disk structure model, we calculate the CO level populations using 
the non-LTE molecular excitation and radiative transfer package LIME 
\citep{brinch10} and molecular data from the LAMDA database 
\citep{schoier05,yang10}.  LIME is then used to generate channel maps matching 
the native channel spacing of the ALMA data (see Table \ref{tab:lines}).  We
then smooth using the Hanning algorithm in the spectral dimension with a three 
channel kernel to mimic the ALMA pipeline \citep[see 
\S\ref{sec:discussion};][]{lundgren12} and calculate model visibilities to match
the Fourier sampling of the measured visibilities.  We evaluate the model fit 
through a $\chi^2$ statistic of the complex visibilities and also by inspection 
of CLEANed channel maps using the same masks and imaging procedure applied to 
the data (see \S\ref{sec:observations}).

It is important to emphasize that the models developed here involve many 
degenerate parameters: the exact structures we present are neither unique nor 
necessarily optimal.  The analysis presented here is only meant to provide a 
basic physical understanding of some new, key aspects of these data.

%**************************
\subsubsection{A Pedagogical Structure}\label{sec:pedagogic}
%**************************

In this section, we will demonstrate how a vertical temperature gradient 
naturally produces the asymmetric emission morphology of the $^{12}$CO $J$=3$-$2
line.  In order to disambiguate the effects of density and temperature, we 
define a single parametric disk density structure and calculate the line 
emission for two simple cases for the temperature structure:
(a) the disk is vertically isothermal and (b) the disk has a 
warm ($T\sim30$\,K) molecular layer above a cold ($T\sim20$\,K) midplane.

\begin{figure}[t!]
\epsscale{0.95}
\plottwo{hd163296/f05.pdf}{hd163296/f06.pdf}
\caption{Temperature ({\it color scale}) and gas density structure 
({\it black contours}; $\log n_{\rm gas} = [4,5,6,7,8,9]$\,cm$^{-3}$) of the 
pedagogical models (\S\ref{sec:pedagogic}).  (a) A vertically isothermal model 
structure. (b) A model structure that features an artificial vertical 
temperature gradient.  The dashed red curves mark the photodissociation 
boundaries, for $\sigma_s = 5\times 10^{-20}$\,cm$^{-2}$.
\label{fig:pedagogical_struct}}
\end{figure}

We assume that the gas is vertically distributed with the Gaussian profile 
appropriate for a vertically isothermal disk in hydrostatic equilibrium 
(neglecting self gravity):
\begin{equation}
\rho_{\rm gas}(r,z) = \frac{\Sigma_{\rm gas}(r)}{\sqrt{2\pi}H} \exp \left[-\frac{z^2}{2H^2}\right].
\end{equation}
For this pedagogical example, the temperature and density structure are 
entirely decoupled. The disk scale height is defined to be 
$H(r) = 16(r/150$\,AU$)^{1.35}$\,AU.  The velocity field is defined such that 
the disk is in circular Keplerian rotation about the central star, with an azimuthal
component
\begin{equation}
{v_K}^2 = \frac{G M_\ast}{r}.
\end{equation}
Depletion due to CO freezeout is not considered in this calculation, but 
photodissociation is included with a threshold density 
$\sigma_s=5\times10^{20}$\,cm$^{-2}$.  The model disk has a gas mass of 
$0.09$\,M$_\odot$, $r_c = 115$\, AU, $\gamma = 0.8$, and 
$f_{\rm co} = 5\times10^{-5}$.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f07.pdf}
\caption{The $^{12}$CO $J=3-2$ line emission predicted by the models 
described in \S \ref{sec:pedagogic}.  The top set of channel maps shows the 
emission from a vertically isothermal model structure, while the bottom model 
structure has a vertical temperature gradient (see Figure 
\ref{fig:pedagogical_struct}). 
\label{fig:pedagogical}}
\end{figure}

We first calculate the $^{12}$CO $J$=3$-$2 line emission from this disk 
structure assuming that it is vertically isothermal with a radial temperature 
structure set by a power law, $T_{\rm gas} = 65$K\,$(r/100$\,AU$)^{-0.5}$.  The 
disk density and temperature structure is shown in Figure 
\ref{fig:pedagogical_struct}(a), and the corresponding channel maps in Figure 
\ref{fig:pedagogical} (top panels).  The qualitative discussion from 
\S\ref{sec:toy} is confirmed: the morphology of the line is essentially 
symmetric across the major axis with only a slight brightness asymmetry between 
the Northern (near) and Southern (far) sides of the disk (which is reversed if 
the inclination is flipped by $90^\circ$).  This geometric effect, for 
sensitive data, indicates the absolute orientation of the disk and can
qualitatively be understood by considering the emission that originates above 
the disk midplane.  The half of the disk (located in the N) that is tilted 
toward the observer will then have a smaller projection on the observer sky 
plane (and appear dimmer) than the disk half (located in the S) that is tilted 
away.  The double cone structure described in \S\ref{sec:toy} similarly
encapsulated this effect: in Figure \ref{fig:cone_chmaps} the isovelocity 
contours of the near cone (black curve) appear noticeably shorter in the N part
of the disk than in the S for channels near the line center, $|v_{\rm los}| 
\lesssim 1.02$\, km s$^{-1}$.  Optical depth effects and the disk
structure also contribute to the details of this brightness asymmetry.

Next, we build on that calculation by including a vertical gradient to the
temperature structure.  We start by defining a ``warm'' region where the gas is 
a constant $T_ a$ = 30\,K for $z > 20$\,AU\,$(r/100$\,AU) or $r < 150$\,AU 
\citep[the freezeout radius for this disk][]{qi11}.  Outside this 
region (i.e., the disk midplane for $r > 150$\,AU), the gas is a cooler 
$T_ m$ = 20\,K.  Since the gas high above the midplane and close to the star
should be hotter than the 30\,K typical for the outer region of this disk,
we add a hot atmosphere above $z_a = 73$AU\,$(r/200$\,AU$)$ with a power-law 
component $T_a(r,z>z_a) = T_{\rm gas} + $40K\,$(r/200$\,AU$)^{-0.8}$.  
This disk structure is shown in Figure \ref{fig:pedagogical_struct}(b), and the 
corresponding channel maps in Figure \ref{fig:pedagogical} (bottom panels).  
Qualitatively, this model does an excellent job of producing both the 
morphological and brightness asymmetries noted in \S\ref{sec:results}.

The disk structures described above are both massive and geometrically thick: 
the model scale height $H$ is much larger than what the actual midplane 
temperatures would produce.  Specifically, the defined $H(r)$ would correspond 
to a midplane temperature of $T_m = 50$K\,$(r/100$\,AU$)^{-0.3}$, which is too 
warm to reproduce the line asymmetries (too much emission would be generated, 
essentially filling in the hollow cone emission morphology).  However, the 
artificially large scale height is necessary to reproduce the morphology, since 
increasing the height of the emitting layer increases the projected separation 
of the front and back half of the disk (see \S\ref{sec:toy}).  This suggests 
that a warm layer suspended above a cold midplane may help to raise the height 
of the emitting CO layer.  It should be emphasized that a vertical temperature 
gradient is necessary to reproduce the data although it may not be the only 
contributing effect.  Depletion of the gas phase CO due to freezeout onto grains
should also reduce the contribution of the midplane emission, but this effect 
requires the cold temperatures that, by themselves, can produce this 
distinctive morphology.

%**************************
\subsubsection{A Hydrostatic Model Structure}\label{sec:hydro}
%**************************
While the previous model serves to demonstrate how a vertical temperature 
profile can explain the observed asymmetries in the $^{12}$CO $J$=3$-$2 and 
$^{12}$CO $J$=2$-$1 lines, the model structure is purely pedagogical and not 
physically self-consistent.  We now present a hydrostatic disk model where the 
densities are internally consistent with the parametrically defined 
temperatures.  This model roughly reproduces both the morphology and intensity 
of the observed emission simultaneously for the lines of interest, 
$^{12}$CO $J$=3$-$2, $^{12}$CO $J$=2$-$1, $^{13}$CO $J$=2$-$1, and 
C$^{18}$O $J$=2$-$1.

\begin{figure}[t!]
\epsscale{0.45}
\plotone{hd163296/f08.pdf}
\caption{The temperature and density structure of the hydrostatic model 
described in \S\ref{sec:hydro}, using the same legend as Figure 
\ref{fig:pedagogical_struct}.  The lower red curve indicates the upper boundary 
of the cold midplane where the gas phase CO densities are reduced due to 
freezeout onto grains.
\label{fig:hydro_struct}}
\end{figure}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f09.pdf}
\caption{The $^{12}$CO $J$=3$-$2 ({\it top}) and C$^{18}$O $J$=2$-$1 line 
emission ({\it bottom}) of the hydrostatic model described in \S\ref{sec:hydro}.
For both lines the imaged residual visibilities ($\Delta I_\nu=$ data -- model) 
are shown in color ({\it blue} is negative and {\it red} is positive). The 3, 6,
and 10\,$\sigma$ levels are indicated by vertical black lines in the colorbar.
\label{fig:hydrochmaps}}
\end{figure}

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f10.pdf}
\caption{The integrated line profiles of the four observed CO lines 
({\it gray}) compared to the model spectrum ({\it black lines}; 
\S\ref{sec:hydro}).  The spectra were calculated in 10\arcsec\,square boxes 
except for the C$^{18}$O line, which used a 6\arcsec\,square box.
\label{fig:hydrolps}}
\end{figure}

We define the gas temperatures using a slightly modified version of the 
parameterization introduced by \citet{dartois03}, where
\begin{equation}
 T_{\rm gas}(r, z) = \left\{
\begin{array}{ll}
T_a + \left(T_m - T_a\right) \left[\sin{\frac{\pi z}{2 z_q}}\right]^{2\delta(r)} & \mbox{if $z < z_q$} \\
T_a & \mbox{if $z \ge z_q$} 
\end{array}.
\right. 
\label{eq:dartoisII}
\end{equation}
Here, $T_m(r,z=0) = 19$K\,$(r/155$\,AU$)^{-0.3}$ is the temperature at the disk 
midplane and the atmosphere temperature above the height $z_q$ is 
\begin{equation}
T_a(r, z) = 55{\rm K\,}(\sqrt{r^2 + z^2}/200{\rm \,AU})^{-0.5}.
\end{equation}
This prescription produces a smooth, monotonically increasing (with $z$) 
transition between $T_m(r,z=0)$ and $T_a(r,z=z_q)$, with the vertical shape 
controlled by the parameter $\delta = 0.0034*(r-200$\,AU$) + 2.5$ (with the 
requirement that $\delta \geq 0.3$).  The height of the disk atmosphere is 
assumed to have a radial distribution described by a truncated power law, 
$z_q = 63$AU\,$(r/200$\,AU$)^{1.3} \exp[-(r/800$\,AU$)^2]$.  The midplane 
temperature profile we assume has the same CO freezout radius calculated by 
\citet{qi11}, but with a shallower power-law index.  The form of the vertical 
temperature profile (Equation \ref{eq:dartoisII}) was designed to mimic the 
dust temperatures in the disk models developed by 
\citet{dalessio98,dalessio99,dalessio01,dalessio06}.  Since we do not do the 
radiative transfer calculation for a dust disk 
\citep[e.g.,][]{qi11,andrews12}, we adopt this parameterization as a natural 
choice for the outer regions of a disk.

The vertical distribution of the gas is determined by solving the differential 
equation for hydrostatic equilibrium with our parametrically defined temperature
structure:
\begin{equation}
-\frac{\partial \ln \rho_{\rm gas}}{\partial z} =\frac{\partial\ln T_{\rm gas}}{\partial z} + \frac{1}{{c_s}^2}\left[\frac{G M_\ast z}{(r^2 + z^2)^{3/2}}\right],
\end{equation}
where ${c_s}^2 = k_B T_{\rm gas}/\mu m_h$ is the sound speed.  We normalize the 
density so that the total mass of the gas disk is 
$M_{\rm gas} = 0.09$\,M$_\odot$, and assume $\gamma = 0.8$ and $r_c=150$\,AU.  
As before,  we assume that the gas follows circular Keplerian orbits and that the 
self-gravity of the disk is negligible.   We now include freezeout for 
$T_{\rm frz} = 19$\,K and photodissociation with 
$\sigma_s = 9\times10^{20}$\,cm$^{2}$.  For the CO isotopes, we assume the 
relative abundances measured for the ISM by \citet{wilson99}: 
$^{12}$CO/$^{13}$CO$=69\pm6$ and $^{12}$CO/C$^{18}$O$=557\pm30$.  The disk 
density and temperature structure are shown in Figure \ref{fig:hydro_struct}.

In Figure \ref{fig:hydrochmaps} we present channel maps of predicted model 
$^{12}$CO $J$=3$-$2 and C$^{18}$O $J$=2$-$1 emission, with the 
imaged residuals overlaid in color.  This model successfully reproduces the 
distinctive spatial morphology of the $^{12}$CO $J$=3$-$2 line (top panels) 
while also appearing symmetric in the C$^{18}$O $J$=2$-$1 line (bottom panels). 
The majority of the C$^{18}$O emission is contributed from material
much closer to the midplane than in the case of $^{12}$CO.  The resulting 
signature of the temperature asymmetry is much less pronounced in the 
isotopologue lines, and is unresolved by these data.  We show a comparison of 
the integrated line profiles for the data and model predictions in Figure 
\ref{fig:hydrolps} for the four CO lines of interest.

While this model matches the data fairly well, there are systematic residuals at
large radii near the major axis at projected velocities of $\approx 1$\,km
s$^{-1}$ where the model does not produce any significant emission 
$(I_\nu < 3\sigma)$.  We cannot account for this emission by increasing the 
disk size ($r_c$), since emission at larger radii for channels at 
higher velocities as well as at the systemic velocity is overproduced.  One 
interpretation of these significant residuals is that the gas at large radii is 
predicted to be moving {\it too fast} and so the emission morphology of the 
model appears to ``fold'' toward the major axis too slowly (when starting at 
the line center and looking channel by channel toward the blue or red shifted 
line wings).  We address how the model may not accurately describe the true disk
velocity field in the next section.

%**************************
\subsection{A Closer Look at the Gas Velocities}
\label{sec:velocity}
%**************************
Although the models we present in \S\ref{sec:pedagogic} and \S\ref{sec:hydro} 
leave non-negligible residuals compared to the data, they nevertheless serve to 
illustrate how a disk with a realistic vertical temperature structure naturally 
explains the asymmetric morphology of the $^{12}$CO $J$=3$-$2 line.  These data 
are an example of the exquisite sensitivity that ALMA will routinely achieve for
observations of protoplanetary disks.  With this in mind, we now explore how 
these data are sensitive to the bulk motions of the gas in the HD 163296 disk, 
necessitating the use of self-consistent, physically motivated disk structures 
in any detailed analysis of its structure or kinematics.

We begin with a broad discussion of the basic expectations for gas velocities in
protoplanetary disks.  These disks are usually assumed to be rotating 
differentially and in vertical hydrostatic equilibrium, so that there is no bulk
vertical motion of the gas.  At any location in the disk, there is an orbital 
velocity,
\begin{equation}
\vec{v} = v \,\hat{\theta}
\end{equation}
where $\hat{\theta}=(\sin\theta, -\cos\theta,0)$, and the gas is assumed to be 
on circular orbits.  The standard \citep[or tested, e.g.][]{dutrey94} assumption
is that $v$ is equivalent to the Keplerian orbital velocity set by the 
gravitational potential of the central star,
\begin{equation}
\frac{{v_K}^2}{r} = \frac{G M_\ast}{r^2}.
\label{eq:fkep}
\end{equation}
Equation \ref{eq:fkep} implicitly assumes that the disk is geometrically thin, 
so that for a given $r$ the gas at the midplane ($z=0$) is moving at the same 
velocity as the gas suspended above it.  But including the vertical geometry 
actually decreases the force from the star as well as the radial projection of 
that force vector (which sets the orbital velocity).  This effectively reduces 
$v$ as $z$ increases, and implies that the disk has differential rotation in 
both the radial and vertical dimensions,
\begin{equation}
\frac{v^2}{r} = \frac{r}{(r^2 + z^2)^{1/2}} \left(\frac{G M_\ast}{r^2 + z^2}\right).
\label{eq:fgeom}
\end{equation}
Recasting in terms of the toy model described in \S\ref{sec:toy}, Equation
\ref{eq:fgeom} reduces the observed velocity of the CO emitting layer by a 
factor of $(\cos\psi)^{3/2} \sim 0.95$--$0.91$ for a cone angle, $\psi$, of 
15--20$^{\circ}$.  This rough estimate alone suggests caution when adopting a 
Keplerian, thin disk velocity field (Equation \ref{eq:fkep}) for modeling 
emission from molecules located high above the midplane.  In principle, this
assumption can be checked {\it a posteriori} for any disk model by calculating 
$\psi$ (or equivalently $z$) of the $\tau=1$ emitting surface 
(see \S\ref{sec:discussion}).

In addition to the gravitational potential of the central star, any radial 
change in the gas pressure will provide a force term and alter the gas 
velocities from the fiducial Keplerian field, $v_K$.  Including the pressure 
gradient, the force equation becomes
\begin{equation}
\frac{v^2}{r} = \frac{G M_\ast r}{(r^2 + z^2)^{3/2}} + \frac{1}{\rho_{\rm gas}}\frac{\partial P_{\rm gas}}{\partial r}.
\end{equation} 
As both the gas temperature and density tend to decrease with $r$, a 
corresponding negative pressure gradient causes the gas to slow down and orbit 
at sub-Keplerian velocities.  This phenomenon is thought to play an important 
role in the radial migration and growth of solids in these disks 
\citep{weidenschilling77b,takeuchi02,birnstiel10}, but is a subtle effect that 
slows down the gas from the fiducial Keplerian velocities by a rough factor of 
$(1-c_s^2/v_K^2)^{1/2} \sim 0.99$ for a temperature of 30\,K at $r\sim 500$\,AU 
in the disk \citep{armitage09}.  However, if the density is falling faster with 
radius than a power-law, the pressure gradient will be larger and the gas 
velocities markedly slower.  Therefore, the significance of this force term 
depends intimately upon the density and temperature structure.  The popular 
self-similar solution for $\Sigma_{\rm gas}(r)$ that we have utilized features 
an exponential tail that will enhance this effect.

The last force term we consider is the self-gravity of the disk.  Unlike the two
previous terms which tended to slow down the gas and produce sub-Keplerian 
velocities, the additional mass contributed by the disk should increase $v$.  
The composite force equation is then
\begin{equation}
\frac{v^2}{r} = \frac{G M_\ast r}{(r^2 + z^2)^{3/2}} + \frac{1}{\rho_{\rm gas}}\frac{\partial P_{\rm gas}}{\partial r} + \frac{\partial \phi_{\rm gas}}{\partial r},
\label{eq:fullF}
\end{equation} 
where $\phi_{\rm gas}$ is the potential due to self-gravity of the disk.  The 
right hand side of this equation includes three force terms: 
$F_{\rm stellar\,gravity}$, $F_{\rm pressure\,gradient}$, and 
$F_{\rm disk\,gravity}$ (appearing from left to right in Equation 
\ref{eq:fullF}).  To self-consistently construct a disk structure now requires 
iteratively solving the equation of hydrostatic equilibrium,
\begin{equation}
-\frac{\partial \ln \rho_{\rm gas}}{\partial z} = \frac{\partial\ln T_{\rm gas}}{\partial z} +  \frac{1}{{c_s}^2}\left[\frac{G M_\ast z}{(r^2 + z^2)^{3/2}} + 
\frac{\partial \phi_{\rm gas}}{\partial z}\right],
\end{equation}
and Poisson equation,
\begin{equation}
\nabla^2 \phi_{\rm gas} = 4 \pi G \rho_{\rm gas}.
\end{equation}
For most disk configurations, the Poisson equation must be calculated 
numerically which we do using a fixed grid \citep{swarztrauber75} 
{\it without} any iterations with the hydrostatic equilibrium equation (i.e., we
do not include self-gravity when calculating the disk density structure for 
reasons explained below).  As an illustration, we can roughly estimate the
contribution of this correction by considering a special case with an analytic 
solution to the Poisson equation: a thin disk with a surface density profile 
$\Sigma_{\rm gas} = \Sigma_0 r_0 /r$ and no outer edge has 
\begin{equation}
\left.\frac{\partial \phi_{\rm gas}}{\partial r}\right|_{r=R} = 2\pi G\Sigma(R)
\end{equation}
at the midplane \citep{mestel63,lodato07}.  In that scenario, the gas velocities
are increased by a factor of 
$(1+ 2\pi Gr\Sigma(r)/ v_{\rm Kep}^2)^{1/2} \sim 1.005$--$1.001$ for radii 
$\sim 100$--$500$\,AU in the disk.  We note that throughout our analysis we have
only considered circular orbits.  Including eccentricity in the disk 
\citep[e.g.,][]{regaly11} introduces a host of new parameters, asymmetries in the
integrated line profile (which we have not investigated), and is beyond the 
scope of our analysis.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f11.pdf}
\caption{The velocity difference ({\it in color}) between the thin Keplerian 
disk (${v_K}^2 = G M_\ast /r$) and the four models
considered here.  The black contours mark the 
fractional difference (in \%).  For reference, the native channel width of
the $^{12}$CO $J$=3$-$2 data is $0.11$\,km s$^{-1}$. The molecular layer of the 
model (\S\ref{sec:models}) is indicated by the dashed black border.
\label{fig:veloc}}
\end{figure}

We now explore whether these ALMA data can distinguish between models that 
include these subtle effects on the gas velocities: (1) the differential 
rotation due to the vertical geometry of the disk, (2) the sub-Keplerian 
velocities of the gas due to the radial pressure gradient, and (3) the 
super-Keplerian velocities introduced by self-gravity.  In order to disambiguate
these three effects, we will consider separately the three force terms in 
Equation \ref{eq:fullF} that determine $v$: $F_{\rm stellar\,gravity}$, 
$F_{\rm pressure\,gradient}$, and $F_{\rm disk\,gravity}$.
When modeling the last two disk-specific terms, we revert to the thin-disk 
approximation where $F_{\rm stellar\,gravity}=GM_\ast/r^2$.  We use the same 
observing geometry, disk density and temperature structure introduced in 
\S\ref{sec:hydro}, changing only the gas velocities.  As mentioned
previously, the self gravity of the disk affects both the velocity and 
density structures of the disk.  Since both of these structures will impact the 
observed line emission and we are interested here in the effect of the former 
only, we choose to omit the disk potential in the equation of hydrostatic 
equilibrium and leave the disk density structure unaltered.  For completeness, 
we also calculate the emission from a model whose velocities are determined by 
all three terms (Equation \ref{eq:fullF}).  Due to the more complicated velocity
field and the fine spatial sampling of our disk models, we found that the 
packaged raytracer in LIME was prohibitively slow.  Therefore, for these models 
we only used LIME to calculate the non-LTE level populations.  To generate the 
model images, we utilized the axisymmetry of our disk models to interpolate 
onto a fine 2D grid \citep{fan05} and integrated the radiative transfer 
equation.  None of these four models require any additional parameters.

\begin{figure}[t!]
\epsscale{0.86}
\plotone{hd163296/f12.pdf}
\caption{The synthesized channel maps of the models (shown in {\it gray}) 
with the various velocity structures described in \S\ref{sec:velocity}.  
Overplotted in color are the fiducial model $-$ model residuals.  The fiducial 
model (\S\ref{sec:hydro}) assumes that the disk is thin ($z \ll H$) with 
Keplerian orbits ($v^2 = G M_\ast/r$; $r=\sqrt{x^2 + y^2}$).  The residuals do 
{\it not} indicate the quality of the model as a fit to the data: they show how 
the model emission changes when the velocity field is changed.
\label{fig:effects}}
\end{figure}

Figure \ref{fig:veloc} shows the fractional and absolute velocity difference 
compared to the fiducial Keplerian velocity field (${v_K}^2 =G M_\ast/r$) for 
each model structure. As expected, the geometry of the disk height and the disk 
pressure gradient slow down the gas high above the midplane and at large disk 
radii ($r \gtrsim 300$\,AU).  The disk self-gravity speeds up the gas, but has 
a smaller effect than the first two terms.  Combining all three effects 
for our fiducial model, the fractional difference in the gas velocities are on 
the order of a few percent in the midplane.  However, the absolute velocity 
difference in the CO emitting region in the outer disk can be as great as 
$\sim0.1\,$km s$^{-1}$, the same as the native channel width of the Band 7 
observations.

The impact of these changes to the velocity field are significant at the 
$\gtrsim 3 \sigma$ level across many channels.  Figure \ref{fig:effects} shows 
the difference between the model with the standard Keplerian rotation 
($v^2 =G M_\ast/r$; \S\ref{sec:hydro}) and the individual models whose velocity 
fields are summarized in Figure \ref{fig:veloc}.  For all of these models, the 
central channels were the least affected since the projected velocity (for this 
inclination) is small ($v_{\rm los} \propto \cos \theta$).  The model that 
accounts for the vertical geometry of the disk differs most significantly, 
particularly for the higher velocity channels ($v \gtrsim 1.4$\,km s$^{-1}$).  
The radial pressure gradient impacts the emission at large radii, where the 
surface density profile falls off sharply.  By itself, the disk self gravity has
the least significant effect.  However, the combined model is clearly not 
dominated by any one term, indicating that models for data of this high quality 
need to incorporate all of these more physical and self-consistent calculations 
for the velocity field.  The usual assumptions can, of course, be instead tested
{\it a posteriori} as we have done here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We have conducted an analysis of multiple CO emission lines observed toward the
protoplanetary disk hosted by HD 163296.  The exquisite spatial resolution and 
sensitivity of these ALMA SV observations, combined with good spectral 
resolution of the $^{12}$CO $J=$3$-$2 line, clearly reveals that the spatial 
morphology of the emission is asymmetric across the major axis of the disk.  We 
interpret this 
asymmetry as a resolved signature of the vertical temperature gradient in the 
disk generically predicted by physical/chemical models of disk structures 
\citep[e.g.,][]{pavlyuchenkov07,semenov08}.  We developed a series of 
physically self-consistent disk structures, showing that these ALMA data can 
distinguish between models that correct the gas velocities for the vertical 
thickness of the disk and the radial pressure gradient of the gas.

We started with a toy model that can conceptually explain the asymmetric 
morphology of the $^{12}$CO $J$=3$-$2 line.  An emitting surface described by a 
rotating double cone that makes an angle $\psi \sim 15^\circ$ with the disk 
midplane naturally mimicked the observed morphology of this disk.  Such a toy 
model is analogous to a layered disk structure, where the contribution of the 
midplane emission is small due to the cold gas temperatures and depleted 
abundances produced by the freezeout of gas phase CO onto grain surfaces.  We 
confirmed the qualitative behavior of this toy model by calculating the non-LTE 
emission from both pedagogical and hydrostatic axisymmetric disk models, 
concluding with a disk structure that roughly reproduced the observed morphology
of the $^{12}$CO $J$=3$-$2 emission.  Our analysis also showed that a vertical 
temperature gradient, which has been reported in this disk by both 
\citet{qi11} and \citet{akiyama11}, was consistent with the observed 
intensities of the $J$=2$-$1 emission from $^{12}$CO, $^{13}$CO, and C$^{18}$O, 
as well as the symmetric morphology of the rarer isotopologue lines.

Spectral line emission, even when it is spatially resolved, cannot easily define
the absolute orientation of the disk: there is a 180$^\circ$ ambiguity in the 
inclination and position angle.  However, the morphological and brightness 
asymmetry highlighted here resolves the absolute orientation of the disk.  We 
report values of the disk inclination and position angle that differ by 
180$^\circ$ from previous analyses based on less sensitive data 
\citep{isella07,hughes08,qi11}.  For HD 163296, an independent check on the 
viewing geometry is provided by its outflow, where Doppler shift measurements 
of H$\alpha$, [S II], O III], [N II] emission lines and imaging can 
differentiate between the jet (blue shifted, seen on the near side of the disk) 
and counter jet (red shifted, partially hidden behind the disk).  
For HD 163296, the blueshifted jet is oriented in the SW direction while the 
redshifted counter jet is moving NE \citep[in agreement with the disk wind 
observed via $^{12}$CO by][]{klaassen13}.  This is consistent with our reported 
inclination axis, which tilts the NE part of the disk toward the observer.  
Furthermore, the projecion of the disk rotation axis on the observer sky plane 
(42$^\circ$) is aligned with the PA of the jet \citep[measured for the 
counter jet as $42.0 \pm 3.5^\circ$;][]{grady00}.  The 
jet inclination \citep[$51_{-9}^{+11}$$^\circ$;][]{wassell06,gunther13} also 
agrees with the disk inclination within the $1\sigma$ error bars.  We confirm 
that the jet appears to be orthogonal to the disk plane, although the 
large error bars in the jet inclination leaves open the possibility that the
jet is slightly skewed ($\Delta i \lesssim 10^\circ$).

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f13.pdf}
\caption{Map renderings of the angle, $\psi = \tan^{-1}(z/r)$, that the 
$\tau = 1$ surface of the $^{12}$CO $J$=3$-$2 ({\it top}) and $^{12}$CO 
$J$=2$-$1 ({\it bottom}) model emission makes with the disk midplane (for the 
fiducial model described in \S\ref{sec:hydro}). 
\label{fig:psi}}
\end{figure}

A subtle, but important, point is that spatially resolving the asymmetry caused
by the vertical temperature gradient requires that the data has both excellent 
spatial {\it and spectral} resolution.  In Figure \ref{fig:psi} we show a 
pixel map of the angle, $\psi = \tan^{-1}(z/r)$, that the $\tau=1$ surface of 
our fiducial model (\S\ref{sec:hydro}) makes with the disk midplane for the 
$^{12}$CO $J$=3$-$2 and $^{12}$CO $J$=2$-$1 emission.  Both of these lines 
originate at nearly the same height, and both models exhibit an asymmetric 
emission morphology at a $\Delta v = 0.1$km s$^{-1}$ resolution.  However,
when imaged at the ALMA SV resolutions, the $J$=2$-$1 line appears mostly 
symmetric -- just like the data (exceptfor the slight rotation at 
$|v| \approx 1.7$\,km s$^{-1}$).  Figure \ref{fig:spec_effects} shows the 
accumulated degradation of the spectral 
resolution for $^{12}$CO $J$=2$-$1 model emission.  The spectral averaging 
(middle panel), Hanning smoothing (right panel), and Fourier sampling (bottom 
row) all diminish the apparent morphological asymmetry even though the first 
two effects are operations along the spectral dimension.  The muted asymmetry of
the $^{12}$CO $J$=2$-$1 emission can be attributed to the larger beam size and 
coarser spectral resolution of the Band 6 observations and {\it not} to the disk
structure (as is the case for the rarer $^{13}$CO and C$^{18}$O isotopologues). 
The Band 6 data have been averaged and smoothed in the spectral dimension (see 
\S\ref{sec:models}), which smooths the correlated emission in the spatial 
dimension \citep{thompson86}.  The consequence of a coarse spectral resolution 
is worth emphasizing, since in practice it may significantly degrade the utility
of the excellent spatial resolution of ALMA.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{hd163296/f14.pdf}
\caption{This set of panels show the effect that the smoothing and averaging 
in the spectral dimension has on the apparent asymmetry of a model.  The top 
row features the high fidelity $^{12}$CO $J$=2$-$1 model observed at 
$v=-1.36$\,km s$^{-1}$ ({\it left}) which is then spectrally averaged over 
a spectral bin width of $\Delta v = 0.32$\,km s$^{-1}$ ({\it center}), and then 
Hanning smoothed ({\it right}). The bottom row shows the synthesized image for 
each model and all panels are shown on the same color scale with
$5\sigma$ interval contours marked on the color bar.
\label{fig:spec_effects}}
\end{figure}

We explored the sensitivity of the Band 7 observations to the gas kinematics
with a series of models where only the bulk velocities were changed. 
We found that these data could distinguish between models for which those gas 
velocity differences were $\gtrsim 5$\%, corresponding to classes of models that
incorporated the vertical geometry of the disk as well as the radial pressure 
gradient in the gas.  Indeed, a few disks have been identified as 
having sub-Keplerian velocities \citep{wang12}, and we suggest that these 
phenomena provide an intuitive explanation (although might not be sufficient for
such systems).  For our fiducial model, the disk self-gravity did not appear to 
significantly alter the velocity field, but may be important when modeling the 
density structure.  Furthermore, the contributions from the gas pressure 
gradient and self gravity are model dependent, and so may become even more 
important for certain systems \citep[e.g.,][]{cesaroni05,bergin13}.

There are a few aspects of the disk hosted by HD 163296 that accentuate the 
subtle effects we have addressed.  First, the disk is both large and relatively 
nearby so that the longest baselines in this SV dataset ($\sim 400$k$\lambda$) 
correspond to spatial scales that will only be attained by full ALMA for 
T Tauri disks in more distant star-forming clusters.  Second, HD 163296 
is an A star that can passively heat its disk out to large radii.  And so in 
addition to having a large radial extent and thick vertical geometry, the disk 
is warm and bright for sub-millimeter observations.  Lastly, the intermediate 
disk inclination helps to evenly project both the vertical and radial dimension 
of the disk on the sky plane \citep{semenov08}.  The inclination also projects 
a significant fraction of the intrinsic gas motions along the line of sight.
This reduces the smoothing effect that the spectral resolution has upon the 
spatial pattern of the line (see Figure \ref{fig:spec_effects}).

With high-resolution, sensitive instruments such as ALMA and the JVLA, there is 
a host of opportunities for studying circumstellar disks and phenomena 
associated with planet formation
\citep[e.g.][]{wolf05,semenov08,cossins10,cleeves11,gonzalez12,ruge13}.  These 
observations demonstrate the necessity of more sophisticated, physically 
self-consistent approaches when analyzing data from this new generation of 
observing facilities.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}\label{sec:summary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have analyzed sensitive, sub-arcsecond resolution observations of
the $^{12}$CO $J$=3$-$2, $^{12}$CO $J$=2$-$1, $^{13}$CO $J$=2$-$1, and 
C$^{18}$O $J$=2$-$1 emission lines from the protoplanetary disk hosted by HD 
163296.  The key conclusions of our analysis are:
\begin{enumerate}
\item The $^{12}$CO $J$=3$-$2 spectral line features a clear and systematic
morphological and brightness asymmetry across the major axis of the disk.  The 
$^{12}$CO $J$=2$-$1 line exhibits similar, but muted, behavior.  No asymmetries
are observed for the rarer $^{13}$CO and C$^{18}$O isotopologues.
\item  The resolved morphological and brightness asymmetries of the 
$^{12}$CO $J$=3$-$2 line emission as well as the symmetric emission of the 
rarer isotopologues are the signatures of a vertical temperature gradient in the
disk.  A double cone structure encapsulates the salient physical features of a 
disk with a cold midplane and warm atmosphere and mimics the distinctive
emission morphology by moving the surface that dominates the emission above 
(and below) the disk midplane.
\item We presented a series of simple disk structures that demonstrated how the
observed data can be roughly reproduced by a disk with a cold midplane and 
warm atmosphere.  For our fiducial model, the $\tau=1$ surface for the 
$^{12}$CO emission was suspended $\sim 15^\circ$ above the midplane, agreeing 
with our morphological characterization using the toy double cone structure.
\item The excellent spatial and spectral resolution of the Band 7 observations
are suprisingly sensitive to the bulk velocities of the gas and can distinguish 
between models that include the vertical geometry of the disk
and radial pressure gradient (a fractional difference in the bulk gas velocity 
field of $\gtrsim 5$\%).  The inclusion of self-gravity for our fiducial disk 
model is a less important correction, but may become dominant for other disk 
models.
\item  The excellent sensitivity and spatial resolution of these ALMA SV data
require careful processing of models.  In particular, coarse spectral 
resolution of the data can strongly impact the spatial morphology of the 
observed emission.
\end{enumerate}

\section*{Chapter Acknowledgements}
We are grateful to Matt Payne, Diego Mu{\~n}oz, Eugene Chiang, David Knezevic, 
Moritz G{\"u}nther, Marc Metchnik, and Kees Dullemond for insightful 
conversations.  This paper makes use of the following ALMA data: 
ADS/JAO.ALMA\#2011.0.00010.SV.  ALMA is a partnership 
of ESO (representing its member states), NSF (USA) and NINS (Japan), together 
with NRC (Canada) and NSC and ASIAA (Taiwan), in cooperation with the Republic 
of Chile.   The Joint ALMA Observatory is operated by ESO, AUI/NRAO and NAOJ.  
We acknowledge support from NASA Origins of Solar Systems grant No. NNX11AK63.

\clearpage

\begin{table}
\begin{center}
\centering
\caption{Emission Line Results \label{tab:lines}}\medskip
\centering
\begin{threeparttable}
  \renewcommand\TPTminimum{\linewidth} %%% we want full width
\centering
\begin{tabular}{lcccc}
\toprule
Parameters & $^{12}$CO(3-2) & $^{12}$CO(2-1) & $^{13}$CO(2-1) & C$^{18}$O(2-1) \\
\midrule
Channel Width$^a$ [km s$^{-1}$] & 0.11 & 0.32 & 0.33 & 0.33 \\
Beam & $0\farcs65\times0\farcs42$ & $0\farcs81\times0\farcs66$ & $0\farcs87\times0\farcs70$ & $0\farcs87\times0\farcs70$ \\
P.A. & $-87\fdg3$ & $75\fdg8$ & $76\fdg8$ & $76\fdg8$\\
RMS [Jy beam$^{-1}$] & 0.022 & 0.016 & 0.009 & 0.008 \\
Integrated intensity [Jy km s$^{-1}$] & $109\pm11$ & $46\pm5$ & $18\pm2$ & $5.8\pm 0.6$\\
Peak flux [Jy beam$^{-1}$]& 0.86 & 0.87 & 0.45 & 0.23 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[$^a$]{The velocity resolution indicated by the channel width 
is smaller than the true velocity resolution as the data are Hanning 
smoothed along the spectral dimension \citep{lundgren12}.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

### 2015-07-03 ms_radial_092713/ms_radial_chapter.tex
initial commit

\chapter{Fast Radial Flows in Transition Disk Holes}\label{c:radial_flows}

\section*{abstract}

Protoplanetary ``transition'' disks have large, mass-depleted central cavities,
yet also deliver gas onto their host stars at rates comparable to disks
without holes.  The paradox of simultaneous transparency and accretion can
be explained if gas flows inward at much higher radial speeds inside the cavity 
than outside the cavity, since surface density (and by extension optical depth) 
varies inversely with inflow velocity at fixed accretion rate.  Radial speeds 
within the cavity might even have to approach free-fall values to explain the 
huge surface density contrasts inferred for transition disks.  We identify 
observational diagnostics of fast radial inflow in channel maps made in 
optically thick spectral lines.  Signatures include (1) twisted isophotes in 
maps made at low systemic velocities and (2) rotation of structures observed 
between maps made in high-velocity line wings.  As a test case, we apply our 
new diagnostic tools to archival ALMA data on the transition disk HD 142527, 
and uncover evidence for free-fall radial velocities inside its cavity. 
Nevertheless the observed kinematics are also consistent with a disk warp;
disentangling the two scenarios calls for detecting optically thin
line emission to measure the low surface densities expected to
accompany fast inflow.  How material in the disk cavity sheds its
angular momentum wholesale to fall freely onto the star
is an unsolved problem; gravitational torques exerted by giant planets or
brown dwarfs are discussed as a candidate mechanism.

\section{Introduction} \label{intro}

A subset of T Tauri and Herbig Ae disks have large, mass-depleted cavities in 
their inner regions, inferred from their infrared spectra 
\citep[e.g.,][]{Kim13} or spatially resolved imaging at infrared 
\citep[e.g.,][]{Geers2007} and mm wavelengths \citep[e.g.,][]{Brown09}.  The
cavities in these ``transition'' disks have radii ranging from $\sim$4\,AU 
\citep[TW Hya;][]{Calvet02,Hughes07} to 20\,AU \citep[GM Aur;][]{hughes09} to 
40--50\,AU \citep[LkCa 15;][]{Andrews11,Isella12} to 140\,AU \citep[HD 
142527;][]{casassus13}.  The suspicion is that dynamical interactions with 
nascent planetary systems evacuate the inner disk regions 
\citep[e.g.,][]{Zhu11}.

Surprisingly, despite their optically thin cavities, many transition disks 
appear to be accreting gas at fairly normal rates.  Stellar accretion rates 
$\dot{M}_\ast$ in transition disks range from 
$\sim$$10^{-9}\,M_\odot\,{\rm yr}^{-1}$ \citep[TW Hya;][]{muzerolle00} to 
$10^{-8} \,M_\odot\,{\rm yr}^{-1}$ \citep[GM Aur;][]{gullbring98,Calvet05} to 
$10^{-7}\, M_\odot\,{\rm yr}^{-1}$ \citep[HD 142527;][]{garcia06}.  These rates 
overlap with those for conventional disks without cavities, although the median
$\dot{M}_\ast$ for transitional disks may be a factor of $\sim$10 lower than 
for conventional disks \citep{najita07,espaillat12,Kim13}.

It is sometimes thought that transition disk cavities are created by planets 
consuming material from the outer disk and thereby starving the inner disk.  
This is unlikely to be correct --- or at least it cannot be the whole story --- 
because a starved inner disk would also starve the host star, contrary to the 
observed values of $\dot{M}_\ast$ cited above.  Modeling of disk spectra and 
images reveals that surface density contrasts inside and outside cavities can 
be $\gtrsim 10^3$ \citep[e.g.,][]{Calvet05}.  Planets inside the cavity cannot
simultaneously eat $\gtrsim 99.9$\% of the mass flowing inward from the outer 
disk and still leave enough for their host star to accrete at near-normal 
rates.  If we interpret the factor-of-10 lower $\dot{M}_\ast$ for transition 
disks as arising from planets consuming $\sim$90\% of the accretion flow 
from the outer disk and leaving the remaining $\sim$10\% for the central star 
\citep[e.g.,][]{lubow06}, then the surface density just inside the planet's 
orbit would decrease by only a factor of 10, which is almost certainly too 
small to render the inner disk optically thin --- all other factors being 
equal.

In this paper, we describe a way for transition disks to have their cake 
(possess optically thin cavities) and eat it too (accrete at normal rates).  
The dilemma can be resolved by (somehow) increasing the radial accretion 
velocity $v_R$ inside the cavity, thereby concomitantly lowering the surface 
densities $\Sigma$ there.  For a given radius $R$ and steady disk accretion 
rate $\dot{M} = 2\pi R \Sigma v_R$, we have $\Sigma \propto 1/v_R$.  Just 
outside the cavity, $\Sigma$ is large because accretion in the outer disk is 
diffusive (``viscous") and slow ($v_R \sim \alpha (H/R)^2 v_{\rm K} \ll v_{\rm
 K}$, where $H$ is the disk vertical thickness, $\alpha \ll 1$ is the 
Shakura-Sunyaev stress parameter, and $v_{\rm K}$ is the Keplerian orbital 
velocity).  Just inside the cavity, the accretion velocity $v_R$ is somehow 
boosted --- $v_R$ could, in principle, approach its maximum of 
$\sim$$v_{\rm K}$ --- so that $\Sigma$ is lowered, conceivably by orders of 
magnitude, while maintaining the same mass flow rate $\dot{M}$ as in the outer 
disk.

Gravitational (as opposed to magnetic or hydrodynamic) torques are an obvious 
candidate for catastrophically removing angular momentum from gas inside the 
cavity.  In galaxies, gravitational torques are exerted by stars that arrange 
themselves into non-axisymmetric patterns.  A stellar bar (a perturbation with
azimuthal wavenumber $m=2$) eviscerates the interior of the $\sim$3-kpc 
molecular ring at the center of our Galaxy \citep{Schwarz81,Binney91}; an 
eccentric stellar disk ($m=1$) may be funneling gas onto the star-forming 
nucleus of M31 \citep{Chang07}; and a complex nest of $m=1$ and $m=2$ stellar 
potentials force-feeds gas onto quasars \citep{Shlosman89,Hopkins11}.  In these 
galaxy-scale examples, radial inflow velocities can approach free-fall rates.

The role of stars in galaxies might be played by planets in transition disks.  
We like this interpretation, but it is not without difficulty.  The cavities 
are so large (they are not annular gaps) and so clean that appeals are made to 
rather extreme parameters.  As many as four planets, each weighing up to 
$\sim$10 $M_{\rm Jup}$, have been posited to carve out a given cavity 
\citep{Zhu11,Dodson11}.\footnote{Even multiple super-Jovian planets are 
reportedly not enough: grain growth and dust filtration at the cavity rim are 
invoked in addition to planets to make the cavity sufficiently transparent 
\citep{Zhu11,Dong12}.}  At least in the case of the transition disk around LkCa 
15, there might be one (but so far only one) actively accreting super-Jovian 
planet located well inside the disk cavity \citep{Kraus12}.  Some assistance in 
driving the accretion flow inside the cavity may be obtained from the 
gravitational potential of the outer disk itself, if that disk is sufficiently 
clumpy (for examples of clumpy disks, see IRS 48 or HD 142527; 
\citealt{vandermarel13} and \citealt{casassus13}, respectively).  Just inside a 
non-axisymmetric outer disk, gas streamlines can cross, shock, and plunge 
inward \citep[e.g.,][]{pac77}.

Disk radial velocities driven by embedded massive planets or brown dwarfs can 
be healthy fractions of orbital (azimuthal) velocities.  We derive a crude 
estimate\footnote{Our ``zero-dimensional'' analysis ignores the fact that 
actual disk flows around planets are two-dimensional and dominated by
non-axisymmetric streams flowing in and around planetary Hill spheres.
Still, some indirect empirical support for the scalings in equation 
(\ref{eqn:simple}) has been found by \cite{Fung13}.
}
 for $v_R$ by equating the one-sided Lindblad torque --- the amount of 
angular momentum transmitted outward across a planet's orbit 
\citep{goldreich80} --- with the angular momentum advected inward by accretion, 
$\dot{M} v_{\rm K} R$.  This balance yields
\begin{equation}
v_R \sim \left( \frac{M_p}{M_\ast} \right)^2 \left( \frac{H}{R} \right)^3 v_{\rm K} \label{eqn:simple}
\end{equation}
where we have used the total linear Lindblad torque, integrated over all 
resonances outside the torque cut-off zone at distances $\gtrsim H$ away from 
the planet (see, e.g., \citealt{goodman01} or \citealt{crida06}, and references 
therein).  For a planet mass $M_p \sim 10\,M_{\rm Jup}$, a host stellar mass 
$M_\ast \sim 1\,M_\odot$, and a typical disk aspect ratio $H/R \sim 0.1$, 
equation (\ref{eqn:simple}) evaluates to
\begin{equation}
v_R \sim 0.1 \,v_{\rm K} \sim 0.5 \left( \frac{30\,{\rm AU}}{R} \right)^{1/2}\,{\rm km} \,\, {\rm s}^{-1}. 
\label{eqn:simple1}
\end{equation}
Modern heterodyne receivers on (sub)millimeter wavelength interferometers 
routinely observe disks at $\sim$0.1 km s$^{-1}$ resolution, and the Atacama 
Large Millimeter Array (ALMA) is expected to provide high sensitivity data at 
finer velocity scales yet.  Evidence for free-fall radial velocities can be 
found in the HCO$^+$ filament within the disk cavity around HD 142527.  
\citet{casassus13} discovered this filament to be oriented nearly radially; in 
estimating $\dot{M}$ for this system, they assumed $v_R \sim v_{\rm K}$ 
(although the velocities of the HCO$^+$ filament were not actually quoted in
their study).

Regardless of what is actually accelerating the accretion, we ask here whether 
we can test observationally the idea that transition disk cavities have fast 
(nearly free-fall) radial flows.  We present in \S \ref{sec:twist} some 
predicted observational signatures of radial inflow in spectrally and spatially 
resolved (sub)millimeter-wavelength maps of molecular emission lines.  A sample
case study based on the HD 142527 disk is presented in \S \ref{sec:prospects}.  
We conclude in \S \ref{sec:conclude}.

\section{Observational Signature of Radial Inflow} \label{sec:twist}

Interferometers observing the (sub)millimeter and radio spectral line 
transitions of molecular gas tracers produce a spatially resolved map of the 
source for each frequency (velocity) channel.  A disk in Keplerian rotation has 
well-ordered gas velocities, and so line emission in a particular channel
``highlights'' disk material that is appropriately Doppler shifted 
\citep[e.g.,][]{omodaka92,beckwith93}.  Adding a fast radial inflow to the 
normal Keplerian velocity field alters the morphology of channel map emission
\citep[cf.][]{tang12}.  We identify two related observational signatures of 
radial inflow in a disk cavity: (1) twisted isophotes in the channel map
corresponding to the systemic velocity (\S \ref{subsec:isotwist}), and (2) 
rotation of features seen between maps made in high-velocity wings (\S\ref{subsec:wingrot}).
While distinctive, neither of these features is entirely unique, and must be 
disentangled from other non-Keplerian effects (\S \ref{subsec:disentangle}).

Our analysis uses two axisymmetric models: a razor-thin disk and a 
three-dimensional (3D) disk.  The razor-thin model is a pedagogical tool to 
develop intuition.  The 3D model is more realistic and presents a platform to 
test the razor-thin model as well as analyze mock data products.

\begin{figure}[t!]
\epsscale{1.}
\plottwo{ms_radial_092713/f01.eps}{ms_radial_092713/f02.eps}
\caption{
{\it Left}: Regions of a razor-thin disk model ($i=20^\circ$, $\chi=0$) where 
the velocities $v'_{\rm los}$ fall within each labeled spectral channel of 
width $\Delta v'_{\rm los} = \pm \,0.212$ km s$^{-1}$.  {\it Right}: Simulated 
$^{12}$CO $J$=3$-$2 channel maps for a 3D disk with the same $i$ and $\chi$.  
Projected major (horizontal) and minor (vertical) axes of the disk are shown as 
dashed red lines.  The spatial scales of all panels are identical for a 
distance of 145\,pc.  The morphology of line emission in the channel maps 
(right) respects the velocity field (left).  The synthesized beam for the 
simulated 3D model is 0\farcs12$\times$0\farcs09, with the 
beam major axis oriented at position angle PA = $-3^\circ$. \label{fig:models}}
\end{figure}

Both models treat the gas velocities within the cavity ($R<R_{\rm cav}$) as 
having an inward radial component that is some constant fraction $\chi$ of the 
local circular Keplerian velocity.  The velocity field is then
\begin{align}
v_x =& -v_{\rm K} \sin\theta - \chi v_{\rm K} \cos \theta \\
v_y =& +v_{\rm K}  \cos\theta - \chi v_{\rm K} \sin \theta \,,
\end{align}
both for the razor-thin model and the 3D model.  Here $\theta$ is the azimuthal 
angle in the disk plane, and variations in bulk velocity with height $z$ are 
ignored (but see \S \ref{subsec:disentangle}).  Outside the cavity, gas is 
assumed to follow circular Keplerian orbits.  Our assumption that the radial 
inflow is axisymmetric is probably unrealistic --- certainly so if the inflow 
is driven by large-scale gravitational torques as we have described in 
\S\ref{intro}.  Nevertheless we assume axisymmetry to gain a foothold on the 
problem.

The disk is inclined by a viewing angle $i$ about the $x$ axis, where 
$i=0^\circ$ corresponds to a disk seen face-on.  Note that the effects we will 
discuss are sensitive to the spin orientation of the disk, and so $i$ can range 
between $0^\circ$ and $180^\circ$ without degeneracy.  Inclinations $0^\circ < i < 
90^\circ$ imply that the line-of-sight component of the orbital angular momentum 
points toward the observer, whereas $90^\circ < i < 180^\circ$ implies 
that the line-of-sight component points away.

For the razor-thin model, the projected line-of-sight velocity is\footnote{We 
can set $v'_{\rm los}$  equal to either $+ v_y \sin i$ or $- v_y \sin i$.  The 
degeneracy arises because we can choose the disk's ascending ($\equiv$ toward 
the observer) node on the sky plane to coincide with either the $x=x'$ axis 
(our convention in this paper) or the $-x = -x'$ axis.  Another way to say this 
is that $i$ is degenerate with $-i$.}
\begin{equation}
v'_{\rm los}= -v_y \sin i = -v_{\rm K}\sin i\cos\theta+\chi v_{\rm K} \sin i\sin\theta.
\label{eq:vlos}
\end{equation}
We utilize primed coordinates to denote values measured by the observer; the 
spatial coordinates in the sky plane are $x'=x$ and $y'=y\cos i$.  The azimuth 
$\theta$ measured in the disk plane is related to the on-sky angle $\theta'$ by 
$\tan\theta'= \cos i \tan \theta$.  The left panel of Figure \ref{fig:models} 
shows the morphology of line emission for a razor-thin disk with no inflow.

We compute an analogous 3D disk model that simulates the $^{12}$CO $J$=3$-$2 
line emission from a disk with a total gas mass of 0.01\,$M_\odot$, a spatially 
constant CO:H$_2$ number-abundance fraction of $10^{-4}$, and a Gaussian 
vertical density distribution ($\exp(-z^2/2H^2)$ with a scale height of 
$H=0.1R$).  The surface density profile is that of a thin, viscously accreting 
disk \citep{bellpringle74}: $\Sigma\propto R^{-1}\exp(-R/R_s)$ with scale 
radius $R_s=50$ AU and an outer truncation radius of 300 AU.  Since our focus 
in this paper is on the velocity field and not on the density structure, we 
have not included any reduction of $\Sigma$ within the cavity.  This 
simplification should not introduce serious error as long as the cavity region 
remains optically thick in $^{12}$CO emission, as is apparently the case in 
reality \citep{bruderer13}.  The model is vertically isothermal with a 
power-law radial temperature profile: $T(R) = 25\,(R/100$\,AU$)^{-0.5}$\,K.  
The mass of the central star is 1\,$M_\odot$, the distance to the observer is 
145\,pc, and the cavity radius is $R_{\rm cav} = 50$\,AU.  The disk inclination
$i$ and radial inflow parameter $\chi$ are variable.  We use the modeling 
capabilities of {\tt LIME} \citep{brinch10} to simulate high-resolution channel 
maps assuming LTE conditions and a turbulent linewidth of $\xi = 
0.01$\,km\,s$^{-1}$.  We then use {\tt CASA} to sample and deconvolve the 
Hanning-smoothed visibilities with a spectral resolution of 0.212\,km\,s$^{-1}$ 
and a spatial resolution of 0\farcs1.  The right panel of Figure 
\ref{fig:models} shows the channel maps from a 3D disk model ($\chi=0$), zoomed 
into the inner $\pm$0\farcs5.  Comparing the left and right panels of Figure 
\ref{fig:models} reveals how the velocity field of the razor-thin model offers 
a guide to the morphology of line emission in the channel maps of the 3D model 
\citep[see also][]{beckwith93}.

\begin{figure}[t!]
\epsscale{1.0}
\plotone{ms_radial_092713/f03.eps}
\caption{Twisted isophotes at the systemic velocity. {\it Top}: Observed 
velocity fields $v'_{\rm los}$ for razor-thin disks with varying degrees of 
radial inflow.  Within the cavity radius $R_{\rm cav}=50$\,AU, gas flows 
axisymmetrically inward with a velocity $\chi$ times the local circular 
Keplerian velocity.  The dotted gray circle traces the cavity boundary, the 
solid white bar indicates $\theta'_{\rm sys}$, and the black arc traces $\Delta 
\theta'_{\rm sys} = \theta'_{\rm sys} - 90^\circ$.  Dashed lines show the 
projected major (horizontal) and minor (vertical) axes.  {\it Bottom}: 
Simulated channel maps at the systemic velocity for 3D disks having the same 
$i$ and $\chi$ as the razor-thin models shown in the top panels.  The isophotes 
twist at the location of the cavity radius ($\sim$0\farcs34 from the map 
centers). \label{fig:thetasys}}
\end{figure}

\begin{figure}[t!]
\epsscale{1.0}
\plottwo{ms_radial_092713/f04.eps}{ms_radial_092713/f05.eps}
\caption{Rotation angles $\Delta \theta'_{\rm sys}$ ({\it left}) and $\Delta 
\theta'_{\rm wing}$ ({\it right}) for $\chi = 0$ (black), 0.1 (red), and 1 
(blue) in 3D disk models with $i=20^\circ$ and $R_{\rm cav} = 50$ AU.  $\Delta 
\theta'_{\rm sys}$ is measured in the systemic velocity channel ($v'_{\rm los} 
= 0$) as a function of radius, and $\Delta \theta'_{\rm wing}$ is derived 
against channel velocity.  Analytic predictions from razor-thin disk models are 
drawn with horizontal dashed lines.  The lower abscissa of the left panel plots 
projected distance along the disk's major axis, while the upper abscissa of the 
right panel plots the radius for which $v_{\rm K} \sin i = v'_{\rm los}$.  
\label{fig:vsys}}
\end{figure}

\subsection{Twisted Isophotes at the Systemic Velocity} \label{subsec:isotwist}

The gas with no bulk motion relative to the systemic velocity of the disk has 
$v'_{\rm los} = 0$. From equation (\ref{eq:vlos}), this gas satisfies 
\begin{equation}
0=-v_{\rm K}\sin i\cos\theta_{\rm sys}+\chi v_{\rm K} \sin i\sin\theta_{\rm sys} \,,
\end{equation}
which implies that $\theta_{\rm sys}=\arctan(\chi^{-1})$ in the disk plane, and 
$\theta'_{\rm sys} = \arctan(\chi^{-1} \cos i)$ on the sky.  For $\chi = 0$ (no 
radial inflow), $\theta'_{\rm sys}=90^{\circ}$: the systemic velocity channel 
probes gas that lies along the minor axis of the disk as seen in projection 
(see Figure \ref{fig:models}, central panels).  When $\chi \neq 0$, the 
isovelocity contours at $v'_{\rm los} = 0$ will be rotated by an angle $\Delta 
\theta'_{\rm sys} = \theta'_{\rm sys} - 90^\circ$.  This rotation is evident in 
the top set of panels in Figure \ref{fig:thetasys}, which show the 
line-of-sight velocity fields of a razor-thin disk for various values of $\chi$ 
inside the disk cavity.  The rotation can also be seen in the isophotes 
observed at the systemic velocity of the corresponding 3D model; the bottom 
panels of Figure \ref{fig:thetasys} show that the isophotes in the channel maps 
``twist'' at $R = R_{\rm cav}$.

The isophote twist in the systemic velocity channel map can be measured as a 
function of radius.  We perform this measurement by first dividing the 3D disk 
model into concentric annuli, and further splitting each annulus into its top 
($y'>0$) and bottom ($y'<0$) halves.  For each half-annulus, an 
intensity-weighted centroid (i.e., a ``center of brightness'') is computed.  
The rotation angle $\Delta \theta'_{\rm sys}$ is then evaluated from the line 
joining the centroid positions of the top and bottom halves of a given annulus 
at radius $R$.  The left panel of Figure \ref{fig:vsys} shows $\Delta 
\theta'_{\rm sys} (y')$ for three values of $\chi$.  Note how $\max |\Delta 
\theta'_{\rm sys}|$ is accurately predicted by the razor-thin model, at least 
for the example inclination angle $i = 20^\circ$.  In the mock observations of 
the 3D model, the transition between $\Delta\theta_{\rm sys}'=0$ and $\max 
|\Delta\theta'_{\rm sys}|$, occurring near the cavity radius ($R_{\rm cav} = 
50$\,AU), is smeared by the 0\farcs1 beam.

\begin{figure}[t!]
\epsscale{1.00}
\plotone{ms_radial_092713/f06.eps}
\caption{Wing rotation. {\it Top}: Rotation angles $\Delta
\theta'_{\rm wing}$ (white arcs), overlaid on velocity fields of razor-thin 
disk models.  Other symbols are the same as in Figure \ref{fig:thetasys}.  Note 
how $\Delta \theta'_{\rm wing} \neq \Delta \theta'_{\rm sys}$.  {\it Bottom}: 
Simulated channel maps for various line-of-sight velocities from corresponding 
3D disk models.  Comparison of $\chi=0$ with $\chi=1$ shows how wing emission 
brightens and rotates clockwise from the projected major axis. 
\label{fig:thetawing}}
\end{figure}

\subsection{Maximum LOS Velocities: Wing Rotation} \label{subsec:wingrot}

We now consider gas moving at the highest line-of-sight velocities, responsible 
for emission in the spectral line wings.  Maximizing equation (\ref{eq:vlos}) 
with respect to $\theta$ yields
\begin{equation}
0=v_{\rm K}\sin i\sin\theta_{\rm wing}+\chi v_{\rm K}\sin i\cos\theta_{\rm wing} \,,
\label{eq:wingrot}
\end{equation}
or $\theta_{\rm wing} =\arctan(-\chi)$ in the disk plane and $\theta'_{\rm 
wing} = \arctan(-\chi \cos i)$ on the sky.  Radial inflow causes the maximum 
isovelocity contour to rotate away from the disk major axis by $\Delta 
\theta'_{\rm wing} = \theta'_{\rm wing} (\chi) - \theta'_{\rm wing} (\chi = 0) 
= \theta'_{\rm wing} (\chi)$.  Figure \ref{fig:thetawing} shows the velocity 
fields for three razor-thin disk models, together with simulated channel maps 
for the corresponding 3D models.  Where there is inflow, emission from the 3D 
model is systematically rotated away from the disk major axis.  For extreme 
values of $\chi \sim 1$, line-of-sight velocities within the cavity are 
noticeably greater than for the case with no inflow (see top panels of Figure 
\ref{fig:thetawing}).  Correspondingly, emission from the line wings is 
brighter --- compare channels of the same velocity between the $\chi = 0$ and 
$\chi = 1$ models at fixed $i = 20^\circ$ (bottom panels of Figure 
\ref{fig:thetawing}) --- although it should be remembered that any decrease in 
the surface density inside the cavity (not modeled here) may reduce the 
intensity enhancement.  The last panel of Figure \ref{fig:thetawing} 
demonstrates that $\Delta\theta'_{\rm wing} \neq \Delta\theta'_{\rm sys}$.  
This difference is further quantified in Figure \ref{fig:comparetheta}, which 
shows that $|\Delta\theta'_{\rm wing} -\Delta\theta'_{\rm sys}|$ increases with 
increasing $i$ and $\chi$.  Note how $|\Delta\theta'_{\rm wing}|$ is reduced 
for highly inclined disks, just opposite to the behavior of 
$|\Delta\theta'_{\rm sys}|$.  Clearly, disk orientation is a consideration when 
designing observations to detect either rotational signature.

\begin{figure}[t!]
\epsscale{0.50}
\plotone{ms_radial_092713/f07.eps}
\caption{Rotation angles $\Delta \theta'_{\rm sys}$ (blue) and $\Delta 
\theta'_{\rm wing}$ (red) for razor-thin disk models as a function of the 
radial inflow parameter $\chi$.  Solid and dashed curves are for disk 
inclinations $i = 20^\circ$ and 60$^\circ$, respectively.  At fixed $\chi$ and 
$i$, $|\Delta \theta'_{\rm sys}| > |\Delta \theta'_{\rm wing}|$.  Measuring 
$\Delta \theta'_{\rm sys}$ is easier at low inclinations, while measuring 
$\Delta \theta'_{\rm wing}$ is easier at high inclinations. 
\label{fig:comparetheta}}
\end{figure}

To measure $\Delta \theta'_{\rm wing}$ for the 3D disk model, we employ a 
procedure similar to the one we devised for $\Delta \theta'_{\rm sys}$.  For 
each spectral channel, we measure the position of the intensity-weighted 
centroid (fitting two-dimensional Gaussians produces consistent results).  
Pairs of these positions are constructed from channels blue- and red-shifted by 
the same velocity relative to the systemic channel.  For a given velocity pair, 
the angle $\Delta \theta'_{\rm wing}$ is calculated from the line segment 
joining the two centroid positions.  The second panel in Figure \ref{fig:vsys} 
shows $\Delta \theta'_{\rm wing}$ as a function of channel velocity for a few 
values of $\chi$.  Rotation is evident for all channel velocities equal to the
Kepler velocities of material within the cavity.  As was the case for $|\Delta 
\theta'_{\rm sys}|$, the maximum value of $|\Delta \theta'_{\rm wing}|$ 
measured in the 3D model is accurately predicted by the razor-thin model.

\subsection{Disentangling Radial Inflow from Other Non-Keplerian Effects}
\label{subsec:disentangle}

The rotational signatures of inflow identified in \S\ref{subsec:isotwist} and 
\S\ref{subsec:wingrot} can be mimicked to varying degrees by other phenomena.  
We discuss four possible ``contaminants'' and how one might disambiguate 
between these different scenarios.

\subsubsection{Radial outflow} \label{out}

Molecular gas can be entrained in winds flowing up and radially outward from 
the disk surface \citep[e.g.,][]{pontoppidan11,bast11,klaassen13}.  Radial 
outflow would cause the systemic velocity isophotes and the wing emission to 
rotate in the direction opposite to that for radial inflow.  Distinguishing 
inflow from outflow requires that we resolve the spin orientation of the disk; 
i.e., we need to decide whether $i < 90^{\circ}$ or $i > 90^{\circ}$.  The way 
to break this degeneracy is to use spatially resolved channel maps to decide 
which limb is approaching and which limb is receding, and then to combine this 
information with some asymmetry along the projected minor axis.  The asymmetry 
could be manifested in the orientation of a polar jet pointed toward the 
observer (i.e., whether the jet points along the positive $y > 0$ or negative 
$y < 0$ minor axis).  
Alternatively, one could observe a spectral line 
\citep{dutrey98,guilloteau98,rosenfeld13} or continuum 
\citep{augereau99,weinberger99,fukagawa06,thalmann10} brightness asymmetry 
arising from the disk's orientation.  Of course it is 
also possible that one molecular line traces an outflowing disk wind, say 
because it probes gas at high altitude above the midplane, while another traces 
radial inflow, closer to the midplane.

\subsubsection{Infall from a molecular envelope}

Gas from the parent molecular cloud may continue to accrete onto the 
protoplanetary disk at early times \citep{stahler94,yorke99,brinch07}.  
Differential rotation and accelerating infall will mimic the spatio-kinematic 
signatures of inflow \citep[cf.][]{brinch08,tang12}.  However, the spatial 
scale of an infalling envelope should be comparable to or larger than the disk 
itself, and should not be confined to a central cavity.  Furthermore, for an 
isolated source, the duration of infall should be considerably shorter than the
lifetime of the disk \citep[e.g.,][]{pietu05}.  Evidence for local cloud 
material might be indicated by high extinction values, very red infrared 
spectral energy distributions, or contamination (self-absorption and/or 
emission on much larger angular scales) in single dish molecular line spectra.  
Most of the currently known transition disks do not exhibit any clear evidence 
for a remnant envelope structure.

\subsubsection{Vertical disk structure} \label{vert}

Line emission from optically thick gas typically originates in disk surface 
layers located at large vertical heights above the midplane
\citep[e.g.,][]{vanzadelhoff01,dartois03,pavlyuchenkov07,semenov08}.
Therefore, in channel maps, the blue- and red-shifted emission structures are 
not generally collinear with the stellar position; rather, they appear rotated 
away from the midplane in opposite directions from one another
\citep{rosenfeld13,degregorio-monsalvo13}.  These rotations of the blue- and 
red-shifted line wings would effectively cancel out in the procedure we 
developed in \S \ref{subsec:wingrot} to measure $\Delta \theta'_{\rm wing}$.
Moreover, vertical structure does not twist the isophotes at the systemic 
velocity (i.e., $\Delta \theta'_{\rm sys} = 0$).  In sum, vertical structure 
should pose little danger of confusion in efforts to characterize radial 
inflow.

\begin{figure}[t!]
\epsscale{1.0}
\plottwo{ms_radial_092713/f08.eps}{ms_radial_092713/f09.eps}
\caption{Degeneracy between a warp and radial inflow.  The same rotation 
angles for inflow models as shown in Figure \ref{fig:vsys} are plotted here as 
open circles, and overlaid with solid curves generated from model warped 
disks.  Warp parameters were chosen to reproduce $\Delta \theta_{\rm wing}'$ 
for the inflow models; evidently these same warp models automatically mimic 
$\Delta \theta_{\rm sys}'$ as well.  For $\chi=1$, the warp parameters are $R_w 
= R_{\rm cav} = 50$\,AU, $i_{\rm rel}=14^\circ$, and $\Omega_{\rm rel} = 
250^\circ$.  For $\chi = 0.1$, $R_w = 50$\,AU, $i_{\rm rel} = 2^\circ$, and 
$\Omega_{\rm rel}=267^\circ$.  \label{fig:vsys_warp}}
\end{figure}

\subsubsection{A warp}\label{warp}

A warp --- introduced, for example, by a perturbing body whose orbit is 
inclined to the disk plane \citep{larwood97,marzari09} --- also changes the 
morphology of line emission in channel maps.  At its most basic, a warp 
comprises an inner disk that is misaligned with an outer disk, and is described 
by three parameters: the radius $R_w$ where the inner disk transitions to the
outer disk, the inclination $i_{\rm inner}$ of the inner disk relative to the 
sky plane, and the orientation of the inner disk, i.e., the longitude 
$\Omega_{\rm inner}$ of ascending ($\equiv$ toward the observer) node, measured 
in the sky plane and referred to the $x=x'$ axis (as a reminder, the outer disk 
has inclination $i_{\rm outer} \equiv i$ relative to the sky plane and 
orientation $\Omega_{\rm outer} \equiv 0^\circ$).  An equivalent description 
uses the outer disk as the reference plane; under this alternate convention, 
the inner disk inclination relative to the outer disk is $i_{\rm rel}$, and its 
longitude of ascending node is $\Omega_{\rm rel}$, measured in the outer disk 
plane and referred to the $x=x'$ axis.  The angles in the two reference systems 
are related by
\begin{align}
i_{\rm rel}=&\arccos\left( \cos i \cos i_{\rm inner}+
    \cos \Omega_{\rm inner} \sin i \sin i_{\rm inner} \right) \\
\Omega_{\rm rel} =& \frac{\pi}{2} + 
    \arctan\left(\frac{\cos i_{\rm inner} \sin i - 
    \cos i \cos \Omega_{\rm inner} \sin i_{\rm inner} }{\sin i_{\rm inner} 
    \sin \Omega_{\rm inner} } \right) \,.
\end{align}

Figure \ref{fig:vsys_warp} shows $\Delta\theta'_{\rm sys}$ and 
$\Delta\theta'_{\rm wing}$ for two 3D disk models with warps (solid curves) 
overlaid on our previous results for disk models with radial inflow (open 
circles).  The similarity between these rotation angle patterns demonstrates 
that a warp with the right parameters (see caption) can masquerade as inflow.  
Furthermore, increasing $i_{\rm rel}$ can brighten wing emission in the same 
way that increasing $\chi$ does \citep[data not shown; see \S 
\ref{subsec:wingrot};][]{rosenfeld12}.

Distinguishing inflow from a warp requires additional data and modeling.  A 
warped disk located in a region that is known to be optically thin at continuum 
wavelengths would need to have its opacity be reduced by large factors by, 
e.g., grain growth or dust filtration at the cavity rim.  By contrast, in the 
fast inflow model, the opacity need not change much, if at all, since the 
surface density reduction that accompanies fast inflow would account for most 
if not all of the cavity's transparency.  Spatially resolved, multi-wavelength 
imaging of the disk cavity can serve to constrain changes in the dust size
distribution.  And one can utilize emission lines from rarer, optically thin 
gas species (e.g., \citealt{bruderer13}) to measure total gas surface 
densities.  In \S\ref{sec:prospects} we wrestle
again with the warp/inflow degeneracy 
using real observations.\footnote{A warp and inflow are not mutually 
exclusive possibilities --- the two phenomena might be driven simultaneously by 
companions inside the disk cavity.}

\section{A Case Study: HD 142527} \label{sec:prospects}

We consider the remarkable transition disk hosted by the young F star HD 
142527.  At a distance of $d=145 \pm 15$ pc \citep{verhoeff11}, the disk has a 
large dust-depleted cavity of radius $R_{\rm cav} \sim 130$ AU that has been 
imaged in near-infrared scattered light 
\citep{fukagawa06,casassus12,rameau12,canovas13} and mid-infrared thermal 
emission \citep{fujiwara06,verhoeff11}.  The cavity is not completely devoid of 
dust; \citet{verhoeff11} required an optically thick disk (and even an 
optically thin halo) interior to $\sim$30 AU to explain the observed 
near-infrared and mid-infrared emission.  ALMA Band 7 (345\,GHz) observations 
presented by \citet{casassus13} resolved the highly asymmetric dust ring at 
$R_{\rm cav}$ that was originally noted by \citet{ohashi08}.  The ALMA data 
clearly demonstrate that molecular gas resides inside the cavity, emitting in
the $^{12}$CO $J$=3$-$2 and HCO$^+$ $J$=4$-$3 lines.  We have already noted the 
nearly free-fall radial velocities implied by the HCO$^+$ filament 
(\S\ref{intro}).

\begin{figure}[t!]
\epsscale{1.00}
\plotone{ms_radial_092713/f10.eps}
\caption{Evidence for fast radial inflow in the HD 142527 disk.  {\it Left:} 
First moment or intensity-weighted velocity map of the $^{12}$CO $J$=3$-$2 
emission line constructed from archival ALMA data.  {\it Center:} First moment 
map of the HCO$^+$ $J$=4$-$3 emission line.  The HCO$^+$ data has lower 
signal-to-noise than the CO data.  {\it Right:} Velocity field for a razor-thin 
disk with the inflow parameter $\chi$ increasing with decreasing $R$ and 
peaking at unity (see text for details).  All panels are colored using the same 
velocity scale, and the orientation of the outer disk is indicated by dashed 
black lines (projected major axis at PA = $-20^\circ$).  The counter-clockwise 
twist, a signature of radial inflow for $i > 90^\circ$, is present in all 
panels; in the first moment maps, the twist appears on larger scales where the 
emission line intensity is strongest.  \label{fig:hd142527_moms}}
\end{figure}

We apply the tools developed in \S\ref{sec:twist} to the archival ALMA data on 
CO and HCO$^+$ line emission. We aim here only for a first look at the 
kinematics, and so forego an extensive modeling effort, fixing in advance the 
disk mass ($0.1\,M_\odot$), scale radius ($R_s = 50$ AU), disk temperature ($T 
= 70 (R/50 \,{\rm AU})^{-0.7}$), and outer truncation radius (300 AU).  We also 
adopt literature values for the stellar mass \citep[$M = 
2.7\,M_\odot$;][]{casassus13} and viewing geometry for the outer disk 
\citep[$i = 160^\circ$, with the projected major axis oriented at position 
angle PA = -20$^\circ$;\footnote{Alternatively, if our $x'$-axis points west and 
our $y'$-axis points north, then $\Omega_{\rm outer} = 
70^\circ$.}][]{fujiwara06,verhoeff11}. The inclination and PA of the outer disk
are based on CO channel maps together with thermal mid-infrared images which 
show that the eastern side of the disk appears brighter than the western side;
on the eastern side we are seeing the portion of the cavity rim located farther 
from the observer and directly illuminated by starlight.  The line-of-sight 
velocities reported below are all relative to systemic; the systemic velocity 
is +3.64\,km\,s$^{-1}$ relative to the local standard of rest, as we deduced 
from visual inspection of the CO channel maps.

Figure \ref{fig:hd142527_moms} shows the first moment maps for the CO
and HCO$^+$ emission lines.  The first moment or intensity-weighted
line-of-sight velocity appears twisted near the disk center ($R'
\lesssim 0\farcs5$).  With respect to the orientation defined by the
outer disk, the twist in the inner disk is counter-clockwise, which
given $i=160^\circ$ implies radial inflow.  Furthermore, the twist is of order
$\sim$1 radian in magnitude, which implies a radial speed comparable to 
Keplerian ($\chi \sim 1$).  Compare the first moment maps (left 
and center panels) with the velocity field of a model razor-thin disk with a 
fast inflow (right panel; cf.~Figure \ref{fig:thetawing}), and recognize that 
the comparison is only meant to be suggestive since the velocity field 
is not the same as the intensity-weighted velocity.

\begin{figure}[t!]
\epsscale{0.50}
\plotone{ms_radial_092713/f11.eps}
\caption{Possible fast inflow in the HD 142527 disk.  The measured rotation,
$\Delta\theta'_{\rm wing}$, in the $^{12}$CO $J$=3$-$2 line wings can be 
approximately reproduced by either a disk with graded radial
inflow (spatially variable $\chi$; red dashed curve), or a graded warp 
(spatially variable $i_{\rm rel}$ and $\Omega_{\rm rel}$; blue dot-dashed 
curve).  Details of the 
models are in the main text.  An inflow model having constant $\chi = 1$ for $R 
< 35$ AU (orange) does less well than the graded inflow model.  The spin 
orientation of the HD 142527 disk ($i = 160^\circ$) is opposite to those of our 
example disks (having $i < 90^\circ$) in \S\ref{sec:twist}, and so 
$\Delta\theta'_{\rm wing}$ is positive rather than negative.  
\label{fig:hd142527_rot}}
\end{figure}

Figure \ref{fig:hd142527_rot} compares the measured $\Delta\theta'_{\rm wing} 
(v'_{\rm los})$ in the CO line with predictions from several models.  A 
constant $\chi$ model does not fit the data as well as a variable $\chi (R < 
140 \, {\rm AU} = [1+(R/33$\,AU$)^6]^{-1}$ model that introduces the twist
incrementally.  The variable $\chi$ model does as well as a warped disk model 
(cf.~\S\ref{warp}) in which the warp is likewise introduced gradually.  To mock 
up a warp induced by a planet as described by linear secular theory 
\citep[e.g.,][]{dawson11}, we ramp up the relative inclination linearly from 0 
to its final value $i_{\rm rel} = 15^\circ$, and ramp down the nodal angle 
linearly from $\Omega_{\rm rel} + 90^\circ$ to $\Omega_{\rm rel} = 69^\circ$, as 
the disk radius decreases from $2R_w$ to $R_w = 25$ AU.

At the highest velocities/smallest radii probed in Figure 
\ref{fig:hd142527_rot}, all our models systematically underpredict $\Delta 
\theta'_{\rm wing}$.
For the inflow model, the fit at small radii could be improved by having the 
azimuthal velocity decrease by a few tens of percent in tandem with the 
increase in radial inflow velocity.  The large synthesized beam size 
(0\farcs64$\times$0\farcs41, PA = 75.5$^\circ$) precludes us from constructing 
a plot of $\Delta \theta'_{\rm sys} (y')$ showing what would presumably be 
large twists at small radii (cf.~Figure \ref{fig:vsys}, which was made assuming 
a beam size of 0\farcs1).

At the smallest velocities/largest radii, there appears in Figure
\ref{fig:hd142527_rot} to be a constant offset in $\Delta \theta'_{\rm
  wing}$ of about 6 degrees. The offset may be due to a systematic
error in the position angle we assumed for the outer disk.  Another
source of error at these low velocities may be in our systemic
velocity (see \citealt{hughes11} for a method to quantify this
uncertainty).

To summarize, our first-cut analysis of the \citet{casassus13} ALMA 
observations of the HD 142527 disk demonstrates that the predicted 
spatio-kinematic signatures of fast radial inflow are eminently detectable, and 
merits a dedicated pursuit of inflow in a general sample of transition disks.  
Additional observations and modeling are posited to distinguish between a fast 
inflow or warp origin for the observed kinematics.  This particular disk is 
especially challenging to interpret because of its complex and non-axisymmetric 
structures \citep[][]{fukagawa06,fujiwara06,verhoeff11,casassus12,honda12,
rameau12,casassus13,canovas13}.  The disk interior to $R\sim 30$ AU that was 
inferred by \citet{verhoeff11} could, in principle, support a warp.  What would 
help clarify the situation would be observations of the central cavity in 
an optically thin line, to confirm the lower gas surface densities that should 
accompany fast inflow.

\section{Conclusions}\label{sec:conclude}

Radial inflow of gas at velocities approaching free fall can account for both 
the depleted densities inside transition disk cavities and the relatively 
normal stellar accretion rates they maintain.  We have demonstrated that high 
spatial and spectral resolution observations of molecular emission lines with
(sub)millimeter wavelength interferometers --- particularly ALMA --- can detect 
inflow by virtue of its two key signatures: twisted isophotes at the systemic 
velocity, and rotated emission patterns in channel maps made in the line wings.
We developed quantitative diagnostics of channel maps to characterize radial 
inflow, and discussed various real-world complications.  Of these, the most 
serious is a warp.

We tested our inflow diagnostics on archival ALMA observations of the HD 142527
transition disk, uncovering a booming signal consistent with free-fall inflow 
velocities at distances of $\sim$25--50 AU from the host star.  Nevertheless a 
warped disk at these radii can also reproduce the large, order-unity rotations 
seen in the CO and HCO$^+$ line wings.  Whether the HD 142527 disk cavity 
contains a fast radial inflow or a warped disk (or both), the only mechanism on 
the table now that can explain either phenomenon involves strong gravitational
torques, exerted by one or more giant planets/brown dwarfs/low-mass stars 
\citep[as yet undetected; see][]{biller12,casassus13}.  Fast flows should also 
produce strong shocks in disks.  We have not investigated the signatures of 
such shocks; the observables depend on details such as the density
and the cooling mechanisms of shocked material.
In the central cavity of the circumbinary disk surrounding GG Tau A,
\cite{beck12} have detected hot H$_2$
that may trace shocks internal to the accretion flow.

\subsection{Chapter Acknowledgements}
We thank Eve Lee for discussions, and the organizers of the 2013 
Gordon Conference on Origins of Solar Systems, Ed Young and Fred Ciesla, for 
providing a great venue for sparking collaboration on this article.  This paper 
makes use of the following ALMA data: ADS/JAO.ALMA\#2011.0.00465.S.  ALMA is a 
partnership of ESO (representing its member states), NSF (USA) and NINS 
(Japan), together with NRC (Canada) and NSC and ASIAA (Taiwan), in cooperation 
with the Republic of Chile.  The Joint ALMA Observatory is operated by ESO, 
AUI/NRAO and NAOJ.  The National Radio Astronomy Observatory is a facility of 
the National Science Foundation operated under cooperative agreement by 
Associated Universities, Inc.

\clearpage

%\bibliography{shortbib}

### 2015-07-14 V4046Sgr2013/ms_v4046sgr2013_chapter.tex
more changes

@@
 \citep{hutchinson90,weaver92,jensen96,jensen97,rodriguez10,oberg11}, the 2MASS 
-\citep{skrutskie06} and {\it IRAS} \citep{beichmann88} point source catalogs, 
+\citep{skrutskie06} and {\it IRAS} \citep{beichman88} point source catalogs, 
 and an archival {\it Spitzer} IRS spectrum.  The SED features the standard 
@@
 natural explanation in the growth and inward radial drift of dust particles 
-\citep{weidenschilling80,brauer08a,birnstiel10}.  Drift is a fundamental feature
+\citep{weidenschilling80,brauer08,birnstiel10}.  Drift is a fundamental feature
 in the migration of disk solids, generated because the radial pressure gradient
@@
 drag-force on the dust that relates to the size of the particle, preferentially
-moving the large particles towards the central star \citep{weidenschilling77,
+moving the large particles towards the central star \citep{weidenschilling77b,
 takeuchi02}.  As these particles drift inwards, their size distribution is 

### 2015-07-14 ms_radial_092713/ms_radial_chapter.tex
more changes

@@
 \citep[e.g.,][]{Kim13} or spatially resolved imaging at infrared 
-\citep[e.g.,][]{Geers2007} and mm wavelengths \citep[e.g.,][]{Brown09}.  The
+\citep[e.g.,][]{geers2007} and mm wavelengths \citep[e.g.,][]{brown09}.  The
 cavities in these ``transition'' disks have radii ranging from $\sim$4\,AU 
-\citep[TW Hya;][]{Calvet02,Hughes07} to 20\,AU \citep[GM Aur;][]{hughes09} to 
-40--50\,AU \citep[LkCa 15;][]{Andrews11,Isella12} to 140\,AU \citep[HD 
+\citep[TW Hya;][]{calvet02,hughes07} to 20\,AU \citep[GM Aur;][]{hughes09} to 
+40--50\,AU \citep[LkCa 15;][]{andrews11b,isella12} to 140\,AU \citep[HD 
 142527;][]{casassus13}.  The suspicion is that dynamical interactions with 
@@
 $\sim$10 $M_{\rm Jup}$, have been posited to carve out a given cavity 
-\citep{Zhu11,Dodson11}.\footnote{Even multiple super-Jovian planets are 
+\citep{Zhu11,dodson11}.\footnote{Even multiple super-Jovian planets are 
 reportedly not enough: grain growth and dust filtration at the cavity rim are 
 invoked in addition to planets to make the cavity sufficiently transparent 
-\citep{Zhu11,Dong12}.}  At least in the case of the transition disk around LkCa 
+\citep{Zhu11,dong12}.}  At least in the case of the transition disk around LkCa 
 15, there might be one (but so far only one) actively accreting super-Jovian 
@@
 non-axisymmetric outer disk, gas streamlines can cross, shock, and plunge 
-inward \citep[e.g.,][]{pac77}.
+inward \citep[e.g.,][]{paczynski77}.
 
@@
 Still, some indirect empirical support for the scalings in equation 
-(\ref{eqn:simple}) has been found by \cite{Fung13}.
+(\ref{eqn:simple}) has been found by \cite{fung13}.
 }
@@
 $H=0.1R$).  The surface density profile is that of a thin, viscously accreting 
-disk \citep{bellpringle74}: $\Sigma\propto R^{-1}\exp(-R/R_s)$ with scale 
+disk \citep{lynden-bell74}: $\Sigma\propto R^{-1}\exp(-R/R_s)$ with scale 
 radius $R_s=50$ AU and an outer truncation radius of 300 AU.  Since our focus 

### 2015-08-03 chapters/conclusion.tex
added intro and conclusion outlines

\chapter{Conclusions} \label{chapter:conclusion}

Summary of previous chapters here.

\section{Recent Developments}

Go through some recent/concurrent papers that may have built on these ideas.

Resolving magnetic fields: \citep{stephens14}

HL Tau: \citep{brogan14}

A gap in Tw Hya at 20 AU? \citep{akiyama15}

Gas density within cavities of transition disks \citep{vandermarel15}

More dust traps: \citep{espaillat15}

\section{A Bright Future}
Circumplanetary disks? \citep{perez15}
Chemistry and stuff \citep{oberg15} \citep{graninger15} 
Chemistry as a signpost of planet formation \citep{cleeves15}
Exploring the outer solar system with New Horizons
Detection of organics by Rosetta? Checkout: \url{http://www.sciencedaily.com/releases/2015/07/150730172518.htm}

Summarize status of EHT and its future.

### 2015-08-03 chapters/introduction.tex
added intro and conclusion outlines

\chapter{Introduction}\label{chapter:introduction}

\section{Structure of Protoplanetary Disks}

Detailed studies versus population statistics

\citep{harris14}

\section{The Event Horizon Telescope}

\section{Basics of Radio Interferometry}
\subsection{Visibilities}
\subsection{Channel Maps}
\subsection{VLBI}

\section{Thesis Outline}

### 2015-08-03 chapters/conclusion.tex
chapter 2 summary

@@
-\chapter{Conclusions} \label{chapter:conclusion}
+\chapter{Conclusions} \label{chap:conclusion}
 
-Summary of previous chapters here.
+One might argue that the space of discovery in Astronomy is spanned by resolution and 
+wavelength.  Technology has enabled us to contemplate phenomena at wavelengths far outside 
+what the realm of our human senses, revolutionizing fields like high energy astrophysics,
+cosmology, and the study of the inter stellar-medium.  Resolution then reveals the hidden
+details and helps us to interpret what these new eyes see.  A recent example is 
+the application of interferometry at sub-mm wavelengths with ground breaking telescopes like the SMA, 
+the amazing new power of ALMA, and the EHT revolution.  
+While the science motivating chapters \ref{chap:v4046_structure} 
+through \ref{chap:radial_flows} may be different than in the final chapter of this thesis, 
+I am left wondering and waiting for what will we see when we look only a little bit closer.
 
-\section{Recent Developments}
-
-Go through some recent/concurrent papers that may have built on these ideas.
+Emboldened by this spirit, we shall conclude not only with summaries of the work presented here,
+but also some brief discussion of recent developments in the field.  In chapter \ref{chap:v4046_structure} we 
+used the SMA to study the protoplanetary disk 
+around the close binary V4046 Sgr.  This disk featured a narrow ring of large mm-sized grains, 
+but an extended gas rich disk.  Detailed modeling of these features demonstrated that they were 
+phenomenologically consistent with the viscous spreading of the gas and the concentration through pressure 
+drag forces of the dust.  This disk is a single example among a number, and more such examples continue to be 
+found, \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A somewhat related phenomenon are the 
+severely asymmetric dust  ``horse-shoes'' that have been recently observed at multiple wavelengths
+\citep{vandermarel13,fukagawa13,isella13,casassus15,marino15}.  Thought to be perhaps generated by gigantic 
+3D Rossby wave instabilities \citep{regaly13}, these are another striking example from the interplay between
+the gas and dust in protoplanetary disks.
 
@@
 \section{A Bright Future}
+What is happening in the inner disks? \citep{ke14}
 Circumplanetary disks? \citep{perez15}
@@
 Summarize status of EHT and its future.
+
+This thesis is a small expression of 
+the science that resolution at these wavelengths can provide and only an exploration for the 
+hard and promising work ahead.

### 2015-08-05 chapters/conclusion.tex
finished rough conclusion draft

@@
 drag forces of the dust.  This disk is a single example among a number, and more such examples continue to be 
-found, \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A somewhat related phenomenon are the 
+found \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14} although it is not universal 
+\citep{huelamo15}.  A somewhat related phenomenon are the 
 severely asymmetric dust  ``horse-shoes'' that have been recently observed at multiple wavelengths
@@
 3D Rossby wave instabilities \citep{regaly13}, these are another striking example from the interplay between
-the gas and dust in protoplanetary disks.
+the gas and dust in protoplanetary disks.  Another exciting example are the narrow gaps in the dust continuum
+of HL Tau. This striking image was obtained using ALMA baselines as long as 15\,km and possible causes include
+planet-disk interactions and condensation fronts \citep{brogan14,zhang15}.
 
-Resolving magnetic fields: \citep{stephens14}
+In chapter \ref{chap:hd163296} we explored the vertical structure of a protoplanetary disk using 
+observations that showed clear, resolved signs of the vertical temperature gradient in the disk.
+The search for azimuthal temperature variations are a natural extension \citep{isella13} and 
+some evidence found \citep{vanderplas14}.  There is much activity to continue designing
+observations that can probe the ambient conditions 
+for planet formation beyond temperature from the ionization environment \citep{cleeves14,teague15} to 
+turbulence \citep{hughes11,guilloteau12,flock15,simon15}.
 
-HL Tau: \citep{brogan14}
+To conclude our work on protoplanary disks, we discussed in \ref{chap:radial_flows} observational indicators of 
+radial gas flows within the holes of transition disks.  \citet{casassus15} have studied the gas kinematics for
+the curious case of HD142527 using the CO $J = 6-5$ line and find that their data is consistent with
+accretion reaching close to free-fall velocities.  The gas surface density of transition disks have also 
+been recently studied using observations taking at sub-millimeter wavelengths 
+\citep{bruderer14,zhang14,canovas15,perez15,vandermarel15} and will help determine how these structures are formed
+\citep{bruderer13}.  Molecular disk winds, outflows, and binary systems are all recently observed 
+with interesting gas kinematics \citep{klaasen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
+future, well resolved, highly sensitive observations of spectral lines may constrain the strength of MRI-drive
+turbulence \citep{simon15}, query the active sites of planet formation \citep{kleeves15,ober15}, or reveal
+circumplanetary disks \citep{perez15}.  The morphology of spectral line emission can provide be a powerful 
+tool for studying astrophysics beyond kinematics.
 
-A gap in Tw Hya at 20 AU? \citep{akiyama15}
-
-Gas density within cavities of transition disks \citep{vandermarel15}
-
-More dust traps: \citep{espaillat15}
-
-\section{A Bright Future}
-What is happening in the inner disks? \citep{ke14}
-Circumplanetary disks? \citep{perez15}
-Chemistry and stuff \citep{oberg15} \citep{graninger15} 
-Chemistry as a signpost of planet formation \citep{cleeves15}
-Exploring the outer solar system with New Horizons
-Detection of organics by Rosetta? Checkout: \url{http://www.sciencedaily.com/releases/2015/07/150730172518.htm}
-
-Summarize status of EHT and its future.
+The Event Horizon Telescope continues to break ground and reports exciting new science with every
+campaign from the early detections of AGN at mm wavelengths \citep{doeleman05} to more recently structures
+at the event horizon of a supermassive black hole \citep{doeleman08,doeleman09,doeleman13}.
+The 2015 campaign represented a new scale of operation, incorporating new telescopes and
+technologies.  In chapter \ref{chap:aphids} we reported on the status of the \APHIDS\, software which
+will post-process the phased-SMA data and make the sampling rate compatible with the rest of the EHT
+network.  This problem is not limited to the SMA and we hope that the development done with \APHIDS\, will
+prove useful for other VLBI experiments.
 

### 2015-08-09 chapters/introduction.tex
finished rough conclusion draft
worked on structure section

@@
 
-Detailed studies versus population statistics
+The canonical picture of star formation breaks this process into several stages.  Picture first a large 
+molecular cloud within which there are complicated sub structures: clumps and cores of 
+concentrated gas and dust.  Over time these self-gravitating units will contract,  the material collapses into 
+a disk to conserve angular momentum, and nascent stars ignite their thermonuclear engines.  Jets, envelopes, 
+winds, and outflows punctuate this 
+process which may likely be disturbed by the swift pull of a passing star or the constant presense of a nearby 
+neighbor.  Whole fields, techniques, and technologies have been developed to study single chapters of this 
+complicated story and ours begins near the very end when only the disk and young star remain.
 
-\citep{harris14}
+Over time, this disk of gas and dust will dissipate leaving trace debris: astroids and comets, dust and planets.
+So while we may be at the end of the star's formation, this protoplanetary disk is the stage where planets are
+born.  Their composition and structure provides the crucial initial conditions that must enable a diverse 
+population of planets.  However, theories of planet formation are still developing and require the consideration 
+of physical processes acting on many physical scales and over a huge range of times.  As such, there 
+work to test and develop these ideas is very diverse, ranging from experiments on grain interactions in 
+microgravity to space missions exploring the larger rocks that remain as testament to the formation of our 
+own Solar system.
+
+Over the past several decades, observations from interferometers operating at millimeter wavelengths have made
+significant contributions to understanding the structure of these disks.  The early maps of CO made using 3 to 5 
+telescopes in an array at 
+Owens Valley revealed bound, rotating disks with diameters 
+similar to the dimensions of own solar system 
+\citep{sargent87,weintraub87,weintraub89,sargent91,koerner93a,koerner93b,koerner95}.  The emission pattern
+matching Keplerian 
+rotation was then used to measure some systems' dynamical masses 
+\citep{dutrey98,guilloteau98,guilloteau99,simon00} and
+maps of multiple species to constrain the two-dimensional structure 
+\citep{aikawa99,aikawa01,vanzadelhoff01,aikawa02}.   Soon arrays detected the giant central dust holes of 
+``transition disks''
+\citep{hughes07,brown09,andrews11}.
+
+There exists several techniques for constraining the structure of protoplanetary disks.  Some geometric 
+properties such as size, inclination, and orientation can be calculated from an image.  The amount of dust 
+in a disk can also be estimated by assuming a dust opacity and measuring the mm-wave flux.  Alternatively, one 
+may try to solve the inverse problem: assume some disk model, calculate the resulting emission, and then compare
+it to what is observed.  Often the density, temperature, and velocity profiles of the disks are assumed to 
+follow power laws or some other parametric function.  There also exist some grids and models that try to 
+calculate abundances in a self-consistent manner.  Additionally, researchers usually either try to constrain 
+populations using 
+surveys \citep[i.e.][]{andrews11,harris14} or present detailed studies of single, interesting disks.  The work
+presented in Chapter \ref{chap:blagh} to \ref{chap:blagh} takes the latter approach and uses simple parametric 
+models to investigate the structure of a choice few, but very interesting, protoplanetary disks.
 

### 2015-08-10 chapters/introduction.tex
added EHT section

@@
 
-\section{The Event Horizon Telescope}
+V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
+protoplanetary disk.  In chapter \ref{chap:V4046Sgr} we present Submillimimeter Array (SMA) observations of this 
+disk 
+that resolve for the
+first time a large central cavity.  In addition, we show that the distribution of gas and dust are non-commitant 
+with each other: the gas disk is large and extended while the mm-sized dust grains are concentrated in a narrow 
+ring.  Furthermore, we show that dust within ring is not completely cleared.  We argue that this configuration is 
+the hallmark of old disk sculpted by at least one large, embedded planet.  HD163296 also has a very large, bright gaseous disk, but no resolved cavity.  In chapter \ref{chap:hd163296} we 
+use 
+science verification data from the Atacama Large Millimimeter Array (ALMA) to study the vertical structure of 
+this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
+its distance from mid-plane and explore additional, nuanced aspects of the disk structure and how they may 
+affect the observables.   In chapter \ref{chap:radial_flows} we consider what impact radial motions of gas within transition disk 
+holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
+with possible sources of confusion such as warps in the disk.
 
@@
 
-\section{Thesis Outline}
+\section{Sample Rate Conversion for the EHT}
+
+The Event Horizon Telescope (EHT) uses VLBI to observe supermassive black holes, obtaining resolution high enough
+to resolve event horizon scales of nearby targets.  These observations provide a 
+unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
+and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
+center and the giant elliptical galaxy M87 \citep{doeleman08,doeleman09}.  These data have provided indirect 
+evidence for existence of an event horizon \citep{broderick15} along with constraints on the degree of 
+magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
+observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
+well as new studies of black hole time variability \citep{doeleman09} and black hole accretion \citep{chan15}.
+
+The bright future of the EHT rests on its key ability to expand by adding more telescopes and
+incorporating new technologies into its operations.  The March 2015 observing campaign was a major milestone in
+the EHT expansion with participation from the LMT in Mexico, CARMA in California, the SMA and JCMT in Hawaii, 
+SMT in Arizona and 30m at Pico Valeta in Spain.  However, the data taken at the SMA must be processed before it
+can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the \APHIDS\, pipeline
+which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
+is intended to be a long-term solution to enable VLBI with the SMA and has applications for other sites such 
+as ALMA that operate at a sampling rate that is not compatible with the standard VLBI data product.

### 2015-08-13 chapters/introduction.tex
expanded intro

@@
 \chapter{Introduction}\label{chapter:introduction}
+
+INTRO PARAGRAPH HERE
+
+The data presented in this thesis was obtained using radio interferometers.  This technique combines the signals
+from pairs of radio telescopes to produce data that is sensitive to an angular scale inversely 
+proportional to the distance between the two elements.  By assembling an array with multiple 
+telescopes, an observer can accumulate information about many spatial scales and reconstruct an image.  However,
+the information contained in this image will be limited to the scales originally measured by the interferometer.
+Images with the best fidelity come from arrays that have elements at a wide variety of 
+distances to each other and the highest resolution comes from the most widely separated telescopes.
+
+figure: images of SMA and ALMA
+
+Each pair of antennas in an array forms a baseline vector (measured in wavelengths) with length $D_\lambda$ and 
+components $(u,v,w)$.  The signal voltages, $V_1(t)$ and $V_2(t)$, measured by the individual 
+elements are correlated against each other
+\begin{equation}
+r(\tau) = \lim_{\tau \to \infto} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
+\end{equation}
+This operation is often done on site with data streamed directly to a correlator and the resulting 
+measurement, or visibility, is stored on disk.  However, for Very Long Baseline Interferometry (VLBI), the 
+radio telescopes can be separated by entire continents and real-time correlation is not feasible.  For these
+experiments, the signal voltages must be recorded and the visibilities calculated at some later time.  The Event
+Horizon Telescope (EHT) a leader in the developing field of (sub)-millimeter VLBI.  Observations at these 
+short wavelengths becomes more difficult as sources
+tend to become fainter and the atmosphere becomes noiser, time variable, and more opaque.  Consequently, it 
+has only been in the past few decades that facilities have been built to operate at these wavelengths.
+
+The Submillimeter Array \citep[SMA][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is located on the top of 
+Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
+It is composed of eight 6-meter diameter dishes which, like many 
+interferometers, can be moved into different configurations so that there are elements as close as 
+15 meters or as far as 500 meters.  Its recievers can operate at wavelengths from 430\,$\mu$m to 2\,mm and 
+historically have provided up to 4\,GHz of bandwidth.  A major upgrade to the array is increasing the bandwidth
+to 8\,GHz, enabling more sensitive observations and larger spectral coverage.  To participate in VLBI 
+experiments, the SMA can operate as a ``phased array'', essentially turning the interferometer into a single 
+dish telescope.
+
+Located 4450\,km away on the Chile is the Atacama Large Millimeter Array (ALMA).  This facility began early
+science operations in 2011 and has been commissioned in a staged fashion where each observing cycle offers more 
+capabilities.  Once completed, ALMA will host an impressive 66 antennas, 64 of will be 12\, dishes and
+the remaining smaller 7\,m ones.  A fully extended configuration will include baselines up to 16\,km and the 
+array will observe at wavelengths from 400\,$\mu$m to 7\,mm.  The ALMA Phasing Project (APP) will enable this
+giant array to join VLBI experiments and fringes to APEX were found in a January 2015 test \citep{matthews15}.
+
+% https://almascience.nrao.edu/documents-and-tools/alma-technical-notes-a-subset-of-documents-from-the-eoc-memo-series
 
@@
 
+figure: early images of disks
+
 Over the past several decades, observations from interferometers operating at millimeter wavelengths have made
@@
 \citep{hughes07,brown09,andrews11}.
+
+HERE GOES A PARAGRAPH DESCRIBING SIMPLE CONCEPTION OF DISK STRUCUTRE
+figure: disk structure cartoon
 
@@
 ring.  Furthermore, we show that dust within ring is not completely cleared.  We argue that this configuration is 
-the hallmark of old disk sculpted by at least one large, embedded planet.  HD163296 also has a very large, bright gaseous disk, but no resolved cavity.  In chapter \ref{chap:hd163296} we 
+the hallmark of an old disk that is being sculpted by at least one embedded planet.  HD163296 also has a 
+very large, bright gaseous disk, but no resolved cavity.  In chapter \ref{chap:hd163296} we 
 use 
@@
 
-\section{Basics of Radio Interferometry}
-\subsection{Visibilities}
-\subsection{Channel Maps}
-\subsection{VLBI}
-
 \section{Sample Rate Conversion for the EHT}
 
-The Event Horizon Telescope (EHT) uses VLBI to observe supermassive black holes, obtaining resolution high enough
+figure: map of EHT site locations
+
+The Event Horizon Telescope (EHT) observes supermassive black holes, obtaining resolution high enough
 to resolve event horizon scales of nearby targets.  These observations provide a 

### 2015-08-14 chapters/conclusion.tex
draft of conclusion and intro done

@@
-\chapter{Conclusions} \label{chap:conclusion}
+\chapter{Conclusion} \label{chap:conclusion}
 
@@
 The 2015 campaign represented a new scale of operation, incorporating new telescopes and
-technologies.  In chapter \ref{chap:aphids} we reported on the status of the \APHIDS\, software which
+technologies.  In chapter \ref{chap:aphids} we reported on the status of the APHIDS software which
 will post-process the phased-SMA data and make the sampling rate compatible with the rest of the EHT
-network.  This problem is not limited to the SMA and we hope that the development done with \APHIDS\, will
+network.  This problem is not limited to the SMA and we hope that the development done with APHIDS will
 prove useful for other VLBI experiments.

### 2015-08-14 chapters/introduction.tex
draft of conclusion and intro done

@@
 
-INTRO PARAGRAPH HERE
+Thomas Young's double split experiment is often a student's first formal introduction to the ideas of 
+coherence and interference.  By observing the famous fringe pattern, Young could study the wave-like nature of 
+photons and aspects of our physical world that operate on quantum scales.  Today, the techniques of interferometry
+are ubiquitious in technical fields and their applications diverse.  To astronomers, interferometers automatically
+imply high resolution and the cabability to discern, even see, small details.  Interferometers can measure the 
+radii of stars, search for extrasolar planets, and watch stars explode. This thesis uses data from radio
+interferometers in the study of disks around young stars and massive black holes, pursuing methods and goals
+that can help resolve interesting questions in the field.
 
-The data presented in this thesis was obtained using radio interferometers.  This technique combines the signals
-from pairs of radio telescopes to produce data that is sensitive to an angular scale inversely 
+Radio interferometry combines the signals
+from pairs of telescopes to produce data that is sensitive to an angular scale inversely 
 proportional to the distance between the two elements.  By assembling an array with multiple 
@@
 \begin{equation}
-r(\tau) = \lim_{\tau \to \infto} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
+r(\tau) = \lim_{\tau \to \infty} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
 \end{equation}
@@
 similar to the dimensions of own solar system 
-\citep{sargent87,weintraub87,weintraub89,sargent91,koerner93a,koerner93b,koerner95}.  The emission pattern
+\citep{sargent87,weintraub87,weintraub89,sargent91,koerner93,koerner93b,koerner95}.  The emission pattern
 matching Keplerian 
@@
 
-HERE GOES A PARAGRAPH DESCRIBING SIMPLE CONCEPTION OF DISK STRUCUTRE
 figure: disk structure cartoon
+
+The main two ingredients in the disk are gas and dust.  The majority of the disk mass will be in molecular 
+hydrogen, but observations focus on easier detect trace molecules such as $^{12}$CO and its isotopologues.
+The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
+can give a reliable measure of mass (CITE).  Interferometric observations taken at millimeter wavelengths
+can resolve structure of nearby protoplanetary disks on scales of a few tens to hundreds of AU.  The motion of 
+material in the disk can also be studied using these same telescopes by studying how the pattern of line emission
+from molecules such as CO changes with frequency.  The bulk motion of the gas relative towards the observer 
+changes as a function of position in the disk. To an observer scanning over frequency, different parts of the 
+disk seem to light up when its Doppler shifted emission matches the observer's frequency window.
 
@@
 populations using 
-surveys \citep[i.e.][]{andrews11,harris14} or present detailed studies of single, interesting disks.  The work
-presented in Chapter \ref{chap:blagh} to \ref{chap:blagh} takes the latter approach and uses simple parametric 
+surveys \citep[i.e.][]{andrews11,oberg11,harris12} or present detailed studies of single, 
+interesting disks.  The work
+presented in chapters \ref{chap:v4046sgr_structure} to \ref{chap:radial_flows} takes the latter approach and uses simple parametric 
 models to investigate the structure of a choice few, but very interesting, protoplanetary disks.
@@
 V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
-protoplanetary disk.  In chapter \ref{chap:V4046Sgr} we present Submillimimeter Array (SMA) observations of this 
+protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present Submillimeter Array (SMA) observations of this 
 disk 
@@
 the hallmark of an old disk that is being sculpted by at least one embedded planet.  HD163296 also has a 
-very large, bright gaseous disk, but no resolved cavity.  In chapter \ref{chap:hd163296} we 
+very large, bright gaseous disk, but no resolved cavity.
+
+In chapter \ref{chap:hd163296} we 
 use 
@@
 its distance from mid-plane and explore additional, nuanced aspects of the disk structure and how they may 
-affect the observables.   In chapter \ref{chap:radial_flows} we consider what impact radial motions of gas within transition disk 
+affect the observables.
+
+In chapter \ref{chap:radial_flows} we consider what impact radial motions of gas within transition disk 
 holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
@@
 SMT in Arizona and 30m at Pico Valeta in Spain.  However, the data taken at the SMA must be processed before it
-can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the \APHIDS\, pipeline
+can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the APHIDS pipeline
 which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software

### 2015-08-23 V4046Sgr2013/ms_v4046sgr2013_chapter.tex
chapter 2 summary
fixed figure placement

@@
-\chapter{The Structure of the Evolved Circumbinary Disk around V4046 Sgr}{\label{c:v4046_structure}
-
-\section*{abstract}
-We present sensitive, sub-arcsecond resolution Submillimeter Array observations 
+\chapter{The Structure of the Evolved Circumbinary Disk around V4046 Sgr}{\label{chap:v4046sgr_structure}
+
+%%\section*{abstract}
+In this chapter we present sensitive, sub-arcsecond resolution Submillimeter Array observations 
 of the protoplanetary disk around the nearby, pre-main sequence spectroscopic 
@@
 \epsscale{0.95}
-\plotone{V4046Sgr2013/f01.eps} \caption{An observational summary of the V4046
+\plotone{V4046Sgr2013/f01.eps} 
+\caption{An observational summary of the V4046
 Sgr disk.  (a) An image of the 1.3\,mm dust continuum emission, with contours

### 2015-08-23 hd163296/ms_hd163296_chapter.tex
finished rough conclusion draft
fixed figure placement

@@
-\chapter{A Spatially Resolved Vertical Temperature Gradient in the HD 163296 Disk}\label{c:hd163296}
-
-\section*{abstract}
-We analyze sensitive, sub-arcsecond resolution ALMA Science Verification 
+\chapter{A Spatially Resolved Vertical Temperature Gradient in the HD 163296 Disk}\label{chap:hd163296}
+
+%%\section*{abstract}
+In this chapter we analyze sensitive, sub-arcsecond resolution ALMA Science Verification 
 observations of CO emission lines in the protoplanetary disk hosted by the 

### 2015-08-27 chapters/introduction.tex
reworked intro

@@
 photons and aspects of our physical world that operate on quantum scales.  Today, the techniques of interferometry
-are ubiquitious in technical fields and their applications diverse.  To astronomers, interferometers automatically
-imply high resolution and the cabability to discern, even see, small details.  Interferometers can measure the 
-radii of stars, search for extrasolar planets, and watch stars explode. This thesis uses data from radio
-interferometers in the study of disks around young stars and massive black holes, pursuing methods and goals
-that can help resolve interesting questions in the field.
+are ubiquitious over a wide array of technical fields.  Even within astronomy the scientific applications are 
+diverse: interferometers  search for extrasolar planets, find subtle ripples in the early universe and watch 
+stars explode.  But in all of these examples the astronomer is searching for small details in our sky that, 
+without the resolving power of interferometry, might remain hidden.  Going along with this theme, this thesis 
+presents four studies using observations at millimeter wavelenghts that work to reveal fine structures in disks 
+around young stars and massive black holes.  We cover a wide variety of techniques, from enabling the production 
+of useful data products to the basic interpretation of never before seen features.
 
-Radio interferometry combines the signals
-from pairs of telescopes to produce data that is sensitive to an angular scale inversely 
-proportional to the distance between the two elements.  By assembling an array with multiple 
+Millimeter interferometry is intimately tied to the earlier development of techniques and technologies 
+used at longer wavelengths.  The essential operation combines the signals from pairs of telescopes to produce a 
+measure that is sensitive to some angular scale on the sky.  This scale shrinks as the distance
+between the two elements grows.  By assembling an array with multiple 
 telescopes, an observer can accumulate information about many spatial scales and reconstruct an image.  However,
@@
 
-figure: images of SMA and ALMA
+\begin{figure}[t!]
+\plottwo{chapters/sma_small.jpg}{chapters/ALMA_NRAO_padilla_14.jpg}
+\caption{{\it Left:} The SMA in its compact configuration with an antenna transporter on the right hand side 
+({\it photo credit:} J. Weintroub).  {\it Right:} Part of the ALMA array along with the high site control room and correlator building ({\it photo credit}: ALMA (ESO/NAOJ/NRAO), C. Padilla).}
+\label{fig:sma_alma}
+\end{figure}
 
@@
 components $(u,v,w)$.  The signal voltages, $V_1(t)$ and $V_2(t)$, measured by the individual 
-elements are correlated against each other
+elements are correlated against each other:
 \begin{equation}
@@
 measurement, or visibility, is stored on disk.  However, for Very Long Baseline Interferometry (VLBI), the 
-radio telescopes can be separated by entire continents and real-time correlation is not feasible.  For these
-experiments, the signal voltages must be recorded and the visibilities calculated at some later time.  The Event
-Horizon Telescope (EHT) a leader in the developing field of (sub)-millimeter VLBI.  Observations at these 
-short wavelengths becomes more difficult as sources
-tend to become fainter and the atmosphere becomes noiser, time variable, and more opaque.  Consequently, it 
-has only been in the past few decades that facilities have been built to operate at these wavelengths.
+radio telescopes can be separated by entire continents and real-time correlation is not usually feasible.  For 
+these experiments, the signal voltages must be recorded and the visibilities calculated at some later time.  The 
+Event Horizon Telescope (EHT) is a leader in the developing field of millimeter VLBI and relies on existing 
+observatories.  The difficulty of observing at these short wavelengths increases as the atmosphere becomes 
+noiser, time variable, and more opaque.  Consequently, facilities that can operate at these wavelengths are a 
+precious resource.
 
-The Submillimeter Array \citep[SMA][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is located on the top of 
-Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
-It is composed of eight 6-meter diameter dishes which, like many 
-interferometers, can be moved into different configurations so that there are elements as close as 
-15 meters or as far as 500 meters.  Its recievers can operate at wavelengths from 430\,$\mu$m to 2\,mm and 
-historically have provided up to 4\,GHz of bandwidth.  A major upgrade to the array is increasing the bandwidth
-to 8\,GHz, enabling more sensitive observations and larger spectral coverage.  To participate in VLBI 
-experiments, the SMA can operate as a ``phased array'', essentially turning the interferometer into a single 
-dish telescope.
+The Submillimeter Array \citep[SMA][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is 
+located on the top of Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
+It is composed of eight 6 meter diameter dishes which, like many interferometers, can be moved into different 
+configurations so that there are elements as close as 15 meters or as far as 500 meters (see Figure 
+\ref{fig:sma_alma}).  Its recievers can operate at wavelengths from 430\,microns to 2\,millimeters and 
+historically have provided up to 4\,GHz of bandwidth.  Chapter \ref{chap:v4046sgr_structure} features an 
+exquisite dataset from the SMA that combines observations taken by multiple configurations.  A major upgrade to 
+the array is increasing the bandwidth to 8\,GHz, enabling more sensitive observations and larger spectral 
+coverage.  To participate in VLBI experiments, the SMA can operate as a ``phased array'', essentially turning 
+the interferometer into a single dish telescope.  Chapter \ref{chap:aphids} addresses a key step of how to 
+integrate the SMA into millimeter VLBI observations such as those taken by the EHT.
 
-Located 4450\,km away on the Chile is the Atacama Large Millimeter Array (ALMA).  This facility began early
+Located 4450\,km away in Chile is the Atacama Large Millimeter Array (ALMA).  This facility began early
 science operations in 2011 and has been commissioned in a staged fashion where each observing cycle offers more 
-capabilities.  Once completed, ALMA will host an impressive 66 antennas, 64 of will be 12\, dishes and
-the remaining smaller 7\,m ones.  A fully extended configuration will include baselines up to 16\,km and the 
-array will observe at wavelengths from 400\,$\mu$m to 7\,mm.  The ALMA Phasing Project (APP) will enable this
-giant array to join VLBI experiments and fringes to APEX were found in a January 2015 test \citep{matthews15}.
-
-% https://almascience.nrao.edu/documents-and-tools/alma-technical-notes-a-subset-of-documents-from-the-eoc-memo-series
+capabilities.  Once completed, ALMA will host an impressive 66 antennas, 64 of will be 12\,meter dishes and
+the remaining smaller 7\,meter ones (see Figure \ref{fig:sma_alma}).  We used publickly available data from 
+this instrument in Chapters \ref{chap:hd163296} and \ref{chap:radial_flows}.  A fully extended configuration 
+will include baselines up to 16\,kilometer and the array will observe at wavelengths from 400\,microns to 
+7\,millimeters.  The ALMA Phasing Project (APP) will enable this giant array to join VLBI experiments and 
+fringes to APEX were found during a January 2015 test \citep{matthews15}.
 
@@
 
-The canonical picture of star formation breaks this process into several stages.  Picture first a large 
+Both the SMA and ALMA are excellent observatories with which to study young stars.  Within a few hundred parsecs 
+of the Sun there are multiple star forming regions and stellar moving groups.  Their relative proximity to us 
+makes it possible to resolve physical structures in these regions and our great observatories have returned 
+spectacular glimpses of the complicated processes that produce a star.  The canonical picture of star formation 
+breaks this process into several stages \citep{shu87,mckee07}.  Picture first a large 
 molecular cloud within which there are complicated sub structures: clumps and cores of 
 concentrated gas and dust.  Over time these self-gravitating units will contract,  the material collapses into 
-a disk to conserve angular momentum, and nascent stars ignite their thermonuclear engines.  Jets, envelopes, 
-winds, and outflows punctuate this 
-process which may likely be disturbed by the swift pull of a passing star or the constant presense of a nearby 
-neighbor.  Whole fields, techniques, and technologies have been developed to study single chapters of this 
-complicated story and ours begins near the very end when only the disk and young star remain.
+a disk to conserve angular momentum, and nascent stars ignite their thermonuclear engines.  Jets, winds, and 
+outflows punctuate this process.  The system might also be perturbed by the swift pull of a nearby passing star 
+or the constant presence of stellar partners forming a hierarchical system.  Whole fields, techniques, and 
+technologies have been developed to study single chapters of this 
+complicated story and ours begins near its end when only the disk and young star remain.
 
@@
 born.  Their composition and structure provides the crucial initial conditions that must enable a diverse 
-population of planets.  However, theories of planet formation are still developing and require the consideration 
-of physical processes acting on many physical scales and over a huge range of times.  As such, there 
-work to test and develop these ideas is very diverse, ranging from experiments on grain interactions in 
-microgravity to space missions exploring the larger rocks that remain as testament to the formation of our 
-own Solar system.
+population of planets.  Theories of planet formation are still developing and require the consideration 
+of physical processes acting on many physical scales and over a huge range of times.  As such, the
+work to test and develop these ideas is quite varied, ranging from theoretical calculations to controlled 
+experiments in the lab to the frontiers of space exploration.
 
-figure: early images of disks
+\begin{figure}[t]
+\vspace{4cm}
+\includegraphics[width=0.5\textwidth]{chapters/koerner93.jpg}
+\caption{\citet{koerner93} imaged the disk around GM Aur using the Owens Valley Millimeter array. This figure
+from their paper 
+show maps of the $^{13}$CO $J$=2$-$1 emission which on the left is integrated over velocity (or frequency) and 
+the right has been weighted by the channel velocity.}
+\end{figure}
 
 Over the past several decades, observations from interferometers operating at millimeter wavelengths have made
-significant contributions to understanding the structure of these disks.  The early maps of CO made using 3 to 5 
-telescopes in an array at 
-Owens Valley revealed bound, rotating disks with diameters 
-similar to the dimensions of own solar system 
-\citep{sargent87,weintraub87,weintraub89,sargent91,koerner93,koerner93b,koerner95}.  The emission pattern
-matching Keplerian 
-rotation was then used to measure some systems' dynamical masses 
-\citep{dutrey98,guilloteau98,guilloteau99,simon00} and
-maps of multiple species to constrain the two-dimensional structure 
-\citep{aikawa99,aikawa01,vanzadelhoff01,aikawa02}.   Soon arrays detected the giant central dust holes of 
-``transition disks''
-\citep{hughes07,brown09,andrews11}.
+significant contributions to understanding the structure of these disks.  The early maps of the CO emission
+made using 3 to 5 telescopes in an array at Owens Valley revealed bound, rotating disks with diameters 
+similar to the dimensions of own solar system \citep{sargent87,weintraub89,koerner93}.  Emission from other 
+molecules were then surveyed \citep{dutrey97} and studies of multiple isotopologues were used to constrain the 
+two-dimensional structure of some disks \citep{dartois03,pietu07,qi11} aided by the development of 
+chemical models \citep{aikawa99,aikawa01}.  Images of the dust also revealed interesting structures such as
+giant holes centered around the star for a class of ``transition disks'' \citep{hughes07,brown09,andrews11} and 
+evidence for binary interactions \citep{akeson98}.
 
-figure: disk structure cartoon
+%% figure: disk structure cartoon?
 
 The main two ingredients in the disk are gas and dust.  The majority of the disk mass will be in molecular 
-hydrogen, but observations focus on easier detect trace molecules such as $^{12}$CO and its isotopologues.
+hydrogen, but observations focus on easier to detect trace molecules such as $^{12}$CO and its isotopologues.
 The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
-can give a reliable measure of mass (CITE).  Interferometric observations taken at millimeter wavelengths
-can resolve structure of nearby protoplanetary disks on scales of a few tens to hundreds of AU.  The motion of 
-material in the disk can also be studied using these same telescopes by studying how the pattern of line emission
-from molecules such as CO changes with frequency.  The bulk motion of the gas relative towards the observer 
+can give a measure of the dust mass \citep{hildebrand83}.  Interferometric observations taken at 
+millimeter wavelengths can resolve structure of nearby protoplanetary disks on scales of a few tens to 
+hundreds of astronomical units.   The bulk motion of material in the disk can also be 
+studied using these same telescopes by studying how the pattern of line emission
+from molecules such as CO changes with frequency \citep{beckwith93}.  The velocity of the gas relative to the 
+observer 
 changes as a function of position in the disk. To an observer scanning over frequency, different parts of the 
@@
 V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
-protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present Submillimeter Array (SMA) observations of this 
-disk 
-that resolve for the
-first time a large central cavity.  In addition, we show that the distribution of gas and dust are non-commitant 
-with each other: the gas disk is large and extended while the mm-sized dust grains are concentrated in a narrow 
-ring.  Furthermore, we show that dust within ring is not completely cleared.  We argue that this configuration is 
-the hallmark of an old disk that is being sculpted by at least one embedded planet.  HD163296 also has a 
-very large, bright gaseous disk, but no resolved cavity.
+protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present SMA observations 
+of this disk that reveal a large central cavity in the dust.  In addition, we show that the distribution dust 
+does not follow that of the gas: the gas disk is large and extended while the 
+mm-sized dust grains are concentrated in a narrow ring.  Furthermore, it appears that the amount of dust left 
+within the ring along with the size of the canvity is inconsistent with photoevaporative scenarios.  We argue 
+that this configuration is the hallmark of an old disk that is being sculpted by at least one embedded planet.
 
-In chapter \ref{chap:hd163296} we 
-use 
-science verification data from the Atacama Large Millimimeter Array (ALMA) to study the vertical structure of 
+HD163296 also has a very large, gas rich disk, but no observed central cavity.  In chapter \ref{chap:hd163296} 
+we use science verification data from ALMA to study the vertical structure of 
 this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
-its distance from mid-plane and explore additional, nuanced aspects of the disk structure and how they may 
-affect the observables.
+its distance from mid-plane and explore how additional, nuanced aspects of the disk structure may affect
+the data.  Chapter \ref{chap:radial_flows} then considers what impact radial motions of gas within transition 
+disk holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
+with possible sources of confusion such as warps in the disk.  One disk has provided intriguing
+evidence of such flows \citep{casassus13,cassasus15} and future observations may provide others.
 
-In chapter \ref{chap:radial_flows} we consider what impact radial motions of gas within transition disk 
-holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
-with possible sources of confusion such as warps in the disk.
+\begin{figure}[t]
+\plotone{chapters/EHT-baselines.png}
+\caption{A selection of EHT baselines viewed looking back from the galactic center where Sgr$^\ast$ lies.  The 
+2015 campaign included sites 
+from Hawaii, California, Arizona, Mexico, and Spain. ({\it image credit}: the EHT collaboration)}
+\end{figure}
 
 \section{Sample Rate Conversion for the EHT}
-
-figure: map of EHT site locations
 
@@
 incorporating new technologies into its operations.  The March 2015 observing campaign was a major milestone in
-the EHT expansion with participation from the LMT in Mexico, CARMA in California, the SMA and JCMT in Hawaii, 
-SMT in Arizona and 30m at Pico Valeta in Spain.  However, the data taken at the SMA must be processed before it
+the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
+Research in Millimeter-wave Astornome (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
+ Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  However, the data taken at the SMA must be processed before it
 can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the APHIDS pipeline

### 2015-08-28 chapters/conclusion.tex
minor updates

--- 2015-08-03 chapters/conclusion.tex
@@
-\chapter{Conclusions} \label{chap:conclusion}
+\chapter{Conclusion} \label{chap:conclusion}
 
-One might argue that the space of discovery in Astronomy is spanned by resolution and 
-wavelength.  Technology has enabled us to contemplate phenomena at wavelengths far outside 
-what the realm of our human senses, revolutionizing fields like high energy astrophysics,
-cosmology, and the study of the inter stellar-medium.  Resolution then reveals the hidden
-details and helps us to interpret what these new eyes see.  A recent example is 
-the application of interferometry at sub-mm wavelengths with ground breaking telescopes like the SMA, 
-the amazing new power of ALMA, and the EHT revolution.  
-While the science motivating chapters \ref{chap:v4046_structure} 
-through \ref{chap:radial_flows} may be different than in the final chapter of this thesis, 
-I am left wondering and waiting for what will we see when we look only a little bit closer.
+We conclude with brief summaries of the preceding chapters along with some discussion of recent developments 
+in the field.
 
-Emboldened by this spirit, we shall conclude not only with summaries of the work presented here,
-but also some brief discussion of recent developments in the field.  In chapter \ref{chap:v4046_structure} we 
+In Chapter \ref{chap:v4046_structure} we 
 used the SMA to study the protoplanetary disk 
-around the close binary V4046 Sgr.  This disk featured a narrow ring of large mm-sized grains, 
-but an extended gas rich disk.  Detailed modeling of these features demonstrated that they were 
-phenomenologically consistent with the viscous spreading of the gas and the concentration through pressure 
-drag forces of the dust.  This disk is a single example among a number, and more such examples continue to be 
-found, \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A somewhat related phenomenon are the 
-severely asymmetric dust  ``horse-shoes'' that have been recently observed at multiple wavelengths
+around the close binary V4046 Sgr.  This disk featured a narrow ring of large mm-sized grains along with
+an extended gas rich disk.  Detailed modeling of these features demonstrated that they were 
+consistent with the viscous spreading of the gas and the concentration of the dust via pressure 
+drag forces.  This disk is one published example among a few and more similar systems continue to be
+identified \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A related phenomena are the 
+severely asymmetric dust  ``horse-shoes'' that have been  observed at multiple wavelengths
 \citep{vandermarel13,fukagawa13,isella13,casassus15,marino15}.  Thought to be perhaps generated by gigantic 
-3D Rossby wave instabilities \citep{regaly13}, these are another striking example from the interplay between
-the gas and dust in protoplanetary disks.
+3D Rossby wave instabilities \citep{regaly13}, these are another striking example of the interplay between
+the gas and dust in protoplanetary disks.  A particular exciting example are the narrow gaps in the dust continuum
+of HL Tau (see Figure \ref{fig:hltau}. This striking image was obtained using ALMA baselines as long as 15\,km 
+and possible causes include
+planet-disk interactions and condensation fronts \citep{brogan14,zhang15}.
 
-Resolving magnetic fields: \citep{stephens14}
+\begin{figure}
+\plotone{chapters/HLTau_nrao.jpg}
+\caption{This ALMA observation of the HL Tau disk revealed multiple rings and gaps \citep{brogan15}.}
+\label{fig:hltau}
+\end{figure}
 
-HL Tau: \citep{brogan14}
+In Chapter \ref{chap:hd163296} we explored the vertical structure of a protoplanetary disk using 
+observations that exhibited clear signs of the kinetic temperature changing vertically in the disk.
+The search for azimuthal temperature variations are a natural extension of this work \citep{isella13} and 
+some evidence has been found \citep{vanderplas14}.  There is much activity designing 
+observations to probe the ambient conditions 
+for planet formation beyond temperature from the local ionization environment \citep{cleeves14,teague15} to 
+turbulence \citep{hughes11,guilloteau12,flock15,simon15}.
 
-A gap in Tw Hya at 20 AU? \citep{akiyama15}
+To conclude our work on protoplanary disks, we discussed in Chapter \ref{chap:radial_flows} observational 
+indicators of 
+radial gas flows within the holes of transition disks.  \citet{casassus15} have studied the gas kinematics for
+the curious case of HD142527 using the CO $J = 6-5$ line and found that their data is consistent with
+accretion reaching close to free-fall velocities.  The gas surface density of transition disks have also 
+been recently studied using observations taking at sub-millimeter wavelengths 
+\citep{bruderer14,zhang14,canovas15,perez15,vandermarel15} and will help determine how these structures are formed
+\citep{bruderer13}.  Molecular disk winds, outflows, and binary systems are all recently observed 
+with interesting gas kinematics \citep{klaasen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
+future, well resolved, highly sensitive observations of spectral lines may constrain the strength of MRI-drive
+turbulence \citep{simon15}, query the active sites of planet formation \citep{kleeves15,ober15}, or reveal
+circumplanetary disks \citep{perez15}.
 
-Gas density within cavities of transition disks \citep{vandermarel15}
+The ground-breaking Event Horizon Telescope continues to report exciting new science with every
+campaign from the early detections of AGN at mm wavelengths \citep{doeleman05} to structures
+at the event horizon of a supermassive black hole \citep{doeleman08,doeleman09,doeleman13}.
+The 2015 campaign represented a new scale of operation, incorporating more telescopes and
+technologies.  In Chapter \ref{chap:aphids} we reported on the status of the APHIDS software which
+will post-process the phased SMA data and make the sampling rate compatible with the rest of the EHT
+network.  This problem is not limited to the SMA and we hope that the development done with APHIDS will
+prove useful for other sites and VLBI experiments.
 
-More dust traps: \citep{espaillat15}
-
-\section{A Bright Future}
-What is happening in the inner disks? \citep{ke14}
-Circumplanetary disks? \citep{perez15}
-Chemistry and stuff \citep{oberg15} \citep{graninger15} 
-Chemistry as a signpost of planet formation \citep{cleeves15}
-Exploring the outer solar system with New Horizons
-Detection of organics by Rosetta? Checkout: \url{http://www.sciencedaily.com/releases/2015/07/150730172518.htm}
-
-Summarize status of EHT and its future.
-
-This thesis is a small expression of 
-the science that resolution at these wavelengths can provide and only an exploration for the 
+This thesis explores
+ence that resolution at these wavelengths can provide and only an exploration for the 
 hard and promising work ahead.

### 2015-08-28 chapters/introduction.tex
minor updates

@@
 giant holes centered around the star for a class of ``transition disks'' \citep{hughes07,brown09,andrews11} and 
-evidence for binary interactions \citep{akeson98}.
+evidence for binary interactions \citep[i.e.][]{akeson98}.
 
@@
 The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
-can give a measure of the dust mass \citep{hildebrand83}.  Interferometric observations taken at 
+can be used to measure of dust mass \citep{hildebrand83}.  Interferometric observations taken at 
 millimeter wavelengths can resolve structure of nearby protoplanetary disks on scales of a few tens to 
@@
 from molecules such as CO changes with frequency \citep{beckwith93}.  The velocity of the gas relative to the 
-observer 
-changes as a function of position in the disk. To an observer scanning over frequency, different parts of the 
-disk seem to light up when its Doppler shifted emission matches the observer's frequency window.
+observer changes as a function of position in the disk. To an observer scanning over frequency, different parts 
+of the disk seem to light up when its Doppler shifted emission matches the observer's frequency window.
 
@@
 properties such as size, inclination, and orientation can be calculated from an image.  The amount of dust 
-in a disk can also be estimated by assuming a dust opacity and measuring the mm-wave flux.  Alternatively, one 
-may try to solve the inverse problem: assume some disk model, calculate the resulting emission, and then compare
-it to what is observed.  Often the density, temperature, and velocity profiles of the disks are assumed to 
+in a disk can also be estimated by assuming a dust opacity and measuring the millimeter wave flux.  
+Alternatively, one may try to solve the inverse problem: assume some disk model, calculate the resulting 
+emission, and then compare the model's prediction to what is observed \citep[i.e.][]{dartois03,qi03}.  Often the 
+density, temperature, and velocity profiles of the disks are assumed to 
 follow power laws or some other parametric function.  There also exist some grids and models that try to 
@@
 interesting disks.  The work
-presented in chapters \ref{chap:v4046sgr_structure} to \ref{chap:radial_flows} takes the latter approach and uses simple parametric 
+presented in chapters \ref{chap:v4046sgr_structure} to \ref{chap:radial_flows} takes the latter approach and 
+uses simple parametric 
 models to investigate the structure of a choice few, but very interesting, protoplanetary disks.
 
-V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
+V4046 Sgr is a system with two stars closely orbiting each other and around this tight binary is a bright
 protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present SMA observations 
@@
 does not follow that of the gas: the gas disk is large and extended while the 
-mm-sized dust grains are concentrated in a narrow ring.  Furthermore, it appears that the amount of dust left 
-within the ring along with the size of the canvity is inconsistent with photoevaporative scenarios.  We argue 
-that this configuration is the hallmark of an old disk that is being sculpted by at least one embedded planet.
+millimeter sized dust grains are concentrated in a narrow ring.  Furthermore, thethe amount of dust left inside
+the cavity is inconsistent with photoevaporative scenarios.  We argue 
+that this configuration is the hallmark of an old disk that is being sculpted by at least one unobserved 
+companion.
 
@@
 this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
-its distance from mid-plane and explore how additional, nuanced aspects of the disk structure may affect
+its distance from the disk mid-plane and explore how additional, nuanced aspects of the disk structure may affect
 the data.  Chapter \ref{chap:radial_flows} then considers what impact radial motions of gas within transition 
@@
 
-The Event Horizon Telescope (EHT) observes supermassive black holes, obtaining resolution high enough
-to resolve event horizon scales of nearby targets.  These observations provide a 
+The Event Horizon Telescope (EHT) observes supermassive black holes with baselines long enough to
+to resolve event horizon scales of nearby targets.  These VLBI observations provide a 
 unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
 and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
-center and the giant elliptical galaxy M87 \citep{doeleman08,doeleman09}.  These data have provided indirect 
-evidence for existence of an event horizon \citep{broderick15} along with constraints on the degree of 
+center and the blaock hole at the center of the giant elliptical galaxy M87 
+\citep{doeleman08,doeleman09}.  These data have provided indirect 
+evidence for the existence of an event horizon \citep{broderick15} along with constraints on the degree of 
 magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
 observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
-well as new studies of black hole time variability \citep{doeleman09} and black hole accretion \citep{chan15}.
+well as new studies of black hole time variability \citep{doeleman09} and accretion \citep{chan15}.
 
-The bright future of the EHT rests on its key ability to expand by adding more telescopes and
-incorporating new technologies into its operations.  The March 2015 observing campaign was a major milestone in
+The bright future of the EHT is partly driven by its ability to add more telescopes and
+incorporate new technologies.  The March 2015 observing campaign was a major milestone in
 the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
-Research in Millimeter-wave Astornome (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
- Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  However, the data taken at the SMA must be processed before it
+Research in Millimeter-wave Astornomy (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
+ Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  However, the data taken at the SMA must be 
+processed before it
 can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the APHIDS pipeline
 which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
-is intended to be a long-term solution to enable VLBI with the SMA and has applications for other sites such 
-as ALMA that operate at a sampling rate that is not compatible with the standard VLBI data product.
+is intended to be a long-term solution enabling VLBI with the SMA and has applications for other sites such 
+as ALMA.

### 2015-08-31 aphids/aphids.tex
added aphids chapter

\newcommand{\z}{\mathpzc{z}}
\newcommand{\falign}{0}
\newcommand{\talign}{7.5}

\chapter{Resampling phased SWARM for VLBI}
\label{chap:aphids}

%%\section*{abstract}

To enbable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
the data before correlating.  In this chapter we describe a software solution to change the 
samping rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
focus on the rate conversion problem and investigate several possible algorithms.  The project
has applications for future VLBI observations at the SMA as well as other sites with similar 
challenges.

\section{Introduction}

Very Long Baseline Interferometry (VLBI) uses similar techniques as the connected element arrays
featured in the previous chapters.  By using detectors located thousands of miles away, perhaps on different 
continents or even in space, this technique can achieve milliarcsecond resolution at radio wavelengths.  Each 
station must have its own clock and, after the observation campaign, the recorded signals is correlated off-site. 
Examples of VLBI applications include geodesy, astrometry, and imaging. Technical developments in the 
field continue to unveil singular scientific opportunities.

One exciting example is the development of mm and sub-mm wave VLBI by collaborations such as the Event Horizon
Telescope (EHT).  The EHT observes nearby super massive black holes on scales of their Schwarzschild radius which
provides a unique lab to study 
general relativity and accretion physics.  One of the challenges that this experiment faces is the inhomogeneity 
of its array since it is a conglomeration of many independently operating observatories.
While many of these telescopes use the same back end hardware such as the state-of-the-art 
ROACH2 VLBI Digital Back-end (R2DBE), a few sites 
cannot use these systems and produce very 
different data products. In particular, the sampling rate of the phased ALMA, SMA, and CARMA arrays are 
incompatible with the standard VLBI rate.  Before the EHT's software correlator
\citep[DIFX;][]{deller07} can find fringes using these data, they must be resampled to the R2DBE rate of 
4096\,MHz.

This chapter summarizes the Adaptive Phased-array and Heterogeneous 
Interpolator, and Downsampler for SWARM (APHIDS), a resampling software designed for 
phased SMA data products.  We concentrate our discussion to early development and design in the project with 
a particular emphasis on the resampling problem.  APHIDS also has very general applications for 
post-processing of VLBI data since it features fast read/write capabilities, a modular threaded pipeline 
and a GPU enabled fast data processing.  The code base is open-source and available online at 
\url{https://github.com/sma-wideband/sdbe}.

The purpose of APHIDS is to take data recorded at the SMA and process it for correlation using
DIFX.  The SMA is a key site for the EHT that provides important East-West baselines and a zero baseline with 
its neighbor JCMT.  The phased sum is calculated by the SMA correlator, the SMA Wideband Astronomical 
ROACH2 Machine (SWARM\footnote{\url{https://www.cfa.harvard.edu/twiki5/view/SMAwideband}}),
which is currently being deployed in stages as the array's bandwidth capabilities are 
upgraded.  For EHT observations in 2015, SWARM recorded at a sampling rate of 2496\,MHz in two separate phased
sums.

In \S\ref{sec:src_basics} we discuss the sampling rate conversion problem and several common solutions.
The current APHIDS resampling algorithm is described in \S\ref{sec:aphids} along with discussions of important
design considerations.  We include measurements of performance as well as a successful
demonstration of the software's operation.  We end by summarizing the goals for near and long-term project 
development.

\section{Sample Rate Conversion} \label{sec:src_basics}

Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  Operationally, 
sample rate conversions is the combination of upsampling and downsampling \citep{oppenheim10,lyons11}.  
Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
time and frequency domain for when $L$ equals 3.

 %% upsampling cartoon
\begin{figure}[t]
\begin{tikzpicture}[
declare function={ 
    funcx(\x) = -0.5*sin(2*pi*0.2*\x r) + 0.2*cos(2*pi*0.32*\x r) + 
                0.3*sin((2*pi*0.9*\x - 0.3) r) + 0.6*sin(2*pi*0.5*\x r);},
]
% frequency domain

%% |X|
\draw [<->, rounded corners, very thick](\falign,1.5) -- (\falign,0) --(\falign+6.5,0) node[below] {$f$};
\node[left] at (\falign,0.7) {$|X|$};
\node[below] at (\falign+1,0) {$f_0/2$};
\foreach \i in {1,...,6}
  \draw[thick] (\i, 0) -- (\i,0.1);
  \draw[very thick] (\falign,1) to [out=0, in=170] 
                    (0.5+\falign,1.03) to [out=350, in=105] 
                    (\falign+1,0);
  \draw[very thick,dashed] (\falign+1,0) to [out=75, in=190]  
                           (\falign+1.5,1.03) to [out=10, in=180] 
                           (\falign+2, 1) to [out=0,in=170] 
                            (\falign+2.5, 1.03) to [out=350, in=105] 
                            (\falign+3,0);
  \draw[very thick,dashed] (\falign+3,0) to [out=75, in=190]  
                            (\falign+3.5,1.03) to [out=10, in=180] 
                            (\falign+4, 1) to [out=0,in=170] 
                            (\falign+4.5, 1.03) to [out=350, in=105] 
                            (\falign+5,0);
  \draw[very thick,dashed] (\falign+6-0,1) to [out=170, in=0] 
                            (\falign+6-0.5,1.03) to [out=190, in=75] 
                            (\falign+6-1,0);

  %% W
  \draw [<->, rounded corners, very thick](\falign,-0.5) -- (\falign,-2) -- (\falign+6.5,-2);
  \node[left] at (\falign,0.7-2) {$|X_L|$};
  \node[below] at (\falign+3,-2) {$Lf_0/2$};
  \foreach \i in {1,...,6}
    \draw[thick] (\i, -2) -- (\i,0.1-2);
  \draw[very thick] (\falign,1-2) to [out=0, in=170] 
                    (0.5,1.03-2) to [out=350, in=105] 
                    (\falign+1,-2);
  \draw[very thick,] (\falign+1,-2) to [out=75, in=190]  
                     (\falign+1.5,1.03-2) to [out=10, in=180] 
                     (\falign+2, 1-2) to [out=0,in=170] 
                     (\falign+2.5, 1.03-2) to [out=350, in=105] 
                     (\falign+3,-2);
  \draw[very thick,dashed] (\falign+3,-2) to [out=75, in=190]  
                            (\falign+3.5,1.03-2) to [out=10, in=180] 
                            (\falign+4, 1-2) to [out=0,in=170] 
                            (\falign+4.5, 1.03-2) to [out=350, in=105] 
                            (\falign+5,-2);
  \draw[very thick,dashed] (\falign+6-0,1-2) to [out=170, in=0] 
                            (\falign+6-0.5,1.03-2) to [out=190, in=75] 
                            (\falign+6-1,-2);

%% H
  \draw [<->, rounded corners, very thick](\falign,-2.5) -- (\falign,-4) -- (\falign+6.5,-4);
  \foreach \i in {1,...,6}
    \draw[thick] (\i, -4) -- (\i,0.1-4);
  \node[left] at (\falign,0.7-4) {$|H|$};
  \node[below] at (\falign+1,-4) {$f_0/2$};
  \node[below] at (\falign+6,-4) {$Lf_0$};
  \draw[very thick] (\falign+0,1-4) -- (\falign+0.8,1-4) -- (\falign+1,0-4);
  \draw[very thick,dashed] (\falign+5,0-4) -- (\falign+5+0.2,1-4) -- (\falign+6,1-4);

  %% Y = H \ast X_L 
  \draw [<->, rounded corners, very thick](\falign,-4.5) -- (\falign,-6) -- (\falign+6.5,-6) node[below] {$f$};
  \foreach \i in {1,...,6}
    \draw[thick] (\i, -6) -- (\i,0.1-6);
  \node[left,align=right] at (\falign,0.7-6) {$|Y|$};
  \node[below] at (\falign+3,-6) {$f_1/2$};
  \draw[very thick] (\falign,1-6) to [out=0, in=170] 
                    (0.5,1.03-6) to [out=350, in=105] 
                    (\falign+1,-6);
  \draw[very thick,dashed] (\falign+6-0,1-6) to [out=170, in=0] 
                            (\falign+6-0.5,1.03-6) to [out=190, in=75] 
                            (\falign+6-1,-6);

  %%%%%%%%%%%%%%%
  % time domain
  %% x
  \draw [<->, rounded corners, very thick](\talign,1.5) -- (\talign,0) -- (\talign+6.5,0) node[below] {$i$};
  \node[left] at (\talign,0.7) {$x$};
  \foreach \i in {0,...,7}
    \draw[thick] (3*\i/4+\talign,0) -- (3*\i/4+\talign,{funcx(\i/2.)});
  \foreach \i in {0,...,7}
    \draw[fill] (3*\i/4+\talign,{funcx(\i/2.)}) circle[radius=0.05];
  \draw[very thick, dashed, domain=0:23,samples=50] plot(\x/4+\talign, {funcx(\x/6.)});

  %% x_L
  \draw [<->, rounded corners, very thick](\talign,-0.5) -- (\talign,-2) -- (\talign+6.5,-2) node[below] {$i$};
  \node[left] at (\talign,0.7-2) {$x_L$};
  \foreach \i in {0,...,7}{
    \draw[thick] (3*\i/4+\talign,-2) -- (3*\i/4+\talign,{funcx(\i/2.)-2});
    \draw[fill] (3*\i/4+\talign,{funcx(\i/2.)-2}) circle[radius=0.05];}
  \foreach \i in {0,...,7}
    \foreach \j in {1,2} 
    \draw[fill] (3*\i/4+\talign+\j/4,0-2) circle[radius=0.05];
  \draw[very thick, dashed, domain=0:23,samples=50] plot(\x/4+\talign, {funcx(\x/6.)-2});

 %% y
  \draw [<->, rounded corners, very thick](\talign,-4.5) -- (\talign,-6) -- (\talign+6.5,-6) node[below] {$j$};
  \node[left] at (\talign,0.7-6) {$y$};
  \foreach \i in {0,...,22}{
    \draw[thick] (\i/4+\talign,-6) -- (\i/4+\talign,{funcx(\i/6.)-6});
    \draw[fill] (\i/4+\talign,{funcx(\i/6.)-6}) circle[radius=0.05];}
  \draw[very thick, dashed, domain=0:23,samples=50] plot(\x/4+\talign, {funcx(\x/6.)-6});

\end{tikzpicture}
\caption{A sketch of upsampling when $L = 3$. The left column shows how the spectrum changes with each operation
and the right column mirrors the effect in the time domain.}
\label{fig:upsampling}
\end{figure}

 %% cartoon explaining down-sampling
 % http://tex.stackexchange.com/questions/124878/declare-function-for-tikzpicture
\begin{figure}[t!]
\begin{tikzpicture}[
declare function={ 
    funcx(\x) = -0.1*sin(2*pi*2.8*\x r) + 0.2*sin(2*pi*1.3*\x r) + 
                0.3*sin((2*pi*0.9*\x - 0.3) r) + 0.6*sin(2*pi*0.5*\x r);
    funchx(\x) = 0.3*sin((2*pi*0.9*\x - 0.3) r) + 0.6*sin(2*pi*0.5*\x r);},
]
  % frequency domain

  %% |X(f)|
  \draw [<->, rounded corners, very thick](0,1.5) -- (0,0) --(6.5,0) node[below] {$f$};
  \node[left] at (0,0.7) {$|X|$};
  \foreach \i in {1,...,6}
    \draw[thick] (\i, 0) -- (\i,0.1);
  \node[below] at (3,0) {$f_0/2$};
  \draw[very thick] (0,1) to [out=10, in=170] (0.5,1.03)
                        to [out=350, in=190] (2,0.95) 
                        to [out=10, in=170] (2.5,1)
                        to [out=-10, in=110] (3,0);
  \draw[very thick,dashed] (6-0,1) to 
                            [out=170, in=10] (6-0.5,1.03) to 
                            [out=190, in=350] (6-2,0.95) to 
                            [out=170, in=10] (6-2.5,1) to [out=190, in=70] (6-3,0);

  %% |H(f)|
  \draw [<->, rounded corners, very thick](0,-0.5) -- (0,-2) -- (6.5,-2);
  \node[left] at (0,0.7-2) {$|H|$};
  \foreach \i in {1,...,6}
    \draw[thick] (\i, -2) -- (\i,0.1-2);
  \node[below] at (1,0-2) {$f_0/2M$}; \node[below] at (6,0-2) {$f_0$};
  \draw[very thick] (0,1-2) -- (0.8,1-2) -- (1,0-2);
  \draw[very thick,dashed] (5,0-2) -- (5.2,1-2) -- (6,1-2);

  %% |W(f)|
  \draw [<->, rounded corners, very thick](0,-2.5) -- (0,-4) -- (6.5,-4);
  \node[left] at (0,0.7-4) {$|HX|$};
  \foreach \i in {1,...,6}
    \draw[thick] (\i, -4) -- (\i,0.1-4);
  \node[below] at (1,0-4) {$f_0/2M$}; \node[below] at (6,0-4) {$f_0$};
  \draw[very thick] (0,1-4) to [out=10, in=170] (0.5,1.03-4) to [out=350, in=105] (1,0-4);
  \draw[very thick,dashed] (6-0,1-4) to [out=170, in=10] (6-0.5,1.03-4) to [out=190, in=75] (6-1,0-4);

  %% Y(f)
  \draw [<->, rounded corners, very thick](0,-4.5) -- (0,-6) -- (6.5,-6) node[below] {$f$};
  \node[left] at (0,0.7-6) {$|Y|$};
  \foreach \i in {1,...,6}
    \draw[thick] (\i, -6) -- (\i,0.1-6);
  \node[below] at (1,0-6) {$f_1/2$}; 
  \draw[very thick] (0,1-6) to [out=10, in=170] (0.5,1.03-6) to [out=350, in=105] (1,0-6);
  \draw[very thick,dashed] (1,0-6) to [out=75, in=190]  (1.5,1.03-6) to [out=10, in=170] 
                            (2, 1-6) to [out=10,in=170] (2.5, 1.03-6) to [out=350, in=105] (3,-6);
  \draw[very thick,dashed] (3,0-6) to [out=75, in=190]  (3.5,1.03-6) to [out=10, in=170] 
                            (4, 1-6) to [out=10,in=170] (4.5, 1.03-6) to [out=350, in=105] (5,-6);
  \draw[very thick,dashed] (6-0,1-6) to [out=170, in=10] (6-0.5,1.03-6) to [out=190, in=75] (6-1,0-6);

  %%%%%%%%%%%%%%%
  % time domain
  %% x
  \draw [<->, rounded corners, very thick](7.5,1.5) -- (7.5,0) -- (13.5,0) node[below] {$i$};
  \node[left] at (7.5,0.7) {$x$};
  \foreach \i in {1,...,22}{
    \draw[thick] (\i/4+7.5,0) -- (\i/4+7.5,{funcx(\i/6.)});
    \draw[fill] (\i/4+\talign,{funcx(\i/6.)}) circle[radius=0.05];}
  \draw[very thick, dashed, domain=0:22,samples=100] plot(\x/4+7.5, {funcx(\x/6.)});

  %% h \ast x
  \draw [<->, rounded corners, very thick](7.5,1.5-4) -- (7.5,0-4) -- (13.5,0-4) node[below] {$i$};
  \node[left] at (7.5,0.7-4) {$h\ast x$};
  \foreach \i in {1,...,22}{
    \draw[thick] (\i/4+7.5,-4) -- (\i/4+7.5,{funchx(\i/6.)-4});
    \draw[fill] (\i/4+\talign,{funchx(\i/6.)-4}) circle[radius=0.05];}
  \draw[very thick, dashed, domain=0:22,samples=50] plot(\x/4+7.5, {funchx(\x/6.)-4});

  %% y 
  \draw [<->, rounded corners, very thick](7.5,1.5-6) -- (7.5,0-6) -- (13.5,0-6) node[below] {$j$};
  \node[left] at (7.5,0.7-6) {$y$};
  \foreach \i in {1,...,22}
    \draw[gray] (\i/4+7.5,-6) -- (\i/4+7.5,{funchx(\i/6.)-6});
  \foreach \i in {0,...,7}{
    \draw[thick] (3*\i/4+7.5,-6) -- (3*\i/4+7.5,{funchx(3*\i/6.)-6});
    \draw[fill] (3*\i/4+\talign,{funchx(3*\i/6.)-6}) circle[radius=0.05];}
  \draw[very thick, dashed, domain=0:22,samples=50] plot(\x/4+7.5, {funchx(\x/6.)-6});

\end{tikzpicture}
%\caption{Series of steps for down sampling shown in the frequency domain when $M=3$.}
\caption{A sketch of downsampling when $M = 3$. The left column shows how the spectrum changes with each operation
and the right column mirrors the effect in the time domain.}
\label{fig:downsampling}
\end{figure}

Downsampling, or decimation, by an integer factor of $M$ requires that the signal be first low-pass filtered to 
avoid aliasing at frequencies greater than the desired rate $f_0/M$.  Next, the sampling rate is reduced to 
$f_0/M$ by selecting every $M$th sample from the filtered signal.  Figure \ref{fig:downsampling} shows this 
process in the time and frequency domain for when $M$ equals 3.

Sample rate conversion by a factor $L/M$ can be conceptualized as a three step process where the original signal 
is expanded by $L$, filtered, and then decimated by $M$.  As illustrated by Figure \ref{fig:resample_basic}, 
the serial anti-imaging and 
anti-aliasing filters are combined into a single filter with cutoff frequency of 
$1/\mathrm{max}(L,M) \times Lf_0/2$.  Under this scheme, interpolation must precede decimation otherwise desired 
frequency components greater than $f_0/M$ cannot be preserved.

Computationally, this basic sample rate conversion scheme is highly inefficient.  The filtering is applied at 
the highest possible sample rate, $Lf_0$, on a time-series that is $(L-1)/L$ zeros by fraction. Furthermore, the 
decimator discards an $(M-1)/M$ fraction of the samples.  As a result, this algorithm requires a lot of memory 
and spends precious clock time with wasted math.  There is a vast literature of techniques and algorithms that 
improve performance including multistage resampling, folded filter structures, and polyphase representations 
\citep{oppenheim10,lyons11,vaidyanathan93}.

\begin{figure}[t]
\centering
\begin{tikzpicture}
   \label{fig:resample_basic}
   \node[dspnodeopen,dsp/label=left]  (c0) {$x[i]$};
   \node[dspsquare,right=of c0]                     (c1) {\upsamplertext{L}};
   \node[dspfilter,right=of c1]                     (c2) {$H(\z)$};
   \node[dspsquare,right=of c2]                    (c3) {\downsamplertext{M}};
   \node[dspnodeopen,right=of c3,dsp/label=right]  (c4) {$y[j]$};
   \foreach \i [evaluate = \i as \j using int(\i+1)] in {0,1,...,3}
       \draw[dspconn] (c\i) -- (c\j);
\end{tikzpicture}
\caption{Basic multirate resampling signal flow graph.}
\end{figure}

There have been recent development of polyphase filter representations for GPU architectures 
\citep[i.e.][]{vanderveldt12,adamek14,kim14a}.  These filter structures can operate at the low sample rate of 
$f_0/M$
by introducing delays to switch the expander and decimatator \citep{crochiere81}.  However, they require 
some filter design and careful book-keeping or buffering to keep track of sample indices \citep{wang01}.
The implementation in \citet{kim14a} assigned each thread to a filter coefficient, parallelizing the 
inner product operation.  However, \citet{kim14a} found that their GPU kernel was dominated by the 
indexing operations.  \citet{adamek14} (building on the work by \citet{vanderveldt12}) presented tantalizing 
implementations of 
polyphase filter banks that could run at data rates in excess of 6.5\,GB/s (their estimate for the single channel 
output of the SKA's Low Frequency Aperture Array).

%  https://github.com/wesarmour/astro-accelerate

A simple option for resampling is to use linear interpolation:
\begin{equation}
y[i] = x[j] + (x[j+1] - x[j]) (if_0/f_1 - j)
\end{equation}
where $j = \mathrm{floor}(if_0/f_1)$.  This is equivalent to applying a ``tent'' FIR filter with $2L$ taps on the 
upsampled $Lf_0$ time series before decimation \citep{oppenheim10}.  Similarly, a $2L$ tap box-car filter is 
equivalent to nearest-neighbor interpolation:
\begin{equation}
y[i] = x[\mathrm{round}(if_0/f_1)].
\end{equation}

Neither linear or nearest-neighbor interpolation consider the frequency content of the original signal and, as a 
result, perform poorly for signals with any significant power near the Nyquist rate \citep{fraser89}.  To further 
emphasize this point, Figure 
\ref{fig:windows} shows the frequency response for both methods when 
the resampling factor $L/M$ is $64/39$.  Note that both methods have large side lobes and so the 
resampled signal will introduce a large slope in the passband.  However, GPUs can compute 
these low-order interpolations directly on the card using texture memory.  \citet{kim14b} utilize this feature 
by using the cuFFT library to upsample the signal in the Fourier domain by some initial factor, $U$, and then 
interpolating in texture memory.  It would be straight forward to 
expand upon the framework introduced by \citet{kim14b} to handle cases where $M$ is greater than $L$.  
This scheme should help reduce loss from aliasing, albeit 
at the cost of two FFTs and much larger memory requirements.

\begin{figure}[t!]
\epsscale{1.0}
\plotone{aphids/windows.eps}
\caption{Frequency response for FIR filters equivalent to linear and nearest interpolation when $L/M = 64/39$.
The orange region shows the first Nyquist zone of the target $L/Mf_0$ sample rate.  Spectral components at all 
other frequencies are aliased into this region.}
\label{fig:windows}
\end{figure}

\subsection{Resampling in the Fourier Domain} \label{sec:fourier_resampling}
In contrast to the previous methods, one can also implement a rational $L/M$ sample rate conversion entirely in 
the Fourier domain \citep{gold69,yeh82}.  After accumulating $kM$ samples at a clock rate of $f_0$ 
(where $k=1,2,3\cdots$), the DFT 
returns spectral components spaced at $f_0/kM$.  If $f_1 > f_0$, the resampled spectrum is generated by inserting
$p$ zeros to match the new $f_1$ while maintaining the correct frequency components: 
\begin{equation}
\frac{f_1}{kM+p} = \frac{f_0}{kM}.
\end{equation} 
Solving for $p = k(L - M)$.  If $f_0 < f_1$, the spectrum is instead trimmed by $p$ samples.  The time series
sampled at $f_1$ can then be constructed from the inverse DFT.  This method is equivalent to sinc 
interpolation using an ideal low-pass filter and is a perfect interpolator
if the original, continuous signal has a discrete spectral density distribution that is band-limited below the 
Nyquist limit.  In practice, small errors may be introduced from aliased spectral leakage.

Trimming or padding the spectrum is equivalent to convolving a normalized sinc function with the original series 
wrapped periodically.  Consequenty, errors will be introduced into the beginning and end of the
resampled signal.  The far this error significantly propagates away from the time series edges depends on the 
width of the sinc function which is set by the resampling factors $L$ and $M$: 
\begin{equation} \label{eq:sinc}
B\,\mathrm{sinc}(B t) \overset{\mathcal{F}}{\Longleftrightarrow} \begin{cases} 1 \quad |f| < B/2 \\ 0 \quad |f| > B/2 \end{cases}
\end{equation}
where $B = \mathrm{min}(f_0,L/Mf_0)$.  
The number of samples for which this makes a large difference is small and can be roughly treated as less than a 
1\% effect when $N > 100\,\mathrm{max}(L/M,1)$.  One option to mitigate this error to stitch together 
overlapping, resampled segments \citep{bi11} which requires more computation and memory.  A second strategy is to 
multiply the original signal by some tapered window function \citep{fraser89}.

For post-processing phased SWARM, DFT resampling is a good fit because one can access arbitrary large chunks of 
the time series without accumulators in hardware and has well-behaved errors
that can be easily modeled.  However, the speed of the FFT depends on the resampling factors ($L/M = 32/39$ for 
the March 2015 EHT observations) and zero-padding in the case that the $L\gg M$ may require large amounts of 
memory.

%% Loss table : /home/krosenfe/resampling/losses.py
% losses for FIR filtering, interpolation schemes, and FFT
\begin{table}
\begin{center}
\centering
\caption{Correlation coefficient loss from resampling \label{tab:loss}}
\begin{threeparttable}
\begin{tabular}{l|ccc}
\toprule
Method & 64/39 & 32/39 & 128/125 \\
\midrule
Nearest-Neighbor            & 13\% & 17\% & 13\% \\
Linear                      &  5\% &  7\% &  5\% \\
Hamming window ($16L$ taps) &  1\% &  2\% &  2\% \\
FFT ($N=M$)                 &  1\% &  1\% & 0.2\% \\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

\section{APHIDS} \label{sec:aphids}

We are developing a software solution to the resampling problem called APHIDS (Adaptive 
Phased-array and 
Heteregeneous Interpolator and Downsampler for SWARM).  This open-source project is written in C/CUDA using
the High Availibility Shared Pipeline 
(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
\ref{fig:aphids_flow_chart} shows how data progresses 
through the pipeline and this section will describe each block covering the major hardware components, software 
design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.

\tikzstyle{line} = [draw,-latex']
\tikzstyle{cpu} = [rectangle, draw, fill=blue!20,
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{gpu} = [rectangle, draw, fill=red!20,
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{resamp} = [rectangle, draw, minimum height=4.5em, align=center, minimum width=15em]

\begin{figure}[t]
\linespread{1.}\selectfont{}
\begin{center}
\begin{tikzpicture} [node distance = 5cm, auto]
\node [cpu] (mark6_in) {Mark6 Reader};
\node[gpu,below of=mark6_in](reader){Reader};
\node[gpu,right of=reader](resampler){Resampler};
\node[cpu,right of=resampler](VDIF_pkt){VDIF packet \\ \& header};
\node[cpu,below of=VDIF_pkt](mark6_out){Mark6 Writer};

\path[line] (mark6_in) -- node[align=center] {BENG @ 2496 GHz \\ quantized to 2-bits}(reader);
\path[line] (reader) -- node[align=center] {spectra of \\ 16k complex64} (resampler);
\path[line] (resampler) --  node[align=center] {time series \\ @ 2048 GHz}(VDIF_pkt);
\path[line] (VDIF_pkt) -- node[align=center, left]{VDIF @ 4096 GHz \\ quantized to 2-bits}(mark6_out);
\end{tikzpicture}
\label{fig:aphids_flow_chart}
\caption{This signal flow chart shows an overview of APHIDS.  The red boxes represent operations on the GPU 
while blue boxes are for the CPU.  Once the B-engine frames have been unpacked, each of the two phased sum 
streams are processed independently and in a serial fashion.}
\end{center}
\end{figure}

\subsection{Hardware and Data description}

APHIDS features three main hardware components: an I/O data system, CPU, and GPU server.  For I/O, the 
EHT observations taken by SWARM in March, 2015 have been stored on four disk-packs that each hold eight, 
6\,TB hard drives (totaling $48 \times 4 = 192$\,TB of data).  The Mark6 VLBI data system has four modules that 
can hold one disk-pack and we use
two seprate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
that controls the GPU server.

The GPU server as four GTX 980 GPUs, each with 4GB of local memory.  Data from the CPU (or host) is copied over 
a PCI-E bus into the local GPU (or device) memory.  Once the data is transferred, the host directs the device
to perform functions (or kernels) on the data which must be again moved from the local device memory onto 
on-chip memory for the individual streaming multiprocessors that make up the GPU.  After computation is done, 
the host transfers the data off of the device memory.

The observations were recorded using a scatter/gather file system which disperses the 
data composing individual scans over the four modules.  APHIDS uses fast, general purpose software called 
sgcomm (Scatter/Gather Communication) to read in 
the scattered data into CPU shared memory over a 10 Gigabit Ethernet network.  Data from phased SWARM was 
streamed through a special version of the R2DBE called the SDBE (Swarm Digital Back End) and packaged into
a customized B-engine format.  Each B-engine frame is composed of 1024, 1056B VDIF frames 
\footnote{\url{http://www.vlbi.org/vdif/}} with 128 snapshots of a 
16384 sample SWARM spectra.  However, the current SDBE bitcode packs
neither the snapshots contiguously in time nor the spectrum samples contiguously in frequency.  Furthermore, 
each B-frames overlaps the preceding one in memory.  Our resampling algorithm handles $19968$\,SWARM samples
simultaneously, so in order to reduce the amount of book-keeping the inout thread buffers 40 B-engine frames to 
produce $39 \times 128$ complete SWARM snapshots.  The inout thread does preliminary checks to ensure that the 
B-engine frames are complete before assigning each buffer to a GPU.  The data is written as a time series using 
following the VDIF specification.

As mentioned previously, the data is split into two phased sums that together cover the desired 4096\,MHz of 
observing bandwidth.  Since DIFX can handle the chunks independently we do not stitch the spectra together. 
However, the observation was designed to include guard bands which must be removed by APHIDS 
(see Figure \ref{fig:swarm_amp_spec}).

\begin{figure}[t!]
\epsscale{1.0}
\plotone{aphids/SWARM_amp_spectra.eps}
% /home/krosenfe/sdbe/software/prototyping/cuda/SWARM_amp_spectra.eps
\caption{This figure shows both channels of the detrended amplitude spectral density of phased SWARM along with 
the R2DBE (SMA single-dish in gray).}
\label{fig:swarm_amp_spec}
\end{figure}

\subsection{Resampling Block}\label{sec:resamp_block}

APHIDS resamples the data in the Fourier domain (see \S \ref{sec:fourier_resampling}).  There were several 
contributing factors that led us to make
this choice rather than apply filters in the time domain.  First, the GPU cluster can handle relatively large 
chunks of data at a time so losses from the DFT are very manageable.  Second, regardless of 
what resampling algorithm is used the SWARM spectra must 
be transformed into the time domain, the guard bands trimmed and the signal modulated.  These operations are done 
naturally in the Fourier domain.  However, the cost of the resampling operation will be largely dominated by 
these operations.  Lastly, we can utilize the optimized cuFFT library
\footnote{\url{https://developer.nvidia.com/cuFFT}} to ensure good performace, reduce development time and 
maintain a manageably sized codebase.

The APHIDS resampling block is composed of three 1-dimensional FFTs (A, B, and C; see Figure 
\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performace, we utilize batched
FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
and ensure that transforms operate on contiguous chunks of memory.  We also use single precision transforms to 
reduce the memory transfer bandwidth and ensure that 
the resampling block can be run on a single device.  The latter point is important since it allows the inout
thread to parallelize the resampling operation across the cluster and provide a roughly x4 speedup.

\begin{figure}[t!]
\linespread{1.}\selectfont{}
\centering
\begin{tikzpicture}[node distance = 2.2cm, auto]
\node[align=center](init){};
\node[resamp,below of=init](rfft){batched real FFT \\ dimension of $39 \times N$};
\node[resamp,below of=rfft](trim){Trim band and shift frequency};
\node[resamp,below of=trim](rifft){batched real IFFT \\ dimension of $32 \times N$};
\node[align=center,below of=rifft](final){};

\path[line](init) -- node[align=left]{ time series \\ @ 2496\,MHz} (rfft);
\path[line](rfft) -- (trim);
\path[line](trim) -- (rifft);
\path[line](rifft) -- node[align=left]{ time series \\ @ 2048\,MHz}(final);
\end{tikzpicture}
\label{fig:resampling_block}
\caption{The resampling block trims the phased SWARM guard bands and shifts the passband to DC as part of the 
resampling operation.}
\end{figure}

The first FFT (A) takes the complex valued, 16384 sample SWARM spectrum into the time domain, returning 
32768 real samples.  Since SWARM omits the Nyquist sample which is far outside the SWARM cutoff frequency, 
we insert a zero as the 16385th sample.  Since this transform is radix-2, the performance should benefit from 
the Cooley-Tukey algorithm.  However, on our cluster the cuFFT library exhibits linear improvement in Gflops 
with transform size up to N of 16384 (see Figure \ref{fig:C2R_performance}).  This behavior is consistent with 
the CUDA 7.0 performance 
report\footnote{\url{http://on-demand.gputechconf.com/gtc/2015/webinar/gtc-express-cuda7-performance-overview.pdf}} 
and we find that this transform, just ouside of the optimal regime, operates at 6.7 Gsps.

Next, we Fourier transform chunks of $N_B = 19968$ consecutive time series samples (B).  The resulting 
spectrum has a frequency resolution of $2496$\,MHz$/19958=125$\,kHz.  The folllowing step is to apply a third
complex-to-real transform (C) on a subset of the spectrum, using only frequences from $150$\,MHz to $1174$\,MHz.  
The final time series has size $N_C= 2\times(1174-150)/0.125=16384$ per batch element and our desired 
sampling rate of 
$16384 \times 0.125$\,MHz$ = 2048$\,MHz.  This simple operation simultaneously resamples the signal, modulates 
the signal, and trims the guard bands.  Furthermore, it is easily implemented using cuFFT by using the same 
batch size as the second real-to-complex transform, increasing the input pointer index by $150 / 0.125 = 1200$
(equivalent to masking out the first 1200 channels) and then using a input stride of $9980$ for the 
$N=16384$ complex-to-real transform.

The choice of FFT size was set to maximize performance.  In order for the target sampling rate to be achieved
$N_B$ must be a multiple of 39.  Furthermore, so that the 150\,MHz guard band can be removed, we must ensure that 
150 is a multiple of $2496/N_B$:
\begin{equation}
\frac{150\times39}{2496} = \frac{75}{32} = \frac{i}{j}, \quad i,j = 1, 2,3, \ldots
\end{equation}
To ensure this, $j$ should be a multiple of 32 so that 
\begin{equation}
N_B = 39j = 1248i, \quad i=1,2,3,\ldots
\end{equation}
Lastly, the size of the C transform is then determined:
\begin{equation}
N_C = \frac{32}{39} N_B.
\end{equation}
The dashed line in Figure \ref{fig:C2R_performance} shows the performance for candidate $N_B$s.  Unlike for 
the radix-2 cases, the configuration that has the best Gflop performance does not also provide the best Gsps
rate.  This is likely because the Gflop calculation is not an actual flop count, but a scaling from the 
original Cooley-Tukey algorithm \citep{cooley65}.  The Gsps metric is perhaps more useful in this case as it
shows that choices of $N_B <= 19968$ will have good performance.  However, the figure also demonstrates
that this non radix-2 FFT is about $2\times$ slower than the radix-2 case.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{aphids/C2R_performance.eps}
\caption{This figure shows the performance of radix-2, single precision, batched, complex-to-real, 1D transforms 
measured on our GPU cluster. The benchmark conventions follow FFTW  (http://www.fftw.org/speed/method.html).}
\label{fig:C2R_performance}
\end{figure}

\subsection{Timing Performance}

In this section we report on timing measurements of the APHIDS resampling kernels.  We consider the cost of 
loading the data onto the device separately and do not include it in the scores of the individual kernels.  
Furthermore, we do not include the cost of building the FFT plan since this operation is done only once.
We have designed the resampling block to operate on a single GPU so there is no time spent moving data 
between devices and we can parallelize across each card.  The
kernels (except for the depacketer) must be called
twice, once on each phased sum.  Our current implementation uses synchronous calls (one kernel function is 
called at a time).  We tested an asynchronous setup, but saw no speedup.  This indicates that the majority of 
the GPU resources are being used.  Table \ref{tab:kernel_times} lists the time it takes for the kernel to be 
executed a single time and also the speed of the kernel assuming it is operating on a buffered 0.32\,Gb data 
chunk (see \S\ref{sec:resamp_block}).

%% Loss table : /home/krosenfe/resampling/losses.py
% losses for FIR filtering, interpolation schemes, and FFT
\begin{table}
\begin{center}
\caption{Timing measurements of APHIDS kernels \label{tab:kernel_times}}
\begin{threeparttable}
\begin{tabular}{l|cc}
\toprule
Kernel & time [ms] & rate [Gbps] \\
\midrule
Reorder                & 10.5 & 31.5 \\
FFT A (N=32768)        & 25.4 & 12.9 \\
FFT B (N=19968)        & 38.0 & 8.6  \\
FFT C (N=16384)        & 13.8 & 23.7 \\
Quantize (2-bit)       & 12.0 & 27.3 \\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

\subsection{FFT Accuracy}
One concern with using multiple FFTs to change the sampling rate is the error introduced by bit growth.  The 
floating point error grows as $\mathcal{O}(\sqrt{\log{N}})$ \citep{schatzman96} on average for the Cooley-Tukey
algorithm, but errors can be exacerbated by innacurate twiddle factors or our choice of FFT algorithm (see
benchFFT\footnote{\url{http://www.fftw.org/accuracy/}}).  We have computed the accuracy of the cufft library 
using single precision for 
radix-2 transforms as well as the $N_b = 19968$ case in Figure \ref{fig:fft_accuracy}.  These values 
were computed using benchFFT and we compare against fftw3\footnote{\url{http://www.fftw.org/}}
results also computed on our GPU cluster.  For random input, the cufft library follows the expected 
$\mathcal{O}(\sqrt{\log{N}})$ behavior, but with larger scatter and slightly worse performance than fftw3.

\begin{figure}[t]
\plotone{aphids/fft_accuracy.eps}
\caption{Accuracy of the FFT algorithm as measured by the $L_2$ error from benchFFT} \label{fig:fft_accuracy}
%/home/krosenfe/testing/cufft_accuracy/make_figure.py
\end{figure}

Round off error and bit growth is also a general concern for scientific computing on GPUs.  
Increasing the precision to double can significantly slow down performance and then 
require reoptimization or even the redesign of kernels.  Generally, well implemented FFTs can 
be very stable options.  The results of ours test suggest that 
bit growth will not be a large source of loss for the resampling scheme presented in \S\ref{sec:resampling_block}.

\subsection{Quantization}
Another source of loss and distortion to consider is the requantization back to 2-bits that is done
before writing back to disk on the Mark6 (see Figure \ref{fig:aphids_flow_chart}).  In theory, APHIDS could 
record higher bit data (or even the full 32 bits for floating point), but this adds complications for 
correlation 
in DIFX and would incur costs for both disk space and output data rate.  While the former cost is monetary, the 
latter could be an important limiting factor for real-time operation of APHIDS at the back end of the SDBE.  
The signal-to-noise loss expected from 2-bit quantization is a function of the quantization 
thresholds, but for Gaussian noise and Nyquist sampling is at best 12\% \citep{cooper70,thompson01}.

\subsection{Inverting the Polyphase Filterbank}
Under normal operation, SWARM uses a polyphase filter bank (PFB) to generate its spectra.  The PFB is an 
efficient method of producing spectra where each channel has a frequency response much better than the 
sinc pattern of the DFT.  This is achieved by using a window function and a polyphase filter structure to reduce 
the effects of leakage and scalloping \citep{lyons11}.  However, for VLBI observations we must transform the 
spectrum into the original time series to resample the data.  Inverting the PFB is a non-trivial operation so 
the operators use a version of the SWARM bit-code that replaces the PFB with an FFT.  However, for the first 
two nights of the 2015 EHT campaign this version was not used and so we must consider this problem.

One option is to simply apply the generic APHIDS pipeline on the PFP spectra which uses an DFT to invert the 
SWARM spectra into time series.  The resulting pseudo time-series is a series of weighted 
averages of the original time series.  The weights are determined by the window function used in the 
PFB and the number of elements in the sum by the number of taps, $N_T$.  For example, if the window function is 
a box car, then the weights are equal and the correlation coefficient for a white noise time stream goes like:
\begin{equation}
\rho = \frac{1}{\sqrt{N_T}}.
\end{equation}
SWARM uses four taps and a Hanning filter.  We simulated the resulting correlation loss and saw a 34\% loss, 
exceeding the largest loss we expect in the pipeline from the 2-bit quantization (12\%).  Using the PFB inversion
technique developed by Richard Shaw\footnote{\url{https://github.com/jrs65/pfb-inverse}} we see excellent 
recovery with a loss of only 0.5\% although the inversion routine takes about 10 times longer than numpy's
iFFT and already uses the highly optimized LAPACK library.  CUDA libraries offer many of the required functions 
to implement this solution on the GPU and we have ported Shaw's solution to CUDA using a combination of 
CUBLAS and CUSolver.  The current implementation runs about 5 times slower than an inverse FFT batched over 
the same size data chunk.

\subsection{Current Status and Results}

The APHIDS software is still under development, but we have tested the resampling blocks on a small sample of
the data.  To do so we took several hundred milliseconds of data collected by phased SWARM and correlated against
a single dish in the SMA array to find a non-cosmic zero-baseline fringe.  The ability to do this was extremely 
important as it allowed us to characterize how the data was ordered and guide efforts to find fringes at longer 
baselines.  A cosmic fringe detection between the LMT in Mexico and the SMA (both the single dish and phased 
SWARM) were confirmed in May (see Figure \ref{fig:sma_lmt}).

\begin{figure}[t!]
\includegraphics[bb=0 0 100 100]{aphids/SMA-LMT.png}
\caption{This figure shows the cosmic fringe detection at 1.3\,mm between phased SWARM and the LMT, a deprojected 
baseline of more than 4 G$\lambda$ (figure from Mike Titus).}
%% Delay rate is red curve and multiband delay rate is blue curve.
\label{fig:sma_lmt}
\end{figure}

\subsection{Conclusion and Future Development}

APHIDS promises to be an important tool for the EHT.  It crucially enables the correlation in
DIFX of non-standard sampling rates and also provides a platform for any general pre-processing of 
VLBI data.  It features fast data IO from Mark6 recorders and GPU accelarated processing.  
We have demonstrated the correctness and speed of this design and its immediate applicability to 
the 2015 EHT data as well as technical tests with phased ALMA.  However,
there remain a number of challenges for future development of the software.

For future VLBI campaigns at the SMA, it would be ideal if APHIDS could run in real-time 
at the backend of the SDBE.  This would reduce the media requirements and also the 
hassle of post-processing the data offsite.  However, the performance will need to improve as the
SWARM data rates increase and if the front-end of the pipeline is optimized to reduce the 
amount of SNR loss from quantization.  Additionally, some analysis on the cooling requiremeents 
of the cards should be conducted to ensure that the GTX980s used in out GPU cluster can safely run 
at the altitude of the SMA (4100\,meters).  Regardless of where it operates, APHIDS will need 
maintence to remain compatible with SWARM as it and the SDBE are updated.  The current 
git repository and its wiki is a good solution for keeping track of versioning and performance status.

In order to handle phased ALMA data products, a separate APHIDS resampling block would need 
to be developed.  In some ways this is a simpler design since there are no guard-bands, but 
there are more phased sums for book-keeping and larger data requirements for the higher 
sampling rate. APP expertise is located at Haystack and in close collaboration with members of 
the APHIDS team.

\section{The APHIDS Team}
The text to this document was written by KR with referral to intermediate documents written by
members of the APHIDS team.  LB ran simulations on 2-bit quantization and suggested the DFT resampling 
architecture appropriate for phased SWARM. AY wrote sgcomm, the front and back-end kernels de-and re-packetizing 
the data along with quantization. RP handled HASHPIPE and the general pipeline design.  The B-engine data format 
along with the SDBE bit code was developed by LV and RP.

### 2015-08-31 frontmatter/abstract.tex
added abstract

% the abstract

Interferometric observations at millimeter wavelengths provide a precious, detailed view 
of certain astrophysical objects.  This thesis is composed of studies that both rely on and enable this 
technique to study the structure of planet-forming disks and eventually image the closest regions around 
super-massive black holes.  Young stars form out of a cloud of gas and dust that, before its eventual dissipation,
flattens to a disk.  However the disk population is diverse and recent high-resolution images have 
revealed a wide variety of interesting features.  To understand these observations we use detailed radiative 
transfer models to motivate various physical scenarios.  First we identify a set of traits in the disk around
V4046 Sgr that marks the coupled progression of the gas and dust distributions in the presence of at least one
embedded companion.  Next, we investigate how the vertical temperature structure of the disk can be spatially 
resolved and apply our framework to observations of the disk around HD163296.  Lastly, we show how the radial 
flow of gas may be imprinted in the data and question how this phenomenon might be distinguished from other 
scenarios such as warps or outflows.  The last chapter summarizes a solution on how to change the sampling 
rate of data taken at the SMA so that it may be used for VLBI campaigns.

### 2015-08-31 frontmatter/authorlist.tex
added abstract

{\large
\linespread{1.}\selectfont{}
\noindent The following authors contributed to Chapter 1: \newline
Sean M. Andrews, David J. Wilner, J. H. Kastner, M. K. McClure.

\noindent The following authors contributed to Chapter 2: \newline
Sean M. Andrews, A. Meredith Hughes, David J. Wilner.

\noindent The following authors contributed to Chapter 3: \newline
Eugene Chiang and Sean M. Andrews.

\noindent The following authors contributed to Chapter 4: \newline
Lindy Blackburn, Rurik Primiani, Laura Vertatschitsch, and Andre Young.
}

### 2015-09-02 aphids/aphids.tex
updated aphids chapter

@@
 
-To enbable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
+To enable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
 the data before correlating.  In this chapter we describe a software solution to change the 
 samping rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
-focus on the rate conversion problem and investigate several possible algorithms.  The project
+focus on the sampling rate conversion problem and investigate several possible algorithms.  The project
 has applications for future VLBI observations at the SMA as well as other sites with similar 
@@
 
-Very Long Baseline Interferometry (VLBI) uses similar techniques as the connected element arrays
+Very Long Baseline Interferometry (VLBI) uses similar techniques as the connected element arrays such as the 
+SMA and ALMA
 featured in the previous chapters.  By using detectors located thousands of miles away, perhaps on different 
@@
 station must have its own clock and, after the observation campaign, the recorded signals is correlated off-site. 
-Examples of VLBI applications include geodesy, astrometry, and imaging. Technical developments in the 
-field continue to unveil singular scientific opportunities.
+Examples of VLBI applications include geodesy, astrometry, and imaging \citep{felli89,sovers98}.  Connected
+elements arrays may also participate in these experiments as a single, phased array site \citep{thompson01}.
 
-One exciting example is the development of mm and sub-mm wave VLBI by collaborations such as the Event Horizon
-Telescope (EHT).  The EHT observes nearby super massive black holes on scales of their Schwarzschild radius which
-provides a unique lab to study 
-general relativity and accretion physics.  One of the challenges that this experiment faces is the inhomogeneity 
-of its array since it is a conglomeration of many independently operating observatories.
-While many of these telescopes use the same back end hardware such as the state-of-the-art 
-ROACH2 VLBI Digital Back-end (R2DBE), a few sites 
+Technical developments in the field continue to open new scientific opportunities.  One such exciting example 
+is the development of mm and sub-mm wavelength VLBI by collaborations such as the Event Horizon
+Telescope (EHT).  The EHT observes nearby super massive black holes on scales of their Schwarzschild radii and
+their observations provide a unique lab to study general relativity and accretion physics.  One of the 
+challenges that this experiment faces is the inhomogeneity of its array which is composed of many independently 
+operated observatories.  While some of these telescopes use the same back-end hardware such as the 
+state-of-the-art ROACH2 VLBI Digital Back-end \citep[R2DBE;][]{vertatschitsch16}, a few sites 
 cannot use these systems and produce very 
@@
 a particular emphasis on the resampling problem.  APHIDS also has very general applications for 
-post-processing of VLBI data since it features fast read/write capabilities, a modular threaded pipeline 
-and a GPU enabled fast data processing.  The code base is open-source and available online at 
-\url{https://github.com/sma-wideband/sdbe}.
+post-processing of VLBI data since it features fast I/O, a modular threaded pipeline 
+and Graphics Processor Unit (GPU) accelerated data processing.  The code base is open-source and available 
+online at \url{https://github.com/sma-wideband/sdbe}.
 
 The purpose of APHIDS is to take data recorded at the SMA and process it for correlation using
-DIFX.  The SMA is a key site for the EHT that provides important East-West baselines and a zero baseline with 
-its neighbor JCMT.  The phased sum is calculated by the SMA correlator, the SMA Wideband Astronomical 
+DIFX.  The SMA is a key site for the EHT that provides important East-West baselines and a near zero baseline 
+with its neighbor the JCMT (see Figure \ref{fig:eht_baselines}).  The phased sum is calculated by the SMA correlator, the SMA Wideband Astronomical 
 ROACH2 Machine (SWARM\footnote{\url{https://www.cfa.harvard.edu/twiki5/view/SMAwideband}}),
 which is currently being deployed in stages as the array's bandwidth capabilities are 
-upgraded.  For EHT observations in 2015, SWARM recorded at a sampling rate of 2496\,MHz in two separate phased
-sums.
+upgraded.  For EHT observations taken in March 2015, SWARM recorded at a sampling rate of 2496\,MHz in 
+two separate phased sums.
 
 In \S\ref{sec:src_basics} we discuss the sampling rate conversion problem and several common solutions.
-The current APHIDS resampling algorithm is described in \S\ref{sec:aphids} along with discussions of important
+The current APHIDS resampling algorithm is described in \S\ref{sec:aphids} along with several important
 design considerations.  We include measurements of performance as well as a successful
@@
 development.
-
-\section{Sample Rate Conversion} \label{sec:src_basics}
-
-Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
-($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
-ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  Operationally, 
-sample rate conversions is the combination of upsampling and downsampling \citep{oppenheim10,lyons11}.  
-Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
-by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
-anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
-$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
-time and frequency domain for when $L$ equals 3.
 
@@
 
+\section{Sample Rate Conversion} \label{sec:src_basics}
+
+Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
+($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
+ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  A sample rate 
+conversion can be helpfully thought of as the combination of upsampling and downsampling 
+\citep{oppenheim10,lyons11}.  
+Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
+by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
+anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
+$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
+time and frequency domain for when $L$ equals 3.
+
  %% cartoon explaining down-sampling
@@
 \begin{tikzpicture}
-   \label{fig:resample_basic}
    \node[dspnodeopen,dsp/label=left]  (c0) {$x[i]$};
@@
 \caption{Basic multirate resampling signal flow graph.}
+\label{fig:resample_basic}
 \end{figure}
 
-There have been recent development of polyphase filter representations for GPU architectures 
+There have been recent development of polyphase filter representations for GPUs
 \citep[i.e.][]{vanderveldt12,adamek14,kim14a}.  These filter structures can operate at the low sample rate of 
@@
 inner product operation.  However, \citet{kim14a} found that their GPU kernel was dominated by the 
-indexing operations.  \citet{adamek14} (building on the work by \citet{vanderveldt12}) presented tantalizing 
-implementations of 
-polyphase filter banks that could run at data rates in excess of 6.5\,GB/s (their estimate for the single channel 
-output of the SKA's Low Frequency Aperture Array).
+indexing operations.  \citet{adamek14} (building on the work by \citet{vanderveldt12}) presented an
+implementation of a polyphase filter bank that could run at data rates in excess of 6.5\,GB/s 
+(their estimate for the single channel output of the SKA's Low Frequency Aperture Array).
 
 %  https://github.com/wesarmour/astro-accelerate
+\begin{figure}[t!]
+\epsscale{1.0}
+\plotone{aphids/windows.eps}
+\caption{Frequency response for FIR filters equivalent to linear and nearest interpolation when $L/M = 64/39$.
+The orange region shows the first Nyquist zone of the target $L/Mf_0$ sample rate.  Spectral components at all 
+other frequencies are aliased into this region.}
+\label{fig:windows}
+\end{figure}
 
-A simple option for resampling is to use linear interpolation:
+For testing and development we also considered two elementary options for the resampling: linear and nearest 
+neighbor interpolation.  Linear interpolation is:
 \begin{equation}
@@
 
-\begin{figure}[t!]
-\epsscale{1.0}
-\plotone{aphids/windows.eps}
-\caption{Frequency response for FIR filters equivalent to linear and nearest interpolation when $L/M = 64/39$.
-The orange region shows the first Nyquist zone of the target $L/Mf_0$ sample rate.  Spectral components at all 
-other frequencies are aliased into this region.}
-\label{fig:windows}
-\end{figure}
-
 \subsection{Resampling in the Fourier Domain} \label{sec:fourier_resampling}
 In contrast to the previous methods, one can also implement a rational $L/M$ sample rate conversion entirely in 
-the Fourier domain \citep{gold69,yeh82}.  After accumulating $kM$ samples at a clock rate of $f_0$ 
-(where $k=1,2,3\cdots$), the DFT 
+the Fourier domain \citep{gold69}.  After accumulating $kM$ samples at a clock rate of $f_0$ 
+(where $k=1,2,3\ldots$), the DFT 
 returns spectral components spaced at $f_0/kM$.  If $f_1 > f_0$, the resampled spectrum is generated by inserting
@@
 if the original, continuous signal has a discrete spectral density distribution that is band-limited below the 
-Nyquist limit.  In practice, small errors may be introduced from aliased spectral leakage.
+Nyquist limit.  In practice, errors may be introduced from aliased spectral leakage.
 
@@
 wrapped periodically.  Consequenty, errors will be introduced into the beginning and end of the
-resampled signal.  The far this error significantly propagates away from the time series edges depends on the 
-width of the sinc function which is set by the resampling factors $L$ and $M$: 
+resampled signal.  How far and significantly this error propagates away from the time series' edges depends 
+on the width of the sinc function which is set by the resampling factors $L$ and $M$: 
 \begin{equation} \label{eq:sinc}
@@
 
-For post-processing phased SWARM, DFT resampling is a good fit because one can access arbitrary large chunks of 
-the time series without accumulators in hardware and has well-behaved errors
+For post-processing phased SWARM, DFT resampling is a good fit because one can access large chunks of 
+the time series without accumulators in hardware and it has well-behaved errors
 that can be easily modeled.  However, the speed of the FFT depends on the resampling factors ($L/M = 32/39$ for 
 the March 2015 EHT observations) and zero-padding in the case that the $L\gg M$ may require large amounts of 
-memory.
+memory.  In practice, we also found that the memory space of the GPU could be a limiting factor not only 
+on how much data could be simultaneously resampled, but also storing FFT plans (which depends on
+values of $L$ and $M$).
 
@@
 
-\section{APHIDS} \label{sec:aphids}
-
-We are developing a software solution to the resampling problem called APHIDS (Adaptive 
-Phased-array and 
-Heteregeneous Interpolator and Downsampler for SWARM).  This open-source project is written in C/CUDA using
-the High Availibility Shared Pipeline 
-(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
-modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
-and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
-vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
-The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
-\ref{fig:aphids_flow_chart} shows how data progresses 
-through the pipeline and this section will describe each block covering the major hardware components, software 
-design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.
-
 \tikzstyle{line} = [draw,-latex']
@@
 
-\begin{figure}[t]
+\begin{figure}[t!]
 \linespread{1.}\selectfont{}
@@
 \end{tikzpicture}
-\label{fig:aphids_flow_chart}
 \caption{This signal flow chart shows an overview of APHIDS.  The red boxes represent operations on the GPU 
@@
 streams are processed independently and in a serial fashion.}
+\label{fig:aphids_flow_chart}
 \end{center}
@@
 
+\section{APHIDS} \label{sec:aphids}
+
+We are developing a software solution to the resampling problem called APHIDS.  This 
+open-source project is written in C/CUDA using
+the High Availibility Shared Pipeline 
+(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
+modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
+and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
+vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
+The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
+\ref{fig:aphids_flow_chart} shows how data progresses 
+through the pipeline and this section will describe each block covering the major hardware components, software 
+design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.
+
 \subsection{Hardware and Data description}
 
-APHIDS features three main hardware components: an I/O data system, CPU, and GPU server.  For I/O, the 
-EHT observations taken by SWARM in March, 2015 have been stored on four disk-packs that each hold eight, 
-6\,TB hard drives (totaling $48 \times 4 = 192$\,TB of data).  The Mark6 VLBI data system has four modules that 
-can hold one disk-pack and we use
+APHIDS features three main hardware components: an I/O data system, CPU, and GPU cluster.  For I/O, the 
+EHT observations taken by SWARM in March 2015 have been stored on four disk-packs that each hold eight, 
+6\,TB hard drives (totaling 192\,TB of disk space).  The Mark6 VLBI data system has four modules 
+that can hold one disk-pack and we use
 two seprate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
-that controls the GPU server.
+that controls the GPU cluster.
 
-The GPU server as four GTX 980 GPUs, each with 4GB of local memory.  Data from the CPU (or host) is copied over 
+The GPU cluster has four GTX 980 GPUs, each with 4GB of local memory.  Data from the CPU (or host) is copied over 
 a PCI-E bus into the local GPU (or device) memory.  Once the data is transferred, the host directs the device
@@
 The observations were recorded using a scatter/gather file system which disperses the 
-data composing individual scans over the four modules.  APHIDS uses fast, general purpose software called 
+individual scan data over the four modules in a Mark6.  APHIDS uses fast, general purpose software called 
 sgcomm (Scatter/Gather Communication) to read in 
-the scattered data into CPU shared memory over a 10 Gigabit Ethernet network.  Data from phased SWARM was 
-streamed through a special version of the R2DBE called the SDBE (Swarm Digital Back End) and packaged into
-a customized B-engine format.  Each B-engine frame is composed of 1024, 1056B VDIF frames 
+the scattered data into CPU shared memory over a 10 Gigabit Ethernet.  Data from phased SWARM was 
+streamed through a specialized version of the R2DBE called the SDBE (Swarm Digital Back End) and packaged into
+a customized B-engine format.  Each B-engine frame is composed of 2048, 1056B VDIF frames 
 \footnote{\url{http://www.vlbi.org/vdif/}} with 128 snapshots of a 
@@
 However, the observation was designed to include guard bands which must be removed by APHIDS 
-(see Figure \ref{fig:swarm_amp_spec}).
+(see Figure \ref{fig:swarm_amp_spec} and \S\ref{sec:resamp_block}).
 
 \begin{figure}[t!]
-\epsscale{1.0}
-\plotone{aphids/SWARM_amp_spectra.eps}
-% /home/krosenfe/sdbe/software/prototyping/cuda/SWARM_amp_spectra.eps
-\caption{This figure shows both channels of the detrended amplitude spectral density of phased SWARM along with 
-the R2DBE (SMA single-dish in gray).}
-\label{fig:swarm_amp_spec}
+\linespread{1.}\selectfont{}
+\centering
+\begin{tikzpicture}[node distance = 2.2cm, auto]
+\node[align=center](init){};
+\node[resamp,below of=init](irfft){(A) batched real IFFT \\ dimension of $32768$};
+\node[resamp,below of=irfft](rfft){(B) batched real FFT \\ dimension of $39 \times 512$};
+\node[resamp,below of=rfft](trim){Trim band and shift frequency};
+\node[resamp,below of=trim](rifft){(C) batched real IFFT \\ dimension of $32 \times 512$};
+\node[align=center,below of=rifft](final){};
+
+\path[line](init) -- node[align=left]{ complex spectra \\ @ 2496\,MHz} (irfft);
+\path[line](irfft) -- (rfft);
+\path[line](rfft) -- (trim);
+\path[line](trim) -- (rifft);
+\path[line](rifft) -- node[align=left]{ time series \\ @ 2048\,MHz}(final);
+\end{tikzpicture}
+\caption{The resampling block trims the phased SWARM guard bands and shifts the passband to DC as part of the 
+resampling operation.}
+\label{fig:resampling_block}
 \end{figure}
@@
 FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
-and ensure that transforms operate on contiguous chunks of memory.  We also use single precision transforms to 
+and operate on contiguous chunks of memory.  We also use single precision transforms to 
 reduce the memory transfer bandwidth and ensure that 
@@
 
-\begin{figure}[t!]
-\linespread{1.}\selectfont{}
-\centering
-\begin{tikzpicture}[node distance = 2.2cm, auto]
-\node[align=center](init){};
-\node[resamp,below of=init](rfft){batched real FFT \\ dimension of $39 \times N$};
-\node[resamp,below of=rfft](trim){Trim band and shift frequency};
-\node[resamp,below of=trim](rifft){batched real IFFT \\ dimension of $32 \times N$};
-\node[align=center,below of=rifft](final){};
-
-\path[line](init) -- node[align=left]{ time series \\ @ 2496\,MHz} (rfft);
-\path[line](rfft) -- (trim);
-\path[line](trim) -- (rifft);
-\path[line](rifft) -- node[align=left]{ time series \\ @ 2048\,MHz}(final);
-\end{tikzpicture}
-\label{fig:resampling_block}
-\caption{The resampling block trims the phased SWARM guard bands and shifts the passband to DC as part of the 
-resampling operation.}
-\end{figure}
-
 The first FFT (A) takes the complex valued, 16384 sample SWARM spectrum into the time domain, returning 
-32768 real samples.  Since SWARM omits the Nyquist sample which is far outside the SWARM cutoff frequency, 
+32768 real samples.  Since SWARM omits the Nyquist sample which is outside the SWARM guard band cutoff frequency, 
 we insert a zero as the 16385th sample.  Since this transform is radix-2, the performance should benefit from 
@@
 and we find that this transform, just ouside of the optimal regime, operates at 6.7 Gsps.
+
+\begin{figure}[t!]
+\epsscale{0.95}
+\plotone{aphids/C2R_performance.eps}
+\caption{This figure shows the performance of radix-2, single precision, batched, complex-to-real, 1D transforms 
+measured on our GPU cluster (solid lines). The dashed lines show the same measures for non radix-2, candidate 
+resampling FFT sizes ($N_B$).  The benchmark conventions follow FFTW  (http://www.fftw.org/speed/method.html).}
+\label{fig:C2R_performance}
+\end{figure}
 
@@
 original Cooley-Tukey algorithm \citep{cooley65}.  The Gsps metric is perhaps more useful in this case as it
-shows that choices of $N_B <= 19968$ will have good performance.  However, the figure also demonstrates
+shows that choices of $N_B \leq 19968$ will have good performance.  However, the figure also demonstrates
 that this non radix-2 FFT is about $2\times$ slower than the radix-2 case.
-
-\begin{figure}[t!]
-\epsscale{0.95}
-\plotone{aphids/C2R_performance.eps}
-\caption{This figure shows the performance of radix-2, single precision, batched, complex-to-real, 1D transforms 
-measured on our GPU cluster. The benchmark conventions follow FFTW  (http://www.fftw.org/speed/method.html).}
-\label{fig:C2R_performance}
-\end{figure}
 
@@
 algorithm, but errors can be exacerbated by innacurate twiddle factors or our choice of FFT algorithm (see
-benchFFT\footnote{\url{http://www.fftw.org/accuracy/}}).  We have computed the accuracy of the cufft library 
+benchFFT\footnote{\url{http://www.fftw.org/accuracy/}}).  We have computed the accuracy of the cuFFT library 
 using single precision for 
@@
 were computed using benchFFT and we compare against fftw3\footnote{\url{http://www.fftw.org/}}
-results also computed on our GPU cluster.  For random input, the cufft library follows the expected 
+results also computed on our GPU cluster.  For random input, the cuFFT library follows the expected 
 $\mathcal{O}(\sqrt{\log{N}})$ behavior, but with larger scatter and slightly worse performance than fftw3.
@@
 be very stable options.  The results of ours test suggest that 
-bit growth will not be a large source of loss for the resampling scheme presented in \S\ref{sec:resampling_block}.
+bit growth will not be a large source of loss for the resampling scheme presented in \S\ref{sec:resamp_block}.
 
@@
 
-One option is to simply apply the generic APHIDS pipeline on the PFP spectra which uses an DFT to invert the 
+One option is to simply apply the generic APHIDS pipeline on the PFB spectra which uses an DFT to invert the 
 SWARM spectra into time series.  The resulting pseudo time-series is a series of weighted 
@@
 
-\subsection{Current Status and Results}
-
-The APHIDS software is still under development, but we have tested the resampling blocks on a small sample of
-the data.  To do so we took several hundred milliseconds of data collected by phased SWARM and correlated against
-a single dish in the SMA array to find a non-cosmic zero-baseline fringe.  The ability to do this was extremely 
-important as it allowed us to characterize how the data was ordered and guide efforts to find fringes at longer 
-baselines.  A cosmic fringe detection between the LMT in Mexico and the SMA (both the single dish and phased 
-SWARM) were confirmed in May (see Figure \ref{fig:sma_lmt}).
-
 \begin{figure}[t!]
-\includegraphics[bb=0 0 100 100]{aphids/SMA-LMT.png}
-\caption{This figure shows the cosmic fringe detection at 1.3\,mm between phased SWARM and the LMT, a deprojected 
-baseline of more than 4 G$\lambda$ (figure from Mike Titus).}
-%% Delay rate is red curve and multiband delay rate is blue curve.
-\label{fig:sma_lmt}
+\epsscale{1.0}
+\includegraphics{aphids/kr_autos.eps}
+\caption{This figure shows the spectral density of phased SWARM along with the single reference dish 
+(in purple).  The APHIDS output is also plotted (dotted).}
+\label{fig:swarm_amp_spec}
 \end{figure}
 
-\subsection{Conclusion and Future Development}
+\subsection{Current Status and Future Development}
+
+APHIDS is still under active development, but rapidly nearing full-scale production.  Figure \ref{fig:swarm_amp_spec}
+shows the original SDBE and resampled auto spectra from data taken in March 2015.  Using data resampled by APHIDS, Mike Titus, 
+the correlator engineer at MIT Haystack, has found a cosmic fringe on 3C297 using the short 
+baseline between the JCMT and the SMA (see Figure \ref{fig:aphids_fringe}).  He reported an increase in the fringe amplitude over initial tests
+using linear interpolation.  APHIDS currently runs about 4.5 times slower than real time, sufficient for 
+processing the March 2015 EHT campaign data.  The resampling block with PFB inversion should be available
+shortly.
 
@@
 DIFX of non-standard sampling rates and also provides a platform for any general pre-processing of 
-VLBI data.  It features fast data IO from Mark6 recorders and GPU accelarated processing.  
+VLBI data.  It features fast data I/O from Mark6 recorders and GPU accelarated processing.  
 We have demonstrated the correctness and speed of this design and its immediate applicability to 
-the 2015 EHT data as well as technical tests with phased ALMA.  However,
+the 2015 EHT data.  However,
 there remain a number of challenges for future development of the software.
@@
 amount of SNR loss from quantization.  Additionally, some analysis on the cooling requiremeents 
-of the cards should be conducted to ensure that the GTX980s used in out GPU cluster can safely run 
+of the cards should be conducted to ensure that the GTX980s used in our GPU cluster can safely run 
 at the altitude of the SMA (4100\,meters).  Regardless of where it operates, APHIDS will need 
@@
 
-In order to handle phased ALMA data products, a separate APHIDS resampling block would need 
-to be developed.  In some ways this is a simpler design since there are no guard-bands, but 
-there are more phased sums for book-keeping and larger data requirements for the higher 
-sampling rate. APP expertise is located at Haystack and in close collaboration with members of 
-the APHIDS team.
+%\includepdf[pages={1}]{aphids/089-0810_JS_fix.pdf}
+\begin{figure}
+\includegraphics[width=\linewidth]{aphids/089-0810_JS_fix.pdf}
+\caption{Fourfit plots of a sky fringe found using APHIDS resampled SWARM data and JCMT.}
+\label{fig:aphids_fringe}
+\end{figure}
 
 \section{The APHIDS Team}
-The text to this document was written by KR with referral to intermediate documents written by
-members of the APHIDS team.  LB ran simulations on 2-bit quantization and suggested the DFT resampling 
-architecture appropriate for phased SWARM. AY wrote sgcomm, the front and back-end kernels de-and re-packetizing 
-the data along with quantization. RP handled HASHPIPE and the general pipeline design.  The B-engine data format 
-along with the SDBE bit code was developed by LV and RP.
+Andre Young, Lindy Blackburn, Laura Vertatschitsch, Rurik Primiani, and Katherine Rosenfeld are part of 
+the core APHIDS team located at the Smithsonian Astrophysical Observatory.  Jonathan Weintroub and Sheperd
+Doeleman were responsible for project oversight and invaluable support was provided by Mike Titus and Geoff Crew at MIT
+Haystack.  The EHT is supported by the NSF and the Gordon and Betty Moore Foundation.

### 2015-09-03 chapters/conclusion.tex
basic edits to intro and conclusion

@@
 
-In Chapter \ref{chap:v4046_structure} we 
+In Chapter \ref{chap:v4046sgr_structure} we 
 used the SMA to study the protoplanetary disk 
@@
 drag forces.  This disk is one published example among a few and more similar systems continue to be
-identified \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A related phenomena are the 
-severely asymmetric dust  ``horse-shoes'' that have been  observed at multiple wavelengths
+identified \citep{isella07,panic09,andrews12,degregorio-monsalvo13,zhang14}.  A related phenomenon are the 
+severely asymmetric dust  ``horse-shoe'' disks that have been observed at multiple wavelengths
 \citep{vandermarel13,fukagawa13,isella13,casassus15,marino15}.  Thought to be perhaps generated by gigantic 
-3D Rossby wave instabilities \citep{regaly13}, these are another striking example of the interplay between
-the gas and dust in protoplanetary disks.  A particular exciting example are the narrow gaps in the dust continuum
-of HL Tau (see Figure \ref{fig:hltau}. This striking image was obtained using ALMA baselines as long as 15\,km 
+3D Rossby wave instabilities \citep{regaly12}, these are another striking example of the interplay between
+the gas and dust in protoplanetary disks.  Another particular exciting example is HL Tau where recent
+observations have revealed narrow gaps in the dust 
+continuum (see Figure \ref{fig:hltau}. This image was obtained using ALMA baselines as long as 15\,km 
 and possible causes include
-planet-disk interactions and condensation fronts \citep{brogan14,zhang15}.
+planet-disk interactions and condensation fronts \citep{brogan15,zhang15}.
 
 \begin{figure}
-\plotone{chapters/HLTau_nrao.jpg}
+\centering
+\includegraphics[width=0.5\linewidth]{chapters/HLTau_nrao.jpg}
 \caption{This ALMA observation of the HL Tau disk revealed multiple rings and gaps \citep{brogan15}.}
@@
 radial gas flows within the holes of transition disks.  \citet{casassus15} have studied the gas kinematics for
-the curious case of HD142527 using the CO $J = 6-5$ line and found that their data is consistent with
+the curious case of HD142527 using the CO $J$ = $6$-$5$ line and found that their data is consistent with
 accretion reaching close to free-fall velocities.  The gas surface density of transition disks have also 
@@
 \citep{bruderer13}.  Molecular disk winds, outflows, and binary systems are all recently observed 
-with interesting gas kinematics \citep{klaasen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
+with interesting gas kinematics \citep{klaassen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
 future, well resolved, highly sensitive observations of spectral lines may constrain the strength of MRI-drive
-turbulence \citep{simon15}, query the active sites of planet formation \citep{kleeves15,ober15}, or reveal
+turbulence \citep{simon15}, query the active sites of planet formation \citep{cleeves15,ober15}, or reveal
 circumplanetary disks \citep{perez15}.
 
-The ground-breaking Event Horizon Telescope continues to report exciting new science with every
+The ground-breaking Event Horizon Telescope continues to report exciting, new science with every
 campaign from the early detections of AGN at mm wavelengths \citep{doeleman05} to structures
-at the event horizon of a supermassive black hole \citep{doeleman08,doeleman09,doeleman13}.
-The 2015 campaign represented a new scale of operation, incorporating more telescopes and
-technologies.  In Chapter \ref{chap:aphids} we reported on the status of the APHIDS software which
+at the event horizon of a supermassive black hole \citep{doeleman08,doeleman12}.
+The 2015 campaign represented a new scale of operations, incorporating more telescopes and
+fielding new technologies.  In Chapter \ref{chap:aphids} we reported on the status of the APHIDS software which
 will post-process the phased SMA data and make the sampling rate compatible with the rest of the EHT
@@
 prove useful for other sites and VLBI experiments.
-
-This thesis explores
-ence that resolution at these wavelengths can provide and only an exploration for the 
-hard and promising work ahead.

### 2015-09-03 chapters/introduction.tex
basic edits to intro and conclusion

@@
 the information contained in this image will be limited to the scales originally measured by the interferometer.
-Images with the best fidelity come from arrays that have elements at a wide variety of 
-distances to each other and the highest resolution comes from the most widely separated telescopes.
+Images with the best fidelity come from arrays that have many elements spaced at a wide variety of 
+distances and the highest resolution comes from the telescopes with the farthest separation.
 
@@
 \begin{equation}
-r(\tau) = \lim_{\tau \to \infty} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
+r(\tau) = \lim_{T \to \infty} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
 \end{equation}
@@
 
-The Submillimeter Array \citep[SMA][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is 
+The Submillimeter Array \citep[SMA;][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is 
 located on the top of Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
@@
 capabilities.  Once completed, ALMA will host an impressive 66 antennas, 64 of will be 12\,meter dishes and
-the remaining smaller 7\,meter ones (see Figure \ref{fig:sma_alma}).  We used publickly available data from 
+the remaining smaller 7\,meter ones (see Figure \ref{fig:sma_alma}).  We used publicly available data from 
 this instrument in Chapters \ref{chap:hd163296} and \ref{chap:radial_flows}.  A fully extended configuration 
@@
 breaks this process into several stages \citep{shu87,mckee07}.  Picture first a large 
-molecular cloud within which there are complicated sub structures: clumps and cores of 
+molecular cloud made up of complex sub-structures: clumps and cores of 
 concentrated gas and dust.  Over time these self-gravitating units will contract,  the material collapses into 
-a disk to conserve angular momentum, and nascent stars ignite their thermonuclear engines.  Jets, winds, and 
+a disk to conserve angular momentum, and budding stars ignite their thermonuclear engines.  Jets, winds, and 
 outflows punctuate this process.  The system might also be perturbed by the swift pull of a nearby passing star 
@@
 So while we may be at the end of the star's formation, this protoplanetary disk is the stage where planets are
-born.  Their composition and structure provides the crucial initial conditions that must enable a diverse 
+born.  Protoplanetary disk composition and structure provides the crucial initial conditions that must 
+enable a diverse 
 population of planets.  Theories of planet formation are still developing and require the consideration 
@@
 giant holes centered around the star for a class of ``transition disks'' \citep{hughes07,brown09,andrews11} and 
-evidence for binary interactions \citep[i.e.][]{akeson98}.
+evidence for binary interactions \citep{akeson98}.
 
@@
 density, temperature, and velocity profiles of the disks are assumed to 
-follow power laws or some other parametric function.  There also exist some grids and models that try to 
-calculate abundances in a self-consistent manner.  Additionally, researchers usually either try to constrain 
-populations using 
+follow power laws or some other parametric function.  There also exist some model grids that provide more
+physically motivated and self consistent disk structures \citep{dalessio05,woitke10}.  Additionally, researchers 
+usually either try to constrain populations using 
 surveys \citep[i.e.][]{andrews11,oberg11,harris12} or present detailed studies of single, 
@@
 V4046 Sgr is a system with two stars closely orbiting each other and around this tight binary is a bright
-protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present SMA observations 
-of this disk that reveal a large central cavity in the dust.  In addition, we show that the distribution dust 
-does not follow that of the gas: the gas disk is large and extended while the 
-millimeter sized dust grains are concentrated in a narrow ring.  Furthermore, thethe amount of dust left inside
+protoplanetary disk.  In Chapter \ref{chap:v4046sgr_structure} we present SMA observations 
+of this disk that reveal a large central cavity in the dust.  In addition, we show that the dust 
+distribution is significantly different than that of the gas: the gas disk is large and extended while the 
+millimeter sized dust grains are concentrated in a narrow ring.  Furthermore, the amount of dust left inside
 the cavity is inconsistent with photoevaporative scenarios.  We argue 
@@
 
-HD163296 also has a very large, gas rich disk, but no observed central cavity.  In chapter \ref{chap:hd163296} 
+HD163296 also has a very large, gas rich disk, but no central cavity has yet been observed.  In Chapter 
+\ref{chap:hd163296} 
 we use science verification data from ALMA to study the vertical structure of 
-this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
+this disk.  We argue that certain resolved features in the images can be attributed to how the temperature of 
+the gas changes with
 its distance from the disk mid-plane and explore how additional, nuanced aspects of the disk structure may affect
@@
 with possible sources of confusion such as warps in the disk.  One disk has provided intriguing
-evidence of such flows \citep{casassus13,cassasus15} and future observations may provide others.
+evidence of such flows \citep{casassus13,casassus15} and future observations may be able to detect others.
 
@@
 \plotone{chapters/EHT-baselines.png}
-\caption{A selection of EHT baselines viewed looking back from the galactic center where Sgr$^\ast$ lies.  The 
+\caption{A selection of EHT baselines viewed looking back from the Galactic Center where Sgr$^\ast$ lies.  The 
 2015 campaign included sites 
 from Hawaii, California, Arizona, Mexico, and Spain. ({\it image credit}: the EHT collaboration)}
+\label{fig:eht_baselines}
 \end{figure}
@@
 
-The Event Horizon Telescope (EHT) observes supermassive black holes with baselines long enough to
+The Event Horizon Telescope (EHT) observes supermassive black holes with baselines long enough
 to resolve event horizon scales of nearby targets.  These VLBI observations provide a 
 unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
-and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
-center and the blaock hole at the center of the giant elliptical galaxy M87 
-\citep{doeleman08,doeleman09}.  These data have provided indirect 
+and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the Galactic 
+Center and the black hole at the center of the giant elliptical galaxy M87 
+\citep{doeleman08,doeleman09}.  These data provide indirect 
 evidence for the existence of an event horizon \citep{broderick15} along with constraints on the degree of 
@@
 
-The bright future of the EHT is partly driven by its ability to add more telescopes and
-incorporate new technologies.  The March 2015 observing campaign was a major milestone in
+The bright future of the EHT is partialy driven by its ability to add more telescopes and
+incorporate new technologies.  The recent March 2015 observing campaign marked a major milestone in
 the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
-Research in Millimeter-wave Astornomy (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
- Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  However, the data taken at the SMA must be 
+Research in Millimeter-wave Astronomy (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
+Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  Figure \ref{fig:eht_baselines} shows the
+impressive geographic span of the EHT including sites in Chile and at the South Pole that will soon be 
+incorporated into the array.  However, the data taken at the SMA must be 
 processed before it
-can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the APHIDS pipeline
+can be correlated with the rest of the array.  In Chapter \ref{chap:aphids} we describe the APHIDS pipeline
 which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
-is intended to be a long-term solution enabling VLBI with the SMA and has applications for other sites such 
-as ALMA.
+is intended to be a long-term solution enabling VLBI with the SMA.

### 2015-09-03 frontmatter/abstract.tex
basic edits to intro and conclusion

@@
 of certain astrophysical objects.  This thesis is composed of studies that both rely on and enable this 
-technique to study the structure of planet-forming disks and eventually image the closest regions around 
+technique to study the structure of planet-forming disks and soon image the closest regions around 
 super-massive black holes.  Young stars form out of a cloud of gas and dust that, before its eventual dissipation,
@@
 V4046 Sgr that marks the coupled progression of the gas and dust distributions in the presence of at least one
-embedded companion.  Next, we investigate how the vertical temperature structure of the disk can be spatially 
-resolved and apply our framework to observations of the disk around HD163296.  Lastly, we show how the radial 
-flow of gas may be imprinted in the data and question how this phenomenon might be distinguished from other 
-scenarios such as warps or outflows.  The last chapter summarizes a solution on how to change the sampling 
+embedded companion.  Next, we investigate how the vertical temperature structure of a disk can be spatially 
+resolved and apply our framework to observations of the disk around HD163296.  Lastly, we show how large-scale
+radial flows of gas may be observable and question how this phenomenon might be distinguished from other 
+scenarios such as warps or outflows.  The last chapter summarizes the APHIDS project which changes the sampling 
 rate of data taken at the SMA so that it may be used for VLBI campaigns.

### 2015-09-03 frontmatter/authorlist.tex
basic edits to intro and conclusion

@@
 
-\noindent The following authors contributed to Chapter 4: \newline
-Lindy Blackburn, Rurik Primiani, Laura Vertatschitsch, and Andre Young.
+\noindent The following people contributed to Chapter 4: \newline
+Lindy Blackburn, Geoff Crew, Sheperd Doeleman, Rurik Primiani, Mike Titus, Laura Vertatschitsch, 
+Jonathan Weintroub, and Andre Young
 }

### 2015-09-04 aphids/aphids.tex
spell checked

@@
 
-To enable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
+To enable VLBI observations with heterogeneous arrays, it is sometimes necessary to pre-process
 the data before correlating.  In this chapter we describe a software solution to change the 
-samping rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
+sampling rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
 focus on the sampling rate conversion problem and investigate several possible algorithms.  The project
@@
 by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
-anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
+anti-imaging filter smooths the signal and suppresses the high frequency spectral images greater than the original 
 $f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
@@
 Trimming or padding the spectrum is equivalent to convolving a normalized sinc function with the original series 
-wrapped periodically.  Consequenty, errors will be introduced into the beginning and end of the
+wrapped periodically.  Consequently, errors will be introduced into the beginning and end of the
 resampled signal.  How far and significantly this error propagates away from the time series' edges depends 
@@
 open-source project is written in C/CUDA using
-the High Availibility Shared Pipeline 
+the High Availability Shared Pipeline 
 (HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
@@
 that can hold one disk-pack and we use
-two seprate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
+two separate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
 that controls the GPU cluster.
@@
 these operations.  Lastly, we can utilize the optimized cuFFT library
-\footnote{\url{https://developer.nvidia.com/cuFFT}} to ensure good performace, reduce development time and 
+\footnote{\url{https://developer.nvidia.com/cuFFT}} to ensure good performance, reduce development time and 
 maintain a manageably sized codebase.
@@
 The APHIDS resampling block is composed of three 1-dimensional FFTs (A, B, and C; see Figure 
-\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performace, we utilize batched
+\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performance, we utilize batched
 FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
@@
 report\footnote{\url{http://on-demand.gputechconf.com/gtc/2015/webinar/gtc-express-cuda7-performance-overview.pdf}} 
-and we find that this transform, just ouside of the optimal regime, operates at 6.7 Gsps.
+and we find that this transform, just outside of the optimal regime, operates at 6.7 Gsps.
 
@@
 Next, we Fourier transform chunks of $N_B = 19968$ consecutive time series samples (B).  The resulting 
-spectrum has a frequency resolution of $2496$\,MHz$/19958=125$\,kHz.  The folllowing step is to apply a third
-complex-to-real transform (C) on a subset of the spectrum, using only frequences from $150$\,MHz to $1174$\,MHz.  
+spectrum has a frequency resolution of $2496$\,MHz$/19958=125$\,kHz.  The following step is to apply a third
+complex-to-real transform (C) on a subset of the spectrum, using only frequencies from $150$\,MHz to $1174$\,MHz.  
 The final time series has size $N_C= 2\times(1174-150)/0.125=16384$ per batch element and our desired 
@@
 floating point error grows as $\mathcal{O}(\sqrt{\log{N}})$ \citep{schatzman96} on average for the Cooley-Tukey
-algorithm, but errors can be exacerbated by innacurate twiddle factors or our choice of FFT algorithm (see
+algorithm, but errors can be exacerbated by inaccurate twiddle factors or our choice of FFT algorithm (see
 benchFFT\footnote{\url{http://www.fftw.org/accuracy/}}).  We have computed the accuracy of the cuFFT library 
@@
 DIFX of non-standard sampling rates and also provides a platform for any general pre-processing of 
-VLBI data.  It features fast data I/O from Mark6 recorders and GPU accelarated processing.  
+VLBI data.  It features fast data I/O from Mark6 recorders and GPU accelerated processing.  
 We have demonstrated the correctness and speed of this design and its immediate applicability to 
@@
 at the backend of the SDBE.  This would reduce the media requirements and also the 
-hassle of post-processing the data offsite.  However, the performance will need to improve as the
+hassle of post-processing the data off-site.  However, the performance will need to improve as the
 SWARM data rates increase and if the front-end of the pipeline is optimized to reduce the 
-amount of SNR loss from quantization.  Additionally, some analysis on the cooling requiremeents 
+amount of SNR loss from quantization.  Additionally, some analysis on the cooling requirements 
 of the cards should be conducted to ensure that the GTX980s used in our GPU cluster can safely run 
 at the altitude of the SMA (4100\,meters).  Regardless of where it operates, APHIDS will need 
-maintence to remain compatible with SWARM as it and the SDBE are updated.  The current 
+maintenance to remain compatible with SWARM as it and the SDBE are updated.  The current 
 git repository and its wiki is a good solution for keeping track of versioning and performance status.

### 2015-09-04 chapters/conclusion.tex
spell checked

@@
 
-To conclude our work on protoplanary disks, we discussed in Chapter \ref{chap:radial_flows} observational 
+To conclude our work on protoplanetary disks, we discussed in Chapter \ref{chap:radial_flows} observational 
 indicators of 

### 2015-09-04 chapters/introduction.tex
spell checked

@@
 photons and aspects of our physical world that operate on quantum scales.  Today, the techniques of interferometry
-are ubiquitious over a wide array of technical fields.  Even within astronomy the scientific applications are 
+are ubiquitous over a wide array of technical fields.  Even within astronomy the scientific applications are 
 diverse: interferometers  search for extrasolar planets, find subtle ripples in the early universe and watch 
@@
 without the resolving power of interferometry, might remain hidden.  Going along with this theme, this thesis 
-presents four studies using observations at millimeter wavelenghts that work to reveal fine structures in disks 
+presents four studies using observations at millimeter wavelengths that work to reveal fine structures in disks 
 around young stars and massive black holes.  We cover a wide variety of techniques, from enabling the production 
@@
 observatories.  The difficulty of observing at these short wavelengths increases as the atmosphere becomes 
-noiser, time variable, and more opaque.  Consequently, facilities that can operate at these wavelengths are a 
+nosier, time variable, and more opaque.  Consequently, facilities that can operate at these wavelengths are a 
 precious resource.
 
-The Submillimeter Array \citep[SMA;][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is 
+The Submillimeter Array \citep[SMA;][]{ho04} is one such world class array.  Commissioned in 2003, the SMA is 
 located on the top of Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
@@
 configurations so that there are elements as close as 15 meters or as far as 500 meters (see Figure 
-\ref{fig:sma_alma}).  Its recievers can operate at wavelengths from 430\,microns to 2\,millimeters and 
+\ref{fig:sma_alma}).  Its receivers can operate at wavelengths from 430\,microns to 2\,millimeters and 
 historically have provided up to 4\,GHz of bandwidth.  Chapter \ref{chap:v4046sgr_structure} features an 
-exquisite dataset from the SMA that combines observations taken by multiple configurations.  A major upgrade to 
+exquisite data set from the SMA that combines observations taken by multiple configurations.  A major upgrade to 
 the array is increasing the bandwidth to 8\,GHz, enabling more sensitive observations and larger spectral 
@@
 
-Over time, this disk of gas and dust will dissipate leaving trace debris: astroids and comets, dust and planets.
+Over time, this disk of gas and dust will dissipate leaving trace debris: asteroids and comets, dust and planets.
 So while we may be at the end of the star's formation, this protoplanetary disk is the stage where planets are
@@
 hydrogen, but observations focus on easier to detect trace molecules such as $^{12}$CO and its isotopologues.
-The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
+The dust grains are detected via their bulk thermal emission which, when the emission is optically thin,  
 can be used to measure of dust mass \citep{hildebrand83}.  Interferometric observations taken at 
@@
 
-The Event Horizon Telescope (EHT) observes supermassive black holes with baselines long enough
+The Event Horizon Telescope (EHT) observes super massive black holes with baselines long enough
 to resolve event horizon scales of nearby targets.  These VLBI observations provide a 
@@
 Center and the black hole at the center of the giant elliptical galaxy M87 
-\citep{doeleman08,doeleman09}.  These data provide indirect 
+\citep{doeleman08,doeleman12}.  These data provide indirect 
 evidence for the existence of an event horizon \citep{broderick15} along with constraints on the degree of 
@@
 
-The bright future of the EHT is partialy driven by its ability to add more telescopes and
+The bright future of the EHT is partially driven by its ability to add more telescopes and
 incorporate new technologies.  The recent March 2015 observing campaign marked a major milestone in
@@
 can be correlated with the rest of the array.  In Chapter \ref{chap:aphids} we describe the APHIDS pipeline
-which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
+which performs a sampling rate conversion using a cluster of Graphics Processing Units (GPUs).  This software
 is intended to be a long-term solution enabling VLBI with the SMA.

### 2015-09-04 ms_radial_092713/ms_radial_chapter.tex
chapter 2 summary
fixed figure placement
spell checked

@@
-\chapter{Fast Radial Flows in Transition Disk Holes}\label{c:radial_flows}
+\chapter{Fast Radial Flows in Transition Disk Holes}\label{chap:radial_flows}
 
-\section*{abstract}
+%%\section*{abstract}
 
@@
 \citep[e.g.,][]{Kim13} or spatially resolved imaging at infrared 
-\citep[e.g.,][]{geers2007} and mm wavelengths \citep[e.g.,][]{brown09}.  The
+\citep[e.g.,][]{geers07} and mm wavelengths \citep[e.g.,][]{brown09}.  The
 cavities in these ``transition'' disks have radii ranging from $\sim$4\,AU 
@@
 Still, some indirect empirical support for the scalings in equation 
-(\ref{eqn:simple}) has been found by \cite{fung13}.
+(\ref{eqn:simple}) has been found by \cite{fung14}.
 }
